%!TEX root = forallxbris.tex
\part{Truth-functional logic}
\label{ch.TFL}
\addtocontents{toc}{\protect\mbox{}\protect\hrulefill\par}

\chapter{A Prolegomenon to TFL}
In this part the lecture notes we start developing our theory of validity and logical consequence, that is, we start introducing the logical language. In \S\ref{s:Valid} we already introduced the idea that an argument is valid if and only if the truth of the premises guarantees the truth of the conclusion in virtue of their logical form. It is this idea that suggest to spell out an account of validity in a formal language, that is, to conceive of the logical language as a formal language. This will enable us to single out arguments that are valid in virtue of their form and eventually make sense of the notion of an interpretation we used in our definition of validity in \S\ref{s:Valid}. We can then give a rigorous formal definition of validity of arguments in the formal language we shall devise. This language will be the language of Truth-functional logic (TFL).

Before we introduce the language of TFL, let us take a look at why a formal language may be helpful for capturing validity of arguments, i.e., the validity of arguments in virtue of their form.

Consider this argument:
	\begin{earg}
		\prem It is raining outside.
		\prem If it is raining outside, then Jenny is miserable.
		\conc Jenny is miserable.
	\end{earg}
and another argument:
	\begin{earg}
		\prem Jenny is an anarcho-syndicalist.
		\prem If Jenny is an anarcho-syndicalist, then Dipan is an avid reader of Tolstoy.
		\conc Dipan is an avid reader of Tolstoy.
	\end{earg}
Both arguments are valid, and there is a straightforward sense in which we can say that they share a common structure. We might express the structure thus:
	\begin{earg}
		\prem A
		\prem If A, then B
		\conc B
	\end{earg}
This looks like an excellent argument \emph{structure}. Indeed, surely any argument with this \emph{structure} will be valid.

What about:
	\begin{earg}
		\prem Jenny is miserable.
		\prem If it is raining outside, then Jenny is miserable.
		\conc It is raining outside.
	\end{earg}
The form of this argument is:
\begin{earg}
\prem	$B$
\prem	If $A$ then $B$
\conc $A$
\end{earg}
Arguments of this form are generally invalid.

Be careful, though, not every argument of this form is sure to be invalid.
It’s possible to have an argument of this form that’s valid – see if you can work out how!
But most arguments of this form are invalid.

There a lot more valid argument forms. For example the
argument form
	\begin{earg}
		\prem $A$ or $B$
		\prem not-$A$
		\conc $B$
	\end{earg}
as well as the form
	\begin{earg}
		\prem not-($A$ and $B$)
		\prem $A$
		\conc not-$B$
	\end{earg}
lead to valid arguments independently of what expressions we substitute for `$A$' and `$B$': we can understand (interpret) `$A$' and `$B$' in whatever way we want, as long as we take them to be place holder for sentences the resulting argument will be valid. These examples illustrate the important idea that the validity of the arguments just considered has nothing to do with the meanings of English expressions like `Jenny is miserable', `Dipan is an avid reader of Tolstoy', or any other sentence. If it has to do with meanings at all, it is with the meanings of conjunction-words like `and', `or', `not,' and `if\ldots, then\ldots'. The language of truth-functional logic is built to single out characteristic feature of these conjunction words and this will enable us to fruitfully study the idea of validity of an argument in virtue of its form.

When one introduces a language there are (at least) two task: the first is to specify the vocabulary of the language and equip the language with a grammar, that is, one has to specify how well-formed sentences of the language look like. This aspect of the language is called its \define{syntax}. The second task is to specify the \define{semantics} of the language. The semantics specifies how we are to understand the expressions of the language, what the sentences of the language mean etc. Part \ref{ch.TFL} develops both the syntax and the semantics of the language of TFL. Once this has been established we can consider how TFL may be useful for thinking about arguments in English. This will lead to the idea of symbolizing arguments in TFL and will be picked up in Part \ref{ch.TFLsymb}.

%In Parts \ref{ch.TFL}--\ref{ch.NDTFL}, we are going to develop a formal system which allows us to symbolize many arguments in such a way as to show that they are valid in virtue of their form. That language will be \emph{truth-functional logic}, or TFL. It will have sentences like $$(A\eand (B\eif\enot C)),$$ which we will read ``$A$ and if $B$, then it is not the case that $C$''.

\chapter{Syntax of TFL}\label{ch:syntfl}
\section{Atomic sentences}\label{sec:as}

We started isolating the form of an argument by replacing  \emph{subsentences} of sentences with individual letters. Thus in the first example of this section, `it is raining outside' is a subsentence of `If it is raining outside, then Jenny is miserable', and we replaced this subsentence with `$A$'.

Our artificial language, TFL, pursues this idea absolutely ruthlessly. We start with some \emph{atomic sentences}. These will be the basic building blocks out of which more complex sentences are built. We will use uppercase Roman letters for atomic sentences of TFL (except for $X$, $Y$, and $Z$ which we reserve for metavariables). There are only twenty-three letters $A$--$W$, but there is no limit to the number of atomic sentences that we might want to consider. By adding subscripts to letters, we obtain new atomic sentences. So, here are some different atomic sentences of TFL:
	$$A, B, P, P_1, P_2, A_{234}$$
You can think of atomic sentences as representing certain English sentences but for now this is simply a heuristic (in Part \ref{symb} we shall take atomic sentences to \emph{symbolize} certain English sentence).  For example, you can think of $A$ as representing the English sentence `It is raining outside', and the atomic sentence of TFL, $C$, as representing the English sentence `Jenny is miserable'.

However, if you think of the letter $P$ as representing a particular English sentence it is important to understand that whatever structure the English sentence has, atomic sentence $P$ will not reflect this structure. From the point of view of TFL, an atomic sentence is just a letter. It can be used to build more complex sentences, but it cannot be taken apart.

\newglossaryentry{atomic sentence}
{
name=atomic sentence,
description={A sentence used to represent a basic sentence; a single letter in TFL, or a predicate symbol followed by names in FOL}
}

\section{Connectives}
\label{s:TFLConnectives}

In the previous section, we introduced the atomic sentences of TFL.  In TFL we have counterparts to the conjunction-words that play an important role for spelling out arguments in English, that is, in TFL we have expression that play a similar role to the role expressions like `and', `or' and `not' play in English. These are the \emph{connectives}---they can be used to form new sentences out of old ones. In TFL, we will make use of logical connectives to build complex sentences from atomic components. There are four logical connectives in TFL. This table summarises them, and they are explained throughout this section.

\newglossaryentry{connective}
{
name=connective,
description={A logical operator in TFL used to combine \glspl{atomic sentence} into larger sentences}
}
	\begin{table}[h]
	\center
	\begin{tabular}{l l l}

	\textbf{symbol}&\textbf{what it is called}&\textbf{rough meaning}\\
	\hline
	\enot&negation&`It is not the case that$\ldots$'\\
	\eand&conjunction&`$\ldots$\ and $\ldots$',\\
	\eor&disjunction&`$\ldots$\ or $\ldots$'\\
	\eif&conditional&`If $\ldots$\ then $\ldots$'\\
	%\eiff&biconditional&`$\ldots$ if and only if $\ldots$'\\

	\end{tabular}
	\end{table}

%These are not the only connectives of English of interest. Others are, e.g., `unless', `neither \dots{} nor \dots', and `because'. We will see that the first two can be expressed by the connectives we will discuss, while the last cannot. `Because', in contrast to the others, is not \emph{truth functional}. In \S\ref{??} we shall explain what this means precisely.

If we were to substitute declarative sentences for `\dots' in the right hand column of the table above, we obtain new English sentences. The language of truth-functional logic works in the same way: the connectives `\enot',`\eand',`\eor' and `\eif' combine with sentences as introduced in \S\ref{sec:as} to form new sentences. For example, the atomic sentences `$A$' and `$C$' combine
with `$\eand$' to form the sentence `$A\eand C$'. If think of `$A$' and `$C$' as representing the English sentences
\begin{ekey}
		\item[A] It is raining outside
		\item[C] Jenny is miserable
	\end{ekey}
`$A\eand C$' can be read as:
\begin{itemize}
\item It is raining \emph{and} Jenny is miserable.
\end{itemize}
Similarly, on this understanding we get the following readings:
\begin{itemize}
\item $\enot A$: It is not the case that it is raining outside. (Alternatively, it is not raining outside).
\item $A\eor C$: It is raining outside \emph{or} Jenny is miserable.
\item $A\eif C$: \emph{If} it is raining outside, \emph{then} Jenny is miserable.
%\item $A\leftrightarrow B$: It is raining outside, \emph{if and only if} Jenny is miserable.
 \end{itemize}
 We shall go back to studying the connection between the connectives of truth-functional logic and various conjunction-words of English in \S\ref{sec:tt}, when we discuss the precise meaning of the connectives in truth-functional logic. For now we focus on completing the description of the formal language of truth-functional logic.

 To conclude our discussion of the connectives we focused on the application of the connectives to atomic sentences, but connectives can be applied to all sorts of sentences not only atomic sentences. For example, as `$\enot A$' is a TFL sentence we can use `$\eand$' to conjoin it the sentence `$C$' to form the new sentence `$(\enot A\eand C)$'. Connectives can be applied to all TFL sentences, not only atomic sentences. It is time to say precisely what TFL sentences are.

\section{Sentences}\label{s:TFLSentences}
\todo{This went through a major rewrite. Check it!!!}
We have introduced the basic building blocks of truth-functional logic, the atomic sentences, and the connectives, which allow us to conjoin different sentences to form new sentences. In terms of the \define{vocabulary} of a (written) language we have introduced all the important parts save the punctuation marks. In the language of truth-functional logic we use brackets for this purpose.

What is still missing is to equip the language with a \define{grammar}. The purpose of the grammar is to distinguish wellformed sentences from nonsense, but also to avoid ambiguity.

We wish to have rules that guarantee that
$$(A\eor (B\eand C))$$ and $$\enot (A\eand B)$$
are wellformed sentences of the language of truth-functional logic, while
$$\enot))A\eand()\enot\eor B\eif$$
is not. It is merely a sequence of symbols of TFL --- an \emph{expression} but not a \emph{sentence}. It is gibberish. 
In this respect it is like the following sequence of English words:
\begin{ebullet}
\item Green quickly the without ran ideas.
\item Not; John is happy and, not or Sue is tall only if
\end{ebullet}
These are nonsense: a string of meaningful words, but not a grammatical English sentence.



The second purpose is to avoid ambiguity. In English we use commas to distinguish between two sentences
\begin{earg}
\item[\ex{engamb1}] John's tired, and Sue's tall or Rob's short.
\item[\ex{engamb2}] John's tired and Sue's tall, or Rob's short.
\end{earg}
and without a comma it would be unclear which of the two sentences we intend to convey. In TFL this job of punctuation marks is assumed by brackets, that is, we distinguish between:
$$(A\eand(B\eor C))$$
$$((A\eand B)\eor C)$$
You can think of the former TFL-sentence as representing the English sentence 1, whereas the latter as representing the English sentence 2.

You might know this use of brackets from mathematics:
\begin{earg}
\item[\ex{mathamb}] $9 + 3 \times 4$
\end{earg}
can either be read as:
\begin{earg}
\item[\ex{mathamb1}] $9 + (3 \times 4) \qquad(=9+12=21)$
\item[\ex{mathamb2}] $(9+3) \times 4 \qquad(=12\times 4=48)$
\end{earg}

Importantly, the language of TFL is designed to exclude any form of ambiguity. For example, $A\eand B\eor C$ will not be a sentence of TFL, as it would require disambiguation. Rather the syntactic/grammatical rules of the language of TFL will be such that only expressions that have a \emph{unique} reading can be sentences of TFL. To make this precise we now provide a formal definition of what it is to be a sentence in TFL.

\factoidbox{\label{TFLsentences}
	\begin{enumerate}
		\item Every atomic sentence is a sentence.
		\item If \metaX is a sentence, then $\enot\metaX$ is a sentence.
		\item If \metaX and \metaY are sentences, then $(\metaX\eand\metaY)$ is a sentence.
		\item If \metaX and \metaY are sentences, then $(\metaX\eor\metaY)$ is a sentence.
		\item If \metaX and \metaY are sentences, then $(\metaX\eif\metaY)$ is a sentence.
		%\item If \metaX and \metaY are sentences, then $(\metaX\eiff\metaY)$ is a sentence.
		\item Nothing else is a sentence.
	\end{enumerate}
	}

	\newglossaryentry{sentence of TFL}
{
name=sentence of TFL,
description={A string of symbols in TFL that can be built up according to the recursive rules given on p.~\pageref{TFLsentences}}
}

The definition specifies rules according to which sentences of the language can be formed. To understand this definition let us pick it apart and consider the rules individually.



\begin{enumerate}
\item[1.] Tells us that atomic sentences as discussed in \S\ref{sec:as} are sentence of TFL.
\end{enumerate}
Recall that any uppercase Roman letters $A$--$W$, or with subscripts, e.g., $A_1, B_3, A_{100}, J_{375}$, are atomic sentences of TFL. Notice $X$, $Y$, and $Z$ are not atomic sentences. They are so-called \define{metavariables} and used as place holders for sentences of TFL (see more on metavairables in \S\ref{s:UseMention}).



Our second rule says:
\begin{enumerate}
\item[2.]
If $\metaX$ is a sentence of TFL, then so is $\enot \metaX$.
\end{enumerate}
By rule 1, we know that $A$ is a sentence. Rule 2 then allows us to conclude that $\enot A$ is also a sentence. We could then apply it again and conclude that $\enot\enot A$ is also a sentence. More generally, if, by whatever rule we have constructed a sentence \metaX, Rule 2 tells us that $\enot X$ will also be a sentence of TFL.

\define{Formation trees} help us keep track of this process. For the case of $\enot\enot A$ this would be:
\begin{center}
\begin{forest}
	[$\mainconnective{\enot}\enot A$
		[$\mainconnective{\enot}A$
			[$A$]
		]
	]
\end{forest}
\end{center}

Our third rule says:
\begin{enumerate}
\item[3.] If \metaX and \metaY are sentences, then so is $(\metaX\eand\metaY)$.
\end{enumerate}
By rule 1, $B_1$ and $D$ are both sentences. So rule 3 allows us to conclude that $(B_1\eand D)$ is a sentence. We might then apply rule 2 to conclude that $\enot(B_1\eand D)$ is also a sentence.
\begin{center}
\begin{forest}
	[$\mainconnective{\enot} (B_1 \eand D)$
		[$(B_1\mainconnective{\eand} D)$
			[$B_1$]
			[$D$]
		]
	]
\end{forest}
\end{center}

The rules 4 and 5 then tell us how the $\eor$- and $\eif$-connective respectively can be used to produce new sentences of TFL. Rule 6, in contrast, tells us that sentences of TFL must be formed using the rules 1-5: if an expression cannot be obtained by consecutively applying rules 1-5, then the expression is not a sentence of TFL. Again formation trees are helpful to understand this: rule 7 tells us that all nodes of the formation tree must be sentences of TFL.

For example, consider $(A \eand (B \eor C))$ we can check this is a sentence by drawing the following formation tree:
\label{S:formationtree}
\begin{center}
\begin{forest}
	[$(A\mainconnective{\eand} (B\eor C))$
		[$A$]
		[$(B\mainconnective{\eor} C)$
			[$B$]
			[$C$]
		]
	]
\end{forest}
\end{center}
Each of the steps here tracks one of the rules of what it is to be a sentence. So we can conclude that this is a sentence of TFL. This also helps us see how to read it.
It has a different formation tree from $((A\eand B)\eor C)$:
\begin{center}
\begin{forest}
	[$((A{\eand} B)\mainconnective{\eor} C))$
		[$(A\mainconnective{\eand} B)$
			[$A$]
			[$B$]
		]
		[$C$]
	]
\end{forest}
\end{center}
The different formations will be important when we describe truth-tables for these sentences (\S\ref{sec:tt}). $((A\eand B)\eor C)$ and $(A\eand (B\eor C))$ will differ in when they are true.


One more example: consider $\enot (P \eand \enot (\enot Q \eor P))$ we can check this is a sentence by drawing the following formation tree:
\label{S:formationtree}
\begin{center}
\begin{forest}
	[$\mainconnective{\enot}\,  (P \eand \enot (\enot Q \eor P))$
		[$(P \,\mainconnective{\eand}\,  \enot (\enot Q \eor P))$
			[$P$]
			[$\mainconnective{\enot}\,   (\enot Q\eor P)$
				[$\mainconnective{\enot}\,   Q$
					[$Q$]
				]
				[$P$]
			]
		]
	]
\end{forest}
\end{center}
each of the steps here tracks one of the rules of what it is to be a sentence. So we can conclude that this is a sentence of TFL. The sentences further up the tree are formed by one of the formation rules from the sentences further down the tree.

When drawing these trees we have highlighted a particular connective on each of our nodes. We call that connective the \define{main connective} of the sentence.
\factoidbox{The \define{main connective} of sentence is the last connective that was introduced in the construction of the sentence.}

In the case of $((\enot E \eor F) \eif \enot\enot G)$, the main connective is $\eif$. Here we can see that the whole sentence can be described in the form $(\metaX\eif\metaY)$ with both $\metaX$ and $\metaY$ being complete sentences (put $\metaX=(\enot E\eor F)$ and $\metaY=\enot\enot G$). That's enough to see that $\eif$ is the main connective. In the case of $\enot\enot\enot D$, the main connective is the very first $\enot$ sign. This is because we can see the sentence as having the form $\enot\metaX$ with $\metaX$ being the complete sentence $\enot\enot D$. In the case of $(P \eand \enot (\enot Q \eor R))$, the main connective is $\eand$: it's an $(\metaX\eand\metaY)$ with $\metaX$ as $P$ and $\metaY$ as $\enot (\enot Q \eor R)$.

\newglossaryentry{main connective}
{
	name=main connective,
	description={The last connective that you add when you assemble a sentence using the recursive definition.}
}

\newglossaryentry{formation tree}
{
	name=formation tree,
	description={A tree showing the structure of a sentence and its subsentences.}
}





\subsection{Inductive Definition}
The definition of a TFL-sentence is a so-called \emph{inductive} definition. Inductive definitions begin with some specifiable base elements, and then present ways to generate indefinitely many more elements by compounding together previously established ones. To give you a better idea of what an inductive definition is, we can give an inductive definition of the idea of \emph{an ancestor of mine}. We specify a base clause.
	\begin{ebullet}
		\item My parents are ancestors of mine.
	\end{ebullet}
and then offer further clauses like:
	\begin{ebullet}
		\item If x is an ancestor of mine, then x's parents are ancestors of mine.
		\item Nothing else is an ancestor of mine.
	\end{ebullet}
Using this definition, we can easily check to see whether someone is my ancestor: just check whether she is the parent of the parent of\ldots one of my parents. And the same is true for our inductive definition of sentences of TFL. Just as the inductive definition allows complex sentences to be built up from simpler parts, the definition allows us to decompose sentences into their simpler parts. Once we get down to atomic sentences, then we know we are ok.

\section{Bracketing conventions}
\label{TFLconventions}


Strictly speaking, $A\eand B$ is not a sentence of TFL. When we introduce a connective $\eand,\eor$ or $\eif$, strictly speaking, we must include brackets. Only $(A\eand B)$ is strictly speaking a sentence of TFL. The reason for this rule is that we might use $(A\eand B)$ as a subsentence in a more complicated sentence. For example, we might want to negate $(A\eand B)$, obtaining $\enot(A\eand B)$. If we just had $A \eand B$ without the brackets and put a negation in front of it, we would have $\enot A \eand B$. It is most natural to read this as meaning the same thing as $(\enot A \eand B)$, but this may be very different from $\enot(A\eand B)$.

When working with TFL, however, it will make our lives easier if we are sometimes a little less than strict. So, here are two convenient conventions.
\begin{enumerate}
\item We can remove \emph{outermost} brackets of a sentence. Thus we allow ourselves to write $A\eand B$ instead of the sentence $(A\eand B)$. However, we must remember to put the brackets back in, when we want to embed the sentence into a more complicated sentence!
\item It can be a bit painful to stare at long sentences with many nested pairs of brackets. To make things a bit easier on the eyes, we will  allow ourselves to use square brackets, `[' and `]', instead of rounded ones. So there is no logical difference between $(P\eor Q)$ and $[P\eor Q]$, for example.
\todo{Add in the rule that arrow binds stronger}
\end{enumerate}

Combining these two conventions, we can rewrite the unwieldy sentence
$$(((H \eif I) \eor (I \eif H)) \eand (J \eor K))$$
rather more clearly as follows:
$$\bigl[(H \eif I) \eor (I \eif H)\bigr] \eand (J \eor K)$$
The scope of each connective is now much easier to pick out.

\begin{practiceproblems}

\solutions

\problempart 
Add brackets to form a sentence of TFL, strictly speaking (i.e., no dropped brackets), which satisfies the given condition. 
\begin{center}
	\begin{tabular}{>{\centering}p{.5em}>{\centering}p{.5em}>{\centering}p{.5em}>{\centering}p{.5em}>{\centering}p{.5em}>{\centering}p{.5em}>{\centering}p{.5em}>{\centering}p{.5em}>{\centering}p{.5em}}
		\enot &$P$&\eor &$Q$&\eif &\enot &$P_2$&\eand &$R$
	\end{tabular}
\end{center}

\begin{earg}
	\item Where the `$\eif$' is the main connective
	\item Where the `$\eor$' is the main connective.
	\item Where the left `$\enot$' is the main connective. 
	\item Can it be done so that the right `$\enot$' is the main connective?
\end{earg}


\problempart
\label{pr.wiffTFL}
For each of the following: (a) Is it a sentence of TFL, strictly speaking? (b) Is it a sentence of TFL, allowing for our relaxed bracketing conventions? (c) If the answer to (b) is yes, write down the formation tree of each sentence and determin the main connective at each node (if there is one). Is there a main connective for every node of the formation tree of a sentence.
\begin{earg}
\item $(A)$\hfill \myanswer{(a) no (b) no}
\item $J_{374} \eor \enot J_{374}$\hfill \myanswer{(a) no (b) yes}
\item $\enot \enot \enot \enot F$\hfill \myanswer{(a) yes (b) yes}
\item $\enot \eand S$\hfill \myanswer{(a) no (b) no}
\item $(G \eand \enot G)$\hfill \myanswer{(a) yes (b) yes}
\item $(A \eif (A \eand \enot F)) \eor (D \eif E)$\hfill \myanswer{(a) no (b) yes}
\item $[(Z \eif S) \eif W] \eand [J \eor X]$\hfill \myanswer{(a) no (b) yes}
\item $(F \eif \enot D \eif J) \eor (C \eand D)$\hfill \myanswer{(a) no (b) no}
\end{earg}


\problempart
\begin{earg}
	\item  Does $\enot A\eor B$ have the form $\metaX\eor\metaY$?
	\item Does $\enot(A\eor B)$ have the form $\metaX\eor \metaY$?
	\item Does $(A\eor B)\eor C$ have the form $\metaX\eor\metaY$?
\end{earg}
\problempart
\begin{enumerate}
	\item Can there be a sentence of TFL of the form $(\metaX\eand\metaY)$ that includes three different connectives (the connectives are $\enot,\eand,\eor,\eif$)? If so, give an example, if not explain why not.
	\myanswer{\\ yes, for example $(\enot A\eand(B\eor C))$}
	\item Are there any sentences of TFL that contain no atomic sentences? Explain your answer.
	\\\myanswer{No. Atomic sentences contain atomic sentences (trivially). And every more complicated sentence is built up out of less complicated sentences, that were in turn built out of less complicated sentences, \ldots, that were ultimately built out of atomic sentences.}
			\item Are there sentences of TFL that don't have a main connective? Explain your answer.\myanswer{Yes, atomic sentences don't have a main connective.}
\end{enumerate}

\end{practiceproblems}


\chapter{Use and mention}\label{s:UseMention}
We have talked a lot \emph{about} sentences. So we should pause to explain an important, and very general, point.

\section{Quotation conventions}
Consider these two sentences:
	\begin{itemize}
		\item Justin Trudeau is the Prime Minister.
		\item The expression `Justin Trudeau' is composed of two uppercase letters and eleven lowercase letters
	\end{itemize}
When we want to talk about the Prime Minister, we \emph{use} his name. When we want to talk about the Prime Minister's name, we \emph{mention} that name, which we do by putting it in quotation marks.

There is a general point here. When we want to talk about things in the world, we just \emph{use} words. When we want to talk about words, we typically have to \emph{mention} those words. We need to indicate that we are mentioning them, rather than using them. To do this, some convention is needed. We can put them in quotation marks, or display them centrally in the page (say). So this sentence:
	\begin{itemize}
		\item `Justin Trudeau' is the Prime Minister.
	\end{itemize}
says that some \emph{expression} is the Prime Minister. That's false. The \emph{man} is the Prime Minister; his \emph{name} isn't. Conversely, this sentence:
	\begin{itemize}
		\item Justin Trudeau is composed of two uppercase letters and eleven lowercase letters.
	\end{itemize}
also says something false: Justin Trudeau is a man, made of flesh rather than letters. One final example:
	\begin{itemize}
		\item ``\,`Justin Trudeau'\,'' is the name of `Justin Trudeau'.
	\end{itemize}
On the left-hand-side, here, we have the name of a name. On the right hand side, we have a name. Perhaps this kind of sentence only occurs in logic textbooks, but it is true nonetheless.

Those are just general rules for quotation, and you should observe them carefully in all your work! To be clear, the quotation-marks here do not indicate reported speech. They indicate that you are moving from talking about an object, to talking about a name of that object.



\section{Object language and metalanguage}
These general quotation conventions are very important for us. After all, we are describing a formal language here, TFL, and so we must often \emph{mention} expressions from TFL.

When we talk about a language, the language that we are talking about is called the \define{object language}. The language that we use to talk about the object language is called the \define{metalanguage}.
\label{def.metalanguage}
\newglossaryentry{object language}
{
name=object language,
description={A language that is constructed and studied by logicians. In this textbook,
 the object languages are TFL and FOL}
}

\newglossaryentry{metalanguage}
{
name=metalanguage,
description={The language logicians use to talk about the object language. In this textbook, the metalanguage is English, supplemented by certain symbols like metavariables and technical terms like ``valid''}
}

For the most part, the object language in this chapter has been the formal language that we have been developing: TFL. The metalanguage is English. Not conversational English exactly, but English supplemented with some additional vocabulary to help us get along.

Now, we have used uppercase letters as sentence letters of TFL:
	$$A, B, C, Z, A_1, B_4, A_{25}, J_{375},\ldots$$
These are sentences of the object language (TFL). They are not sentences of English. So we must not say, for example:
	\begin{itemize}
		\item $D$ is a sentence letter of TFL.
	\end{itemize}
Obviously, we are trying to come out with an English sentence that says something about the object language (TFL), but `$D$' is a sentence of TFL, and not part of English. So the preceding is gibberish, just like:
	\begin{itemize}
		\item \foreignlanguage{german}{Schnee ist weiß} is a German sentence.
	\end{itemize}
What we surely meant to say, in this case, is:
	\begin{itemize}
		\item `\foreignlanguage{german}{Schnee ist weiß}' is a German sentence.
	\end{itemize}
Equally, what we meant to say above is just:
	\begin{itemize}
		\item `$D$' is a sentence letter of TFL.
	\end{itemize}
The general point is that, whenever we want to talk in English about some specific expression of TFL, we need to indicate that we are \emph{mentioning} the expression, rather than \emph{using} it. We can either deploy quotation marks, or we can adopt some similar convention, such as  placing it centrally in the page.


\section{Metavariables}\label{s:Metavariables}
However, we do not just want to talk about \emph{specific} expressions of TFL. We also want to be able to talk about \emph{any arbitrary} sentence of TFL. Indeed, we had to do this in \S\ref{s:TFLSentences}, when we presented the recursive definition of a sentence of TFL. We used uppercase script letters to do this, namely:
	$$\metaX, \metaY, \metaZ,\metaX_1,\metaY_1,\metaZ_1\ldots$$
These symbols do not belong to TFL. Rather, they are part of our (augmented) metalanguage that we use to talk about \emph{any} expression of TFL. To explain why we need them, recall the second clause of the recursive definition of a sentence of TFL:
	\begin{itemize}
		\item[2.] If $\metaX$ is a sentence, then $\enot \metaX$ is a sentence.
	\end{itemize}
This talks about \emph{arbitrary} sentences. If we had instead offered:
	\begin{itemize}
		\item[2$'$.] If `$A$' is a sentence, then `$\enot A$' is a sentence.
	\end{itemize}
this would not have allowed us to determine whether `$\enot B$' is a sentence. To emphasize:
	\factoidbox{
	  `$\metaX$' is a symbol (called a \define{metavariable}) in augmented English, which we use to talk about expressions of TFL. 	`$A$' is a particular sentence letter of TFL.}

        \newglossaryentry{metavariables}
{
name=metavariables,
description={A variable in the metalanguage that can represent any sentence in the object language}
}
But this last example raises a further complication, concerning quotation conventions. We did not include any quotation marks in the second clause of our inductive definition. Should we have done so?

The problem is that the expression on the right-hand-side of this rule, i.e., `$\enot\metaX$', is not a sentence of English, since it contains~`$\enot$'. So we might try to write:
	\begin{itemize}
		\item[2$''$.] If \metaX is a sentence, then `$\enot \metaX$' is a sentence.
	\end{itemize}
But this is no good: `$\enot \metaX$' is not a TFL sentence, since `$\metaX$' is a symbol of (augmented) English rather than a symbol of TFL.

What we really want to say is something like this:
	\begin{itemize}
		\item[2$'''$.] If \metaX is a sentence, then the result of placing the symbol `$\enot$' in front of the sentence \metaX is a sentence.
	\end{itemize}
	This is impeccable, but rather long-winded. %Quine introduced a convention that speeds things up here. In place of (2$''$), he suggested:
%	\begin{numberlist}
%		\item[2$'''$.] If \metaX and \metaY are sentences, then $\ulcorner (\metaX\eand\metaY)\urcorner$ is a sentence
%	\end{numberlist}
%The rectangular quote-marks are sometimes called `Quine quotes', after Quine. The general interpretation of an expression like `$\ulcorner (\metaX\eand\metaY)\urcorner$' is in terms of rules for concatenation.
But we can avoid long-windedness by creating our own conventions. We can perfectly well stipulate that an expression like `$\enot \metaX$' should simply be read \emph{directly} in terms of rules for concatenation. So, \emph{officially}, the metalanguage expression `$\enot \metaX$'
simply abbreviates:
\begin{itemize}
\item[]	the result of placing the symbol `$\enot$' in front of the sentence \metaX
\end{itemize}
and similarly, for expressions like `$(\metaX \eand \metaY)$', `$(\metaX \eor \metaY)$', etc.


\section{Quotation conventions for arguments}
One of our main purposes for using TFL is to study arguments, and that will be our concern in \S\ref{sec:tt}. In English, the premises of an argument are often expressed by individual sentences, and the conclusion by a further sentence. Since we can symbolize English sentences, we can symbolize English arguments using TFL.

Or rather, we can use TFL to symbolize each of the \emph{sentences} used in an English argument. However, TFL itself has no way to flag some of them as the \emph{premises} and another as the \emph{conclusion} of an argument.  (Contrast this with natural English, which uses words like `so', `therefore', etc., to mark that a sentence is the \emph{conclusion} of an argument.)

%So, if we want to symbolize an \emph{argument} in TFL, what are we to do?

%An obvious thought would be to add a new symbol to the \emph{object} language of TFL itself, which we could use to separate the premises from the conclusion of an argument. However, adding a new symbol to our object language would add significant complexity to that language, since that symbol would require an official syntax.\footnote{\emph{The following footnote should be read only after you have finished the entire book!} And it would require a semantics. Here, there are deep barriers concerning the semantics. First: an object-language symbol which adequately expressed `therefore' for TFL would not be truth-functional. (\emph{Exercise}: why?) Second: a paradox known as `validity Curry' shows that FOL itself \emph{cannot} be augmented with an adequate, object-language `therefore'.}

So, we need another bit of notation. Suppose we want to symbolize the premises of an argument with $\metaX_1$, \dots,~$\metaX_n$ and the conclusion with $\metaZ$. Then we will write:
$$\metaX_1, \ldots, \metaX_n \therefore \metaZ$$
The role of the symbol `$\therefore$' is simply to indicate which sentences are the premises and which is the conclusion.

%Strictly, this extra notation is \emph{unnecessary}. After all, we could always just write things down long-hand, saying: the premises of the argument are well symbolized by $\metaX_1, \ldots \metaX_n$, and the conclusion of the argument is well symbolized by $\metaZ$. But having some convention will save us some time. Equally, the particular convention we chose was fairly \emph{arbitrary}. After all, an equally good convention would have been to underline the conclusion of the argument. Still, this is the convention we will use.

Strictly, the symbol `$\therefore$' will not be a part of the object language, but of the \emph{metalanguage}. As such, one might think that we would need to put quote-marks around the TFL-sentences which flank it. That is a sensible thought, but adding these quote-marks would make things harder to read. Moreover---and as above---recall that \emph{we} are stipulating some new conventions. So, we can simply stipulate that these quote-marks are unnecessary. That is, we can simply write:
$$A, A \eif B \therefore B$$
\emph{without any quotation marks}, to indicate an argument whose premises are (symbolized by) `$A$' and `$A \eif B$' and whose conclusion is (symbolized by)~`$B$'.


\begin{practiceproblems}
\problempart
Add quotation marks to the following sentences where necessary:
\begin{enumerate}
\item \myanswer{`}$\enot$\myanswer{'}, \myanswer{`}$\eand$\myanswer{'}, \myanswer{`}$\eor$\myanswer{'} and \myanswer{`}$\eif$\myanswer{'} are the basic logical connectives of the language of TFL.
\item In a well-formed sentence for every opening bracket `(' there must be a closing bracket \myanswer{`})\myanswer{'}.
\item The sentence \myanswer{`}snow is white\myanswer{'} is true if and only if snow is white. 
\item Kevin is a lecturer, whereas \myanswer{`}Kevin\myanswer{'} names Kevin and the latter is an expression of the language (and not a lecturer).
\end{enumerate}

\problempart
Given our convention on metavariables what do the following expressions convey:
\begin{enumerate}
\item $\metaX\eand\metaY$\myanswer{\\the result of taking the sentence \metaY followed by `$\eand$' followed by \metaZ}
\item $\metaY\eif\metaZ$\myanswer{\\the result of taking the sentence \metaY followed by `$\eif$' followed by \metaZ}
\item $\enot(\metaY\eor\metaZ)$\myanswer{\\the result of placing `$\enot($' before the sentence \metaY, then adding `$\eor$', then \metaZ and finally `$)$'.}\todo{are these right???}
\end{enumerate}

\problempart
Consider the sentence
\begin{itemize}
\item Sharky is so-called because of his teeth.
\end{itemize}
Is `Sharky' used or mentioned in the above sentence? Discuss.
\end{practiceproblems}





\chapter{Truth rules for the connectives}\label{sec:tt}

We have completed introducing the syntax of the language of TFL. It is now time to turn to the semantics of TFL. The idea underlying the semantics is to specify conditions under which sentences of TFL are true. In \S\ref{ch:syntfl} we introduced the idea that an argument is valid if there is no interpretation on which all premises of the argument are true but the conclusion false. Accordingly, if we wish to make a start on making this idea more precise, we need to say when a sentence of TFL is true according to an interpretation and when it is false. The target of this chapter is to give precise rules to this effect. The important feature of truth functional logic is that the truth value of a complex sentence, such as `$A\eor(B\eand C)$' is fully determined by the truths of is component parts, that is `$A$', `$B$' and `$C$'. If we're told whether `$A$', `$B$' and $C$ are true or false, then we will be able to say whether `$A\eor(B\eand C)$' is true or false.

To spell out this idea, we need to describe how the truth values of different sentences (e.g., `$A$', `$B$', and `$C$') are to be combined to obtain the truth value of a sentence  that has been obtained via the formation rules 2-5 (e.g., `$A\eor(B\eand C)$'). To do this we work through each of our connectives describing the rules governing it.

\newglossaryentry{truth value}
{
	name = truth value,
	description = {One of the two logical values sentences can have: True and False}
}


\section{Negation}
The `$\enot$'-connective is called negation. When we introduced the `$\enot$'-connective we said it should roughly be understood as `it is not the case' or, perhaps, simply `not'. Let's make that official:
\factoidbox{
If a sentence can be paraphrased as `it is not the case that \ldots'\\ it can be symbolised as $\enot\metaX$.
}
What does that mean for the truth rules? Consider:
\begin{earg}
\item[\ex{neg-f}] Bristol is not in France.
\item[\ex{neg-f}] Bristol is not in England.
\end{earg}
`Bristol is in France' is false, so `Bristol is not in France' is true.
`Bristol is in England' is true, so `Bristol is not in England' is false.

In general, to determine whether a sentence of the form $\enot \metaX$ is true we check whether $\metaX$ is false. If the answer is yes, then $\enot \metaX$ is indeed true. If to the contrary  $\metaX$ is true, then $\enot \metaX$ is false:
\begin{itemize}
\item If $\metaX$ is true, then $\enot \metaX$ is false.
\item If $\metaX$ is false, then $\enot \metaX$ is true.
\end{itemize}
We record this in shorthand:
\begin{highlighted}
\begin{center}
\begin{tabular}{ccc}
If \metaX is: &&then \enot \metaX is:\\
T &$\leadsto$& F\\
F &$\leadsto$& T
\end{tabular}
\end{center}
\end{highlighted}
We have abbreviated `True' with `T' and `False' with `F'. (But just to be clear, the two truth values are True and False; the truth values are not \emph{letters}!)
\newglossaryentry{negation}
{
name=negation,
description={The symbol \enot, used to represent words and phrases that function like the English word ``not''}
}






\section{Conjunction}
\label{s:ConnectiveConjunction}
The $\eand$-connective, called conjunction, is meant to be understood of the English word `and':
\factoidbox{
%		A sentence can be symbolized as $(\metaX\eand\metaY)$ if it can be paraphrased in English as `Both\ldots, and\ldots', or as `\ldots, but \ldots', or as `although \ldots, \ldots'.
		If a sentence can be paraphrased as `\ldots and\ldots' \\it can be symbolised as $(\metaX\eand\metaY)$.
	}

What is the appropriate truth rule for conjunction? Consider:
\begin{earg}
\item[\ex{conj}] She can speak German and she can speak French.
\end{earg}
If she can speak German and she can speak French, then this is true, but otherwise it is false.

More generally, the rule governing $\eand$ is:
\begin{itemize}
\item If $\metaX$ and $\metaY$ are both true, then $\metaX\eand\metaY$ is true.
\item Otherwise, $\metaX\eand \metaY$ is false.
\end{itemize}
Which we summarise
\begin{highlighted}
\begin{center}
\begin{tabular}{cccc}
If \metaX is:&and \metaY is:&&then $(\metaX\eand\metaY)$ is:\\
T & T &$\leadsto$& T\\
T & F &$\leadsto$& F\\
F & T &$\leadsto$& F\\
F & F &$\leadsto$& F
\end{tabular}
\end{center}
\end{highlighted}
%in the characteristic truth table:
%\begin{center}
%\begin{tabular}{c c |c}
%\metaX & \metaY & $\metaX\eand\metaY$\\
%\hline
%T & T & T\\
%T & F & F\\
%F & T & F\\
%F & F & F
%\end{tabular}
%\end{center}
Note that conjunction is \emph{symmetrical}. The truth value for $(\metaX \eand \metaY)$ is always the same as the truth value for $(\metaY \eand \metaX)$.

\newglossaryentry{conjunct}
{
name=conjunct,
description={A sentence joined to another by a \gls{conjunction}}
}


\section{Disjunction}
The `$\eor$'-connective is called disjunction and is meant to be understood in terms of the English `or'.
\newglossaryentry{disjunction}
{
name=disjunction,
description={The connective \eor, used to represent words and phrases that function like the English word ``or'' in its inclusive sense; or a sentence formed by using this connective}
}

\newglossaryentry{disjunct}
{
name=disjunct,
description={A sentence joined to another by a \gls{disjunction}}
}
\factoidbox{
If a sentence can be paraphrased as `\ldots or \ldots' \\it can be symbolised as $(\metaX\eor\metaY)$.
	}

Whereas the truth rules for the negation and conjunction were relatively straightforward, the rule for disjunction is a bit more subtle.
Consider:
\begin{earg}
\item[\ex{conj}] She can speak German or she can speak French.
\end{earg}
If she cannot speak either German or French, then this is false. If she can speak German but not French, then it is true, and if she can speak French but not German it is also true.
We have the general rules:
\begin{itemize}
\item If $\metaX$ and $\metaY$ are both false, then $(\metaX\eor\metaY)$ is false.
\item  If $\metaX$ is true and $\metaY$ is false, then $(\metaX\eor\metaY)$ is true.
\item  If $\metaX$ is false and $\metaY$ is true, then $(\metaX\eor\metaY)$ is true.
\end{itemize}

But what if she can speak both? Is it true or false?
It seems that in English there are two kinds of disjunctions: an \define{inclusive} and an \define{exclusive} one. For the inclusive \emph{or}, we might whisper a ``or both'' after it; whereas for the exclusive \emph{or}, we'd want to whisper a ``but not both'':
\begin{earg}
\item [\ex{inclor}] She can speak German or
she can speak French (or both).
\item [\ex{exclor}] She can speak German or
she can speak French (but not both).
\end{earg}

In logic there can be no ambiguity. We choose that $\eor$ stands for the \emph{inclusive or}.
That is, we give the final rule:
\begin{itemize}
\item If $\metaX$ and $\metaY$ are both true, then $(\metaX\eor\metaY)$ is true.
\end{itemize}

%So, when we turn to symbolisations of arguments, one should only symbolise a sentence as $(\metaX\eor \metaY)$ if it is to be read as the \emph{inclusive or}. 

Once we have completed our presentation of the semantics for. TFL we will see that more complex sentence: $((\metaX\eor\metaY)\eand \enot(\metaX\eand\metaY))$ has the truth conditions of the \emph{exclusive or}, that is it is true if $\metaX$ is true or $\metaY$ is true but not both $\metaX$ and $\metaY$ are true.
%Sometimes the English is ambiguous, in which case one should point that out when symbolising and give the two alternative symbolisations.

%Consider:
%\begin{earg}
%\item[\ex{exclor}]She either ate pizza or pasta.
%\end{earg}
%Maybe this is an exclusive or, though an alternative treatment is to take this to be an inclusive or, but note that there's a implicit, or missing, premise: that she didn't eat both.

To summarise the rules for $\eor$:
%\begin{center}
%\begin{tabular}{c c |c}
%\metaX & \metaY & $\metaX\eor\metaY$\\
%\hline
%T & T & T\\
%T & F & T\\
%F & T & T\\
%F & F & F
%\end{tabular}
%\end{center}
\begin{highlighted}
\begin{center}
\begin{tabular}{cccc}
If \metaX is:&and \metaY is:&&then $(\metaX\eor\metaY)$ is:\\
T & T &$\leadsto$& T\\
T & F &$\leadsto$& T\\
F & T &$\leadsto$& T\\
F & F &$\leadsto$& F
\end{tabular}
\end{center}
\end{highlighted}
Like conjunction, disjunction is symmetrical.

\section{Conditional}\todo{Say more about ``only if''! Students thought it was $\eiff$}
The `$\eif$'-connective is called the conditional-connective and it is meant to be related to our understanding of \emph{if\ldots, then\ldots} sentences. 
Unlike conjunction and disjunction, the conditional is asymmetric, it matters which order the sentences come in, and $(\metaX\eif\metaY)$ will be different to $(\metaY\eif\metaX)$. It is thus valuable to give a name to indicate which part goes where:
\begin{highlighted}
	In $\metaX\eif\metaY$, \metaX is called the \define{antecedent}, and \metaY is called the \define{consequent}.
\end{highlighted}

\newglossaryentry{antecedent}
{
	name=antecedent,
	description={The  \define{antecedent} of a conditional $\metaX\eif\metaY$ is \metaX.}
}
\newglossaryentry{consequent}
{
	name=consequent,
	description={The  \define{consequent} of a conditional $\metaX\eif\metaY$ is \metaY.}
}

Recall our guidance on the rough meaning of $\eif$:

	\factoidbox{
	 If a sentence can be paraphrased as \\`If \ldots, then \ldots' it can be symbolised as $(\metaX\eif\metaY)$.
	}




\newglossaryentry{conditional}
{
name=conditional,
description={The symbol \eif, used to represent words and phrases that function like the English phrase ``if \ldots{} then \ldots''; a sentence formed by using this symbol}
}
What are the truth rules for the conditional? Consider the sentence:
\begin{earg}
\item[\ex{bartender}] If she is drinking a beer, then she is over eighteen.
\end{earg}
What are the circumstances under which this conditional is false?
Here is what we'll say:
\begin{itemize}
\item If she's drinking beer and is under age, then it is false.
\item It is true in all other circumstances.
\end{itemize}
To understand the rationale for this, let us think about when a bartender would get into trouble (clearly the conditional should be true for everyone drinking beer in a bar). They will get into trouble in case there a woman in drinking beer and it turns out that the woman is under age. They will not get into trouble if the woman is over 18 years old, i.e., if the consequent of the conditional is true. They will also not get in trouble if it turns out that the woman is not drinking beer (but orange juice), i.e., if the antecedent of the conditional is false.  In that case it is irrelevant whether she is 18 or not. The barkeeper doesn't have to check her age. The conditional is true for trivial reasons.


%We get the characteristic truth table:
%\begin{center}
%\begin{tabular}{c c|c}
%\metaX & \metaY & $\metaX\eif\metaY$\\
%\hline
%T & T & T\\
%T & F & F\\
%F & T & T\\
%F & F & T
%\end{tabular}
%\end{center}
We are led to the following rules:
\begin{highlighted}
\begin{center}
\begin{tabular}{cccc}
If \metaX is:&and \metaY is:&&then $(\metaX\eif\metaY)$ is:\\
T & T &$\leadsto$& T\\
T & F &$\leadsto$& F\\
F & T &$\leadsto$& T\\
F & F &$\leadsto$& T
\end{tabular}
\end{center}
\end{highlighted}
In this case, it's very important to remember which way around it goes. The TF-line is different to the FT-line.

This is why the terms `antecedent' and `consequent' are so useful.  
In $\metaX\eif\metaY$, \metaX is called the \define{antecedent}, and \metaY the \define{consequent}.
Using the antecedent/consequent terminology we can summarize the truth rule as follows:
If the antecedent is true and the consequent false, then the conditional sentence is false, otherwise it is true.
\newglossaryentry{antecedent}
{
name=antecedent,
description={The sentence on the left side of a \gls{conditional}}
}


\newglossaryentry{consequent}
{
name=consequent,
description={The sentence on the right side of a \gls{conditional}}
}

The TFL connective $\eif$ is \emph{stipulated} to be governed by these rules. This sometimes marked by calling it the \define{material conditional}. But the truth rules of the $\eif$-connective only tell us part of the story of \emph{if\ldots, then\ldots} sentences in English, as the truth rules do not seem to work well for all these sentences. For example, truth rules for material conditional do not seem to work well with our understanding of the sentence
\begin{earg}
\item[\ex{kangaroo}] If Kangaroos had no tails, they would topple over.
\end{earg}
%We will discuss conditional-sentences of this kind in \S\ref{s:IndicativeSubjunctive} and will look at some problems arising due to understanding $\eif$ in terms of material conditional in \S\ref{s:ParadoxesOfMaterialConditional}.
%
For our purposes $\eif$-connective will be understood in terms of truth rules given in this section.

\section{On Truth-functional connectives}
\label{s:TruthFunctionality}
%We now reflect on truth-functional logic and the connectives we've used.
%\section{Non truth-functional connectives}

%%%%because???
Let's introduce an important idea.
\factoidbox{
	A connective is \define{truth-functional} iff the truth value of a sentence with that connective as its main connective is uniquely determined by the truth value(s) of the constituent sentence(s).
}
\newglossaryentry{truth-functional connective}
{
	name=truth-functional connective,
	description={an operator that builds larger sentences out of smaller ones and fixes the \gls{truth value} of the resulting sentence based only on the truth value of the component sentences}
}

Every connective in TFL is truth-functional. We were able to give rules to determine what the truth value of a sentence $\enot\metaX$ is depending only on the truth value of $\metaX$. The truth value of $\metaX$ uniquely determines the truth value of $\enot \metaX$. The same was true for all the other connectives of TFL ($\eand,\eor,\eif$), as is evidenced by the truth rules and truth tables we gave for these connectives. This is what gives TFL its name: it is \emph{truth-functional logic}.

%The truth value of a negation is uniquely determined by the truth value of the unnegated sentence. The truth value of a conjunction is uniquely determined by the truth value of both conjuncts. The truth value of a disjunction is uniquely determined by the truth value of both disjuncts, and so on.

%This then means that to determine the truth value of any TFL sentence, we only need to know the truth value of the atomic sentences it includes. We will see exactly how to do this in \S\ref{s:CompleteTruthTables}.


%The truth value of a non-atomic sentence of TFL, such as $A\eand B$ depends on the truth values of $A$ and $B$. But once the truth values of $A$, $B$ and $C$ are provided, the truth value of $A\eand B$ is fixed. Indeed, this is the characteristic feature of \emph{truth functional logic}.


In plenty of languages, e.g.~English, there are connectives that are not truth-functional. We here describe just two (Exercise: find further non-truth-functional connectives of English):

\subsection{Necessarily}

In English, for example, we can form a new sentence from any simpler sentence by prefixing it with `Necessarily, \ldots' or  `It is necessary that\ldots'. The truth value of this new sentence is not fixed solely by the truth value of the original sentence. For consider two true sentences:
\begin{earg}
	\item[\ex{nec-math}] $2 + 2 = 4$
	\item[\ex{nec-music}] Shostakovich wrote fifteen string quartets
\end{earg}
Whereas it is necessary that $2 + 2 = 4$, it is not \emph{necessary} tthat Shostakovich wrote fifteen string quartets. If Shostakovich had died earlier, he would have failed to finish Quartet no.\ 15; if he had lived longer, he might have written a few more. So `It is necessary that\ldots' is a connective of English, but it is not \emph{truth-functional}.



\subsection{Subjunctive conditionals}\label{s:IndicativeSubjunctive}



We said that $\eif$ was pretty bad at capturing some uses of `if\ldots, then\ldots' in Englihs. In particular, it is bad at capturing use of \emph{subjunctive conditionals} of English. The problem is that a subjunctive conditional is not truth functional.
Consider the two sentences:
\begin{earg}
	\item[\ex{brownwins1}] If Mitt Romney had won the 2012 election, then he would have been the 45th President of the USA.
	\item[\ex{brownwins2}] If Mitt Romney had won the 2012 election, then he would have turned into a helium-filled balloon and floated away into the night sky.
\end{earg}
Sentence \ref{brownwins1} is true; sentence \ref{brownwins2} is false, but both have false antecedents and false consequents. So the truth value of the whole sentence is not uniquely determined by the truth value of the constituent sentences.

However, the material conditional, which is what we specify $\eif$ to be is the best that can be done at symbolising subjunctive conditionals of English in TFL. TFL just doesn't have the required resources as the subjunctive conditional is not truth-functional.



\begin{practiceproblems}
	
	
	\problempart 
\begin{enumerate}
	\item Suppose $\metaX$ is true, what can you say about $\metaX\eor\metaY$?
\item Suppose $\metaX$ is false, what can you say about $\metaX\eif \metaY$?
\end{enumerate}


\problempart
\begin{enumerate}
	\item 	Is `I know that' truth functional? \\
%	\emph{(Hint: try to think of a situation with two English sentences, $\metaX$ and $\metaY$, which have the same truth value but where `I know that \metaX' and `I know that \metaY' have different truth values) }
	\myanswer{\\No. \begin{center}
			`I know that I have hands'
		\end{center} is true (true $\leadsto$ true). But there are some sentences that are true but I don't know them. E.g., 
		\begin{center}
			`I know that France won the FIFA world cup'
		\end{center} (well, I now know it but I didn't 20 minutes ago. )}
\end{enumerate}
\end{practiceproblems}



\chapter{Truth on a Valuation}
In the previous section we have learned how truth values of constituent sentences determine the truth value of complex sentences. This means that if we are presented with the truth value of the relevant atomic sentences that appear in a complex sentence, we can determine whether that sentence is true or false.

But how do we determine whether a given atomic sentence is true? This is where the notion of an \emph{interpretation} comes into the picture. An interpretation will stipulate (assign) truth values of particular atomic sentences. Let `$B$' stand for the English sentence `Ben is happy'. Then there is one interpretation according to which this is true, that is, $B$ will be assigned the value ``True'' on this interpretation. There is another interpretation according to which Ben is not happy, that is, $B$ is assigned the value ``False'' on this interpretation. In TFL an interpretation of the atomic sentences is called a \define{valuation}:

\factoidbox{
		A \define{valuation} is any assignment of truth values (true or false) to the atomic sentences of TFL.
	}

        \newglossaryentry{valuation}
{
name=valuation,
description={An assignment of \glspl{truth value} to particular atomic \glspl{sentence of TFL}}
}


For example a valuation $v$ might assign:
\begin{center}
	\begin{tabular}{rccccccc}
	&	$A$&$H$&$I$&$B$&$C_{35}$&\ldots\\\hline
	$v$:&	F & T &F&F&F&T&\ldots
	\end{tabular}
\end{center}

Sentences of TFL are true or false \emph{relative to a valuation}. 
The valuation determines the truth values of the atomic sentences, and complex sentences are determined using the truth rules described in the previous section. 
 \factoidbox{\label{Truthint}Let $v$ be a valuation. Then
	\begin{enumerate}
		\item An atomic sentence $X$ is true relative to $v$, if and only if $v$ assigns the value T to $X$.
		\item For a sentence $\enot\metaX$, $(\metaX\eand\metaY)$, $(\metaX\eor\metaY)$, $(\metaX\eif\metaY)$, whether it is true or false relative to $v$ can be determined by applying the appropriate truth rule:
%			\begin{tabular}{@{If \metaX is }c@{ relative to $v$, then $\enot\metaX$ is }c@{ relative to $v$}}
%				true &false\\
%				false &true
%			\end{tabular}
		\begin{center}
			\begin{tabular}{ccc}
				If \metaX is:&&then $\enot\metaX$ is:\\
				T &$\leadsto$& F\\
				F &$\leadsto$& T
			\end{tabular}
		\end{center}
		\begin{center}
			\begin{tabular}{cccc}
				If \metaX is:&and \metaY is:&&then $(\metaX\eand\metaY)$ is:\\
				T & T &$\leadsto$& T\\
				T & F &$\leadsto$& F\\
				F & T &$\leadsto$& F\\
				F & F &$\leadsto$& F
			\end{tabular}
		\end{center}
		\begin{center}
			\begin{tabular}{cccc}
				If \metaX is:&and \metaY is:&&then $(\metaX\eor\metaY)$ is:\\
				T & T &$\leadsto$& T\\
				T & F &$\leadsto$& T\\
				F & T &$\leadsto$& T\\
				F & F &$\leadsto$& F
			\end{tabular}
		\end{center}
		\begin{center}
			\begin{tabular}{cccc}
				If \metaX is:&and \metaY is:&&then $(\metaX\eif\metaY)$ is:\\
				T & T &$\leadsto$& T\\
				T & F &$\leadsto$& F\\
				F & T &$\leadsto$& T\\
				F & F &$\leadsto$& T
			\end{tabular}
		\end{center}
%		\item a sentence $\enot\metaX$ is true relative to $v$, if and only if $\metaX$ is false relative to $v$.
%		\item a sentence $(\metaX\eand\metaY)$ is true relative to $v$, if and only if $\metaX$ and $\metaY$ are both true relative to $v$.
%		\item a sentence $(\metaX\eor\metaY)$ is true relative to $v$, if and only if $\metaX$ or $\metaY$is true relative to $v$.
%		\item a sentence $(\metaX\eif\metaY)$  is true relative to $v$, if and only if $\metaX$ is false or $\metaY$ is true relative to $v$
		%\item If \metaX and \metaY are sentences, then $(\metaX\eiff\metaY)$ is a sentence.
		%\item Nothing else is a sentence.
	\end{enumerate}
}

Consider 
`$(\enot I\eand H)\eif H$'. We will determine whether it is true or false on the valuation given. 
The idea is that we draw the formation tree and then work our way up the formation tree to, in stages, determine the value of our complex sentence.
This is similar to as one might do in mathematics to determine the value of a complex equation in stages starting from the bottom of the tree:
\begin{center}
	\begin{forest}
		[{$7\mainconnective{+}(5\times\sqrt{}4)$}, label=right:{\scriptsize $=17$}
		[{$7$}]
		[{$(5\mainconnective{\times} \sqrt{}4)$}, label=right:{\scriptsize $=10$}
		[{$5$}]
		[{$\mainconnective{\sqrt{}}{4}$}, label=right:{\scriptsize $=2$}
		[{$4$}]
		]
		]
		]
	\end{forest}
\end{center}

Or, it might be more similar to this maths question: Setting $x=7$, $y=5$ and $z=4$,  determine the value of $x+(y\times\sqrt{}z)$. 
\begin{center}
	\begin{forest}
		[{$x\mainconnective{+}(y\times\sqrt{}z)$}, label=right:{\scriptsize $=17$}
		[$\atomicstage{x}$, label=left:{\scriptsize $7=$}]
		[{$y\mainconnective{\times}\sqrt{}z$}, label=right:{\scriptsize $=10$}
		[$\atomicstage{y}$, label=left:{\scriptsize $5=$}]
		[{$\mainconnective{\sqrt{}}z$}, label=right:{\scriptsize $=2$}
		[$\atomicstage{z}$, label=right:{\scriptsize $=4$}]
		]
		]
		]
	\end{forest}
\end{center}

Similarly for TFL, we use the valuation to set the values for our atomic sentences and then determine the values of our complex sentences by applying our truth rules. 

Let us see an example. Consider 
`$(\enot I\eand H)\eif H$'. 
The formation tree will help us know what to do (see \S\ref{TFLsentences}):
\begin{center}
	\begin{forest}
		[$(\enot I\eand H)\mainconnective{\eif} H$
		[$(\enot I\mainconnective{\eor} H)$
		[$\mainconnective{\enot} I$
		[$\atomicstage{I}$]
		]
		[$\atomicstage{H}$]
		]
		[$\atomicstage{H}$]
		]
	\end{forest}
\end{center}

%Now, we will start by a slight reformatting of the formation tree. 
%
%\begin{center}
%	\begin{tabular}{c c d e e ee e e f}
%		
%		&	& $($ & $\enot$ & $\atomicstage{I}$ & $\eor$ & $\atomicstage{H}$ & $)$ & $\mainconnective{\eif}$ & $\atomicstage{H}$ \\
%		%&&&&&&&&T&
%		\\[1ex]
%		& & $\mainconnectivecolor{(}$ & $\enot$ & $\atomicstage{I}$ & $\mainconnective{\eor}$ & $\atomicstage{H}$ & $\mainconnectivecolor{)}$ & &$\atomicstage{H}$ \\
%%			&&&&&T&&&&F
%			\\[1ex]
%		& & & $\mainconnective{\enot}$ & $\atomicstage{I}$ & & $\atomicstage{H}$ & & &\\
%%			&&&F&&&F&&&
%			\\[1ex]
%		& & & & $\atomicstage{I}$ & & & & &  \\
%%			&&&&T&&&&&
%	\end{tabular}
%\end{center}



The lowest point on the tree is ${I}$. For the truth of this, we just consult the valuation. On this valuation:
\begin{center}
	\begin{tabular}{rccccccc}
		&	$A$&${H}$&${I}$&$B$&$C_{35}$&\ldots\\\hline
		$v$:&	F & T &F&F&F&T&\ldots
	\end{tabular}
\end{center}
\begin{center}
	\begin{forest}
	[$(\enot I\eand H)\mainconnective{\eif} H$
	[$(\enot I\mainconnective{\eor} H)$
	[$\mainconnective{\enot} I$
	[$\atomicstage{I}$\\False on $v$]
	]
	[$\atomicstage{H}$]
	]
	[$\atomicstage{H}$]
	]
\end{forest}
\end{center}
%\begin{center}
%	\begin{tabular}{c c d e e ee e e f}
%		
%		&	& $($ & $\enot$ & $I$ & $\eor$ & $H$ & $)$ & $\mainconnective{\eif}$ & $H$ \\
%		&&&&&&&&&
%		\\[1ex]
%		& & $\mainconnectivecolor{(}$ & $\enot$ & $I$ & $\mainconnective{\eor}$ & $H$ & $\mainconnectivecolor{)}$ & &$H$ \\
%					&&&&&&&&&F
%		\\[1ex]
%		& & & $\mainconnective{\enot}$ & $I$ & & $H$ & & &\\
%					&&&&&&F&&&
%		\\[1ex]
%		& & & & $I$ & & & & &  \\
%					&&&&T&&&&&
%	\end{tabular}
%\end{center}

The truth rule for $\enot$ tells us how the truth value of $\enot I$ depends on the truth of ${I}$. 
Since ${I}$ is false on valuation $v$, by consulting the truth rule for $\neg$ we can se that $\enot I$ is true on the valuation $v$:
\begin{center}
	\begin{forest}
		[$(\enot I\eand H)\mainconnective{\eif} H$
		[$(\enot I\mainconnective{\eand} H)$
		[$\mainconnective{\enot} I$\\True on $v$
		[$\atomicstage{I}$\\False on $v$]
		]
		[$\atomicstage{H}$\\True on $v$]
		]
		[$\atomicstage{H}$]
		]
	\end{forest}
\end{center}

Then the rule for $\eand$ tells us how the truth value of $\enot I\eand H$ depends on the truths of $\enot I$ and ${H}$: as $\enot I$ is true (which we determined) and ${H}$ is true (consult the valuation), then $\enot I \eand H$ is also true on the valuation $v$.
\begin{center}
	\begin{forest}
		[$(\enot I\eand H)\mainconnective{\eif} H$\\
		[$(\enot I\mainconnective{\eand} H)$\\True on $v$
		[$\mainconnective{\enot} I$\\True on $v$
		[$\atomicstage{I}$\\False on $v$]
		]
		[$\atomicstage{H}$\\True on $v$]
		]
		[$\atomicstage{H}$\\True on $v$]
		]
	\end{forest}
\end{center}
Finally, the rule for $\eif$ tells us how the truth value of $(\enot I\eand H)\eif H$ depends on those of $(\enot I\eand H)$ and ${H}$.
\begin{center}
	\begin{forest}
		[$(\enot I\eand H)\mainconnective{\eif} H$\\True on $v$
		[$(\enot I\mainconnective{\eand} H)$\\True on $v$
		[$\mainconnective{\enot} I$\\True on $v$
		[$\atomicstage{I}$\\False on $v$]
		]
		[$\atomicstage{H}$\\True on $v$]
		]
		[$\atomicstage{H}$\\True on $v$]
		]
	\end{forest}
\end{center}


To keep track of the answer, we will present it on the truth table. 

\begin{center}
	\begin{tabular}{r c c| c c c c c c c c c}
		&&
		
		& \mainconnectivecolor{(} & {\enot} & $I$ & {\eand} & $H$ & \mainconnectivecolor{)} & \tikzmarknode{r1_imp}{$\mainconnective{\eif}$}& $H$  \\[.2em]
		
		% Row 2: children of Row 1
		&& &
		\mainconnectivecolor{(}
		& {\enot} & $I$ & \tikzmarknode{r2_and}{$\mainconnective{\eand}$} & $H$
		& \mainconnectivecolor{)} & &
		\tikzmarknode{r2_Hr}{$\atomicstage{H}$} & \\[.2em]
		
		% Row 3: children of Row 2 (left side)
		&& & & \tikzmarknode{r3_not}{$\mainconnective{\enot}$} & $I$ & & \tikzmarknode{r3_Hl}{$\atomicstage{H}$} & & & \\[.2em]
		
		% Row 4: child of Row 3 (for the negation)
		Valuation & $I$ & $H$ & & & \tikzmarknode{r4_I}{$\atomicstage{I}$} & & & && & \\[.2em]\hline
		
% % v1: I=T, H=T
%$v_1:$ & T & T & & F & T & F & T & & \TTbf{T} & T \\
%
%% v2: I=T, H=F
%$v_2:$ & T & F & & F & T & F & F & & \TTbf{T} & F \\

% v3: I=F, H=T
$v:$ & F & T & & T & F & T & T & & \TTbf{T} & T \\

% v4: I=F, H=F
%$v_4:$ & F & F & & T & F & F & F & & \TTbf{T} & F \\

	\end{tabular}
	
	\begin{tikzpicture}[remember picture,overlay]
		\tikzset{arrowformationtree/.style={-{Stealth[length=2.2mm,width=1.6mm]},gray,opacity=0.6}}
		
		% Row 2 -> Row 1
		\draw[arrowformationtree] (r2_and) to[bend left=18] (r1_imp);
		\draw[arrowformationtree] (r2_Hr)  to[bend right=18] (r1_imp);
		
		% Row 3 -> Row 2
		\draw[arrowformationtree] (r3_not) to[bend left=18] (r2_and);
		\draw[arrowformationtree] (r3_Hl)  to[bend right=18] (r2_and);
		
		% Row 4 -> Row 3
		\draw[arrowformationtree] (r4_I) to[bend left=18] (r3_not);
	\end{tikzpicture}
\end{center}



We do this because when we consider multiple valuations, we'll need to track where their answers are. 
For example:


\begin{center}
	\begin{tabular}{r c c c| d e e e e e e e e e e e  e e f}
		&&&
		& \tikzmarknode{r1_outerneg}{$\mainconnective{\enot}$} & $( $ & $C$ & $\eor$ & $( $ & $D$ & $\eif$ & $( $ & $A$ & $\eor$ & $\enot$ & $D$ & $)$ & $)$ & $)$ \\[1em]
		
		% Row 2: parents of Row 3; child of Row 1
		& & & &
		& \mainconnectivecolor{(}
		& $C$ & \tikzmarknode{r2_bigor}{$\mainconnective{\eor}$}
		& $( $ & $D$ & ${\eif}$
		& $( $ & $A$ & ${\eor}$
		& ${\enot}$ & $D$
		& $)$ & ${)}$& $\mainconnectivecolor{)}$ \\[1em]
		
		% Row 3: children of Row 2; parents of Row 4
		& & & &
		& & \tikzmarknode{r3_C}{$\atomicstage{C}$}
		& &{\mainconnectivecolor{$($ }}
		& $D$ & \tikzmarknode{r3_implies}{$\mainconnective{\eif}$}
		& $( $ & $A$ & $\eor$ & ${\enot}$ & $D$
		& $)$ & \tikzmarknode{r3_closeCond}{\mainconnectivecolor{)}} & \\[1em]
		
		% Row 4: children of Row 3; parents of Row 5
		& & & & & & & & &
		\tikzmarknode{r4_D}{$\atomicstage{D}$}
		& & & $A$ & \tikzmarknode{r4_inneror}{$\mainconnective{\eor}$}
		& $\enot$ & $D$ & & & \\[1em]
		
		% Row 5: children of Row 4; parent of Row 6
		& & & & & & & & & & & &
		\tikzmarknode{r5_A}{$\atomicstage{A}$}
		& & \tikzmarknode{r5_not}{$\mainconnective{\enot}$}
		& ${D}$ & & & \\[1em]
		
		% Row 6: child of Row 5
		Valuation&$A$ &$C$ &$D$ & & & & & & & & & & & &
		\tikzmarknode{r6_D}{$\atomicstage{D}$}
		& & & \\[.2em]\hline
		
%		$v_1:$ & T & T & T & \TTbf{F} & & T & T & & T & T & & T & T & F & T & & & \\
%		$v_2:$ & T & T & F & \TTbf{F} & & T & T & & F & T & & T & T & T & F & & & \\
%		$v_3:$ & T & F & T & \TTbf{F} & & F & T & & T & T & & T & T & F & T & & & \\
		$v_1:$ & T & F & F & \TTbf{F} & & F & T & & F & T & & T & T & T & F & & & \\
		$v_2:$ & F & T & T & \TTbf{F} & & T & T & & T & F & & F & F & F & T & & & \\
%		$v_6:$ & F & T & F & \TTbf{F} & & T & T & & F & T & & F & F & T & F & & & \\
%		$v_7:$ & F & F & T & \TTbf{T} & & F & F & & T & F & & F & F & F & T & & & \\
%		$v_8:$ & F & F & F & \TTbf{F} & & F & T & & F & T & & F & F & T & F & & & \\
	\end{tabular}
	\begin{tikzpicture}[remember picture,overlay]
		\tikzset{arrowformationtree/.style={-{Stealth[length=2.2mm,width=1.6mm]},gray,opacity=0.6}}
		
		
		% Row 2 -> Row 1
		\draw[arrowformationtree] (r2_bigor) to[bend left=18] (r1_outerneg);
		
		% Row 3 -> Row 2
		\draw[arrowformationtree] (r3_C)        to[bend left=18] (r2_bigor);     
		\draw[arrowformationtree] (r3_implies) to[bend right=18] (r2_bigor);  
		
		% Row 4 -> Row 3
		\draw[arrowformationtree] (r4_D)       to[bend left=18] (r3_implies);    
		\draw[arrowformationtree] (r4_inneror) to[bend right=18] (r3_implies);   
		
		% Row 5 -> Row 4
		\draw[arrowformationtree] (r5_A)   to[bend left=18] (r4_inneror);         
		\draw[arrowformationtree] (r5_not) to[bend right=18] (r4_inneror);        
		
		% Row 6 -> Row 5
		\draw[arrowformationtree] (r6_D) to[bend left=18] (r5_not);               
	\end{tikzpicture}
\end{center}

\begin{practiceproblems}


\problempart Determine the truth values for these sentences on this valuation:

\begin{center}
	\begin{tabular}{rccccccccc}
		& $A$ & $B$ & $C$ & $D$ & $E$ & $H$ & $I$ & $J$ & $C_{35}$\\\hline
		$v$: & F & F & F & T & F & T & F & T & F
	\end{tabular}
\end{center}

\begin{earg}
	\item $\,\enot A \eor (B \eif \enot (C \eor A))\,$
	\item $\, (A \eand H) \eor (I \eand \enot B)\,$
	\item $\, H \eif (A \eor C)\,$
	\item $\, \enot (D \eand E)\,$
	\item $\, (B \eif J) \eand (J \eif H)\,$
	\item $\, (C \eor C_{35}) \eif H\,$
	\item $\, \enot\enot A \eor \enot I\,$
	\item $\, (A \eif B) \eif ((H \eand D) \eor \enot E)\,$
	\item $\, ((J \eand \enot C )\eif D) \eand \enot (J \eand E)\,$
	\item $\, (A \eor B) \eand (\enot A \eor \enot B)$
	\item $\, (H \eif D) \eor (I \eif A)\,$
\end{earg}

	
	\end{practiceproblems}

\chapter{Truth Tables}
\section{Validity and truth tables, the idea}\label{sec:valtt}
In \S\ref{s:Valid} we said that an argument was valid, if and only if there is no interpretation such that all premises are true but the conclusion false. At that point the definition was suggestive, but we lacked a clear understanding of `interpretation' and when a sentence is true relative to an interpretation. But for the language of TFL we can now turn the informal definition in a precise and rigorous definition as, for TFL, interpretations are given by valuations, assignments of truth values to the atomic sentences. 

Recall that an argument consisted of a number of premises together with a conclusion. A TFL-argument then can be written as $X_1,\ldots,X_n\therefore Y$ where $X_1,\ldots,X_n$ and $Y$ are sentences of TFL, and $X_1,\ldots,X_n$ are the premises and $Y$ the conclusion of the argument.

\factoidbox{\label{TFLvalid}A TFL-argument is valid if and only if there is no valuation on which all the premises are true but the conclusion is false. 
	
	That is, `$X_1,\ldots,X_n\therefore Y$' is \define{valid} if and only if there is no valuation $v$ such that $X_1$ is true relative to $v$ and $X_2$ is true relative to $v$, and so on, up to $X_n$ being true relative to $v$, but $Y$ is false relative to $v$.}

So to determine whether a TFL-argument is valid or not, we should show whether or not there is a valuation on which all premises are true but the conclusion false. 

We have discussed how to determine the truth of a sentence relative to a single valuation. To determine whether an argument is valid or not we need to consider all valuations and ask whether there is at least one valuation where the premises are all true but the conclusion is false. 

There are many different valuations. But to determine validity, we only need to consider the valuations' assignment to the atomic sentences that occur in the premises and the conclusion of the argument. Consider the argument:
$$\enot C,\,D\eif C\therefore\enot  D$$
To determine its validity, we only care about the truth values of the three sentence $\enot C$, $D\eif C$ and $\enot D$. So all that matters for our purposes is the valuation's assignment of truth values to the atomic sentences $C$ and $D$.

There will be four possible such valuations. 
\begin{center}
	\begin{tabular}{rcc}
		Valuation&$C$&$D$\\\hline
		$v_1:$&T & T \\
		$v_2:$&T & F \\
		$v_3:$&F & T \\
		$v_4:$&F & F \\
	\end{tabular}
\end{center}

We need to determine the truth values of the three sentences, $\enot C$, $D\eif C$ and $\enot D$ on each valuation to determine whether the argument is valid. 

This will result in:
 \begin{center}
	\begin{tabular}{rcc||c|c||c}
		Valuation&$C$&$D$&$\enot D$&$C\eif D$&$\enot C$\\\hline
		$v_1:$&T & T & F & T & F\\
		$v_2:$&T & F & T & F & F\\
		$v_3:$&F & T & F & T & T\\
		$v_4:$&F & F & {T}& {T} & {T}
	\end{tabular}
\end{center}

$v_4$ is the only valuation on which both the premises (`$\enot D$' and `$C\eif D$') are true; and on this valuation, the conclusion (`$\enot C$') is also true. 
There is thus no valuation where all the premises are true and the conclusion is false. So the argument is valid. 

To allow you to determine this we will show how to construct so-called (complete) truth tables. 
 
\section{Determining all the valuations}

Suppose we are interested in determining the validity of an argument. We need to consider all the valuations for the atomic sentences that appear in the argument. 


If we are only interested in the atomic sentences $A$ there are two ways of assigning truth values to the atomic sentence $A$:
\begin{center}
	\begin{tabular}{ccccccc}
		$A$\\\hline
		T\\
		F 
	\end{tabular}
\end{center}
That is, there are two valuations for the atomic sentence $A$. 

If we are interested in two atomic sentences, say $A$ and $B$, we need to consider all possible assignments of truth values to $A$ and to $B$. 

Whatever the valuation assigns to $A$, it can assign either true or false to $B$. So we ``split'' each valuation we have so far into two versions, one which assigns $B$ as true and another assigning $B$ as false. 

\begin{center}
	\begin{tabular}{cccccc}
		 $A$&$B$\\\hline
	T&T\\
	T&F\\
	F&T\\
	F&F
	\end{tabular}
\end{center}

If we are also interested in a further atomic sentence, say $C$, we further split each of our valuations into two, one which also assigns $C$ as true, and one which assigns $C$ as false.

\begin{center}
	\begin{tabular}{cccc}
		$A$&$B$&$C$&\\\hline
		T& T &T&\\
		T & T & F&\\
		T & F & T& \\
		T & F &F&\\
		F & T &T&\\
		F& T & F & \\
		F & F & T &\\
		F & F & F& 
	\end{tabular}
\end{center}


What if we add a further atomic sentence, say $P_{235}$? We want to consider the valuations for atomic sentences $A,B,C,P_{235}$.
Again we need to split each of our valuations for $A,B,C$ into one which assigns $P_{235}$ as true and another which assigns it as false. 

\begin{center} \begin{tabular}{cccc} $A$ & $B$ & $C$ & $P_{235}$\\\hline T & T & T & T\\ T & T & T & F\\ T & T & F & T\\ T & T & F & F\\ T & F & T & T\\ T & F & T & F\\ T & F & F & T\\ T & F & F & F\\ F & T & T & T\\ F & T & T & F\\ F & T & F & T\\ F & T & F & F\\ F & F & T & T\\ F & F & T & F\\ F & F & F & T\\ F & F & F & F \end{tabular} \end{center}


The list of all possible valuations gets very big very fast.

There are $2^n$ ways to assign truth values to $n$ atomic sentences. Using truth tables to determine validity thus becomes quite difficult and it is one reason we consider an alternative method: that of proofs, which we will see in \ref{ch.NDTFL}.


Here's one strategy that some people find useful for listing all the valuations. 
\factoidbox{
\begin{itemize}
	\item 	For $n$ atomic sentences, there are $2^n$ valuations. 
	
	\item Draw a table with the $n$ atomic sentences in the header row and count $2^n$ rows (make a mark). 
	
	\item We'll now fill out the `T' and `F''s in these $n$-columns and $2^n$ rows:
\begin{itemize}
	 \item	Start with the last atomic sentence, the right-most column and \emph{alternate} between `T’ and `F’ until you've filed out all $2^n$ rows. 
	
\item	Next consider the penultimate atomic sentence, the column to the left of the one just completed,  write \emph{two} ‘T’s, write two ‘F’s,
	and repeat. 
	
\item	For the third-last atomic sentence letter, write \emph{four} ‘T’s followed
	by four ‘F’s. This would yields an eight line truth table. 
	
\item	For the fourth-last atomic sentence, you have \emph{eight} ‘T’s followed by eight ‘F’s. 
	
\item	And so on until you get to the first atomic sentence.
\end{itemize}
\end{itemize}
	}
	
\section{Complete truth tables for a sentence}



\begin{highlighted}
	A \define{complete truth table} for a sentence has a line for each valuation for the atomic sentences appearing in the sentence, and displays the truth value of the sentence on each of the valuations. 
\end{highlighted}


Consider the sentence `$(\enot I\eand H)\eif H$'. We will give a \emph{truth table} which lists all the valuations and says whether this sentence is true or false on each of them.

The valuations assign either the value ``True'' or ``False'' to each atomic sentence. In this case we have two atomic sentences, $I$ and $H$, so we have four ($2^2$) valuations ($v_1,\dots,v_4$) each of which is a line in the truth table:
\begin{center}
	\begin{tabular}{ccc|c}
		Valuation&$I$&$H$&$(\enot I\eand H)\eif H$\\\hline
		$v_1$&T&T&\\
		$v_2$&T&F&\\
		$v_3$&F&T&\\
		$v_4$&F&F&
	\end{tabular}
\end{center}
Our job is to fill out the truth values of `$(\enot I\eand H)\eif H$' on each of the four valuations.

We have already shown how to do this on a particular valuation: one first writes down the formation tree of the sentence. (see \S\ref{TFLsentences}):
\begin{center}
	\begin{forest}
		[$(\enot I\eand H)\mainconnective{\eif} H$
		[$(\enot I\mainconnective{\eand} H)$
		[$\mainconnective{\enot} I$
		[$\atomicstage{I}$]
		]
		[$\atomicstage{H}$]
		]
		[$\atomicstage{H}$]
		]
	\end{forest}
\end{center}
The idea is that we work ourselves from \define{leaves} of the tree (the atomic sentence) to the \define{root} of the formation tree (the sentence that has been constructed). The truth rule for $\enot$ tells us how the truth value of $\enot I$ depends on the truth of $I$. Then the rule for $\eand$ tells us how the truth value of $\enot I\eand H$ depends on the truths of $\enot I$ and $H$; and finally, the rule for $\eif$ tells us how the truth value of $(\enot I\eand H)\eif H$ depends on those of $\enot I\eand H$ and $H$.


We will record the truth value of each subsentence underneath its main connective. 
So we will somewhat recreate the formation tree in the header of the truth table. 
\begin{center}
	\begin{tabular}{r c c| d e e ee e e f}
\multirow{4}{*}{Valuation}	&	\multirow{4}{*}{$H$}&\multirow{4}{*}{$I$}	& $($ & $\enot$ & $I$ & $\eand$ & $H$ & $)$ & $\mainconnective{\eif}$ & $H$ \\[.2em]
	&	& & $\mainconnectivecolor{(}$ & $\enot$ & $I$ & $\mainconnective{\eand}$ & $H$ & $\mainconnectivecolor{)}$ & &$\atomicstage{H}$ \\[.2em]
	&	& & & $\mainconnective{\enot}$ & $I$  & & $\atomicstage{H}$ & & &\\[.2em]
	&	& & & & $\atomicstage{I}$& & & & &  \\\hline 
	$v_1:$&T&T&\\
	$v_2:$&T&F&\\
	$v_3:$&F&T\\
	$v_4:$&F&F
	\end{tabular}
\end{center}

We will work our way up the formation tree, as displayed in the header of the table. 
In the bottom line we just have `$I$', whose truth value is just specified by the valuation. So we can fill that out

\begin{center}
	\begin{tabular}{r c c| d e e ee e e fc}
		\multirow{4}{*}{Valuation}	&	\multirow{4}{*}{$H$}&\multirow{4}{*}{$I$}	& $($ & $\enot$ & $I$ & $\eand$ & $H$ & $)$ & $\mainconnective{\eif}$ & $H$ \\[.2em]
		&	& & $\mainconnectivecolor{(}$ & $\enot$ & $I$ & $\mainconnective{\eand}$ & $H$ & $\mainconnectivecolor{)}$ & &$\atomicstage{H}$ \\[.2em]
		&	& & & $\mainconnective{\enot}$ & $I$  & & $\atomicstage{H}$ & & &\\[.2em]
		&	& & & & $\atomicstage{I}$& & & & & &$\star$ \\\hline 
		$v_1:$&T&T& & &T& & & & &\\
		$v_2:$&T&F& & &F& & & & &\\
		$v_3:$&F&T& & &T& & & & &\\
		$v_4:$&F&F& & &F& & & & &\\
		&&&&&$\star$&&&&&&
	\end{tabular}
\end{center}

The next line is to include the truth value for `$\enot I$' and the truth value for `$H$'. The truth value for `$H$' is just specified by the valuation so that can be copied across. 

\begin{center}
	\begin{tabular}{r c c| d e e ee e e f c}
		\multirow{4}{*}{Valuation}	&	\multirow{4}{*}{$H$}&\multirow{4}{*}{$I$}	& $($ & $\enot$ & $I$ & $\eand$ & $H$ & $)$ & $\mainconnective{\eif}$ & $H$ &  \\[.2em]
		&	& & $\mainconnectivecolor{(}$ & $\enot$ & $I$ & $\mainconnective{\eand}$ & $H$ & $\mainconnectivecolor{)}$ & &$\atomicstage{H}$ & \\[.2em]
		&	& & & $\mainconnective{\enot}$ & $I$  & & $\atomicstage{H}$ & & & & $\star$ \\[.2em]
		&	& & & & $\atomicstage{I}$& & & & &  & \\\hline 
		$v_1:$&T&T& & &T& &T& & & &\\
		$v_2:$&T&F& & &F& &T& & & &\\
		$v_3:$&F&T& & &T& &F& & & &\\
		$v_4:$&F&F& & &F& &F& & & &\\
		&&&&&&&$\star$&&&&
	\end{tabular}
\end{center}


The truth value for `$\enot I$' is determined using the truth rule for negation:
\begin{center}
	\begin{tabular}{ccc}
		If \metaX is & & then $\enot X$ is\\
		T&$\leadsto$&F\\
		F&$\leadsto$&T
	\end{tabular}
\end{center}
So we take the truth value of `$I$' as recorded underneath that `$I$', and apply this rule to determine the truth value for `$\enot I$'. We record the answer underneath the main connective of `$\enot I$', the `$\enot$'. 
This results in:

\begin{center}
	\begin{tabular}{r c c| d e e ee e e fc}
		\multirow{4}{*}{Valuation}	&	\multirow{4}{*}{$H$}&\multirow{4}{*}{$I$}	& $($ & $\enot$ & $I$ & $\eand$ & $H$ & $)$ & $\mainconnective{\eif}$ & $H$ \\[.2em]
		&	& & $\mainconnectivecolor{(}$ & $\enot$ & $I$ & $\mainconnective{\eand}$ & $H$ & $\mainconnectivecolor{)}$ & &$\atomicstage{H}$ \\[.2em]
		&	& & & $\mainconnective{\enot}$ & $I$  & & $\atomicstage{H}$ & & &&$\star$\\[.2em]
		&	& & & & $\atomicstage{I}$& & & & &  \\\hline 
		$v_1:$&T&T& &F&T& &T& & &\\
		$v_2:$&T&F& &T&F& &T& & &\\
		$v_3:$&F&T& &F&T& &F& & &\\
		$v_4:$&F&F& &T&F& &F& & &\\
		&&&&\tikzmarknode{negI}{$\star$}&\tikzmark{I}
	\end{tabular}
\end{center}

\begin{tikzpicture}[remember picture,overlay]
	\draw[-Stealth] (pic cs:I) to[bend left=35] (negI.south);
\end{tikzpicture}



The next line we need to add the truth value for `$H$' again, which is specified by the valuation, and the truth value for `$\enot I\eand H$'.
For `$\enot I\eand H$' we will use the truth rule for $\eand$:
\begin{center}
	\begin{tabular}{ccccc}
		If \metaX is&and \metaY is  && then $\metaX\eand \metaY$ is\\
		T&T&$\leadsto$&T\\
		T&F&$\leadsto$&F\\
		F&T&$\leadsto$&F\\
		F&F&$\leadsto$&F
	\end{tabular}
\end{center}

So we take the truth value for `$\enot I$', as recorded underneath `$\enot$', and the truth value for `$H$', and apply this truth rule, writing the answer under its main connective, `$\eand$'. So we can fill out:



\begin{center}
	\begin{tabular}{r c c| d e e ee e e f c}
		\multirow{4}{*}{Valuation}	&	\multirow{4}{*}{$H$}&\multirow{4}{*}{$I$}	& $($ & $\enot$ & $I$ & $\eand$ & $H$ & $)$ & $\mainconnective{\eif}$ & $H$ \\[.2em]
		&	& & $\mainconnectivecolor{(}$ & $\enot$ & $I$ & $\mainconnective{\eand}$ & $H$ & $\mainconnectivecolor{)}$ & &$\atomicstage{H}$&$\star$ \\[.2em]
		&	& & & $\mainconnective{\enot}$ & $I$  & & $\atomicstage{H}$ & & &\\[.2em]
		&	& & & & $\atomicstage{I}$& & & & &  \\\hline 
		$v_1:$&T&T& &F&T&F&T& &&T\\
		$v_2:$&T&F& &T&F&T&T& &&T\\
		$v_3:$&F&T& &F&T&F&F& &&F\\
		$v_4:$&F&F& &T&F&F&F& &&F\\
&&&&\tikzmarknode{negI}{}&&\tikzmarknode{andStar}{$\star$}&\tikzmarknode{Hcol}{}&&

	\end{tabular}
\end{center}

	\begin{tikzpicture}[remember picture,overlay]
	\draw[-Stealth] (negI) to[bend right=35] (andStar.south);
	\draw[-Stealth] (Hcol.south) to[bend left=35] (andStar.south);
\end{tikzpicture}



Our final stage is to determine the truth value for `$(\enot I\eand H)\eif H$'

So we take our truth value we have determined for `$(\enot I\eand H)$', which is recorded underneath the `$\eand$', and the truth value for `$H$', and apply the truth rule for `$\eif$'
\begin{center}
	\begin{tabular}{cccc}
		If \metaX is&and \metaY is  && then $\metaX\eif \metaY$ is\\
		T&T&$\leadsto$&T\\
		T&F&$\leadsto$&F\\
		F&T&$\leadsto$&T\\
		F&F&$\leadsto$&T
	\end{tabular}
\end{center}
recording the answer underneath `$\eif$'. As this is the final stage, we have determined the truth value for the target sentence, `$(\enot I\eand H)\eif H$', so we will boldface this answer to indicate that it is our final answer, the truth value for `$(\enot I\eand H)\eif H$'. 


\begin{center}
	\begin{tabular}{r c c| d e e ee e e fc}
		\multirow{4}{*}{Valuation}	&	\multirow{4}{*}{$H$}&\multirow{4}{*}{$I$}	& $($ & $\enot$ & $I$ & $\eand$ & $H$ & $)$ & $\mainconnective{\eif}$ & $H$&$\star$ \\[.2em]
		&	& & $\mainconnectivecolor{(}$ & $\enot$ & $I$ & $\mainconnective{\eand}$ & $H$ & $\mainconnectivecolor{)}$ & &$\atomicstage{H}$ \\[.2em]
		&	& & & $\mainconnective{\enot}$ & $I$  & & $\atomicstage{H}$ & & &\\[.2em]
		&	& & & & $\atomicstage{I}$& & & & &  \\\hline 
		$v_1:$&T&T& &F&T&F&T& &\TTbf{T}&T\\
		$v_2:$&T&F& &T&F&T&T& &\TTbf{T}&T\\
		$v_3:$&F&T& &F&T&F&F& &\TTbf{T}&F\\
		$v_4:$&F&F& &T&F&F&F& &\TTbf{T}&F\\
		&&&&&&\tikzmark{wedge}&&&\tikzmarknode{arrow}{$\star$}&\tikzmark{H2}&		
		\end{tabular}
	\end{center}

\begin{tikzpicture}[remember picture,overlay]
\draw[-Stealth] (pic cs:wedge) to[bend right=35] (arrow.south);
\draw[-Stealth] (pic cs:H2) to[bend left=35] (arrow.south);
\end{tikzpicture}


%%%%% The below has separate columns for the different subsetnences. 
%
%So to work out the truth values of $(\enot I\eand H)\eif H$ we first need to work out the truth values of $\enot I$ and $\enot I\eand H$. We expand our truth table with columns for each of these.
%\begin{center}
%	\begin{tabular}{ccc|c|c||c}
%		Valuation&$I$&$H$&$\enot I$&$(\enot I\eand H)$&$(\enot I\eand H)\eif H$\\\hline
%		$v_1$&T&T&&\\
%		$v_2$&T&F&&\\
%		$v_3$&F&T&&\\
%		$v_4$&F&F&&
%	\end{tabular}
%\end{center}
%The first step is $\enot I$. We use the truth table (rule) for negation:
%\begin{center}
%	\begin{tabular}{ccc}
%		If \metaX is & & then $\enot X$ is\\
%		T&$\leadsto$&F\\
%		F&$\leadsto$&T
%	\end{tabular}
%\end{center}
%Now, we can fill out:
%
%\begin{center}
%	\begin{tabular}{ccc|c|c||c}
%		Valuation&$I$&$H$&$\enot I$&$(\enot I\eand H)$&$(\enot I\eand H)\eif H$\\\hline
%		$v_1$&T&T&F&\\
%		$v_2$&T&F&F&\\
%		$v_3$&F&T&T&\\
%		$v_4$&F&F&T&
%	\end{tabular}
%\end{center}
%We worked these out using the following instructions:
%\begin{itemize}
%	\item Go to the column for $I$ and for every valuation do the following:
%	\begin{itemize}
%		\item If the value of $I$ is T, then put F into the column of $\enot I$ at the line of the valuation.
%		\item If the value of $I$ is F, then put T into the column of $\enot I$ at the line of the valuation.
%	\end{itemize}
%\end{itemize}
%The next step is to consider $\enot I\eand H$. For this we will use the truth rule for $\eand$:
%\begin{center}
%	\begin{tabular}{ccccc}
%		If \metaX is&and \metaY is  && then $\metaX\eand \metaY$ is\\
%		T&T&$\leadsto$&T\\
%		T&F&$\leadsto$&F\\
%		F&T&$\leadsto$&F\\
%		F&F&$\leadsto$&F
%	\end{tabular}
%\end{center}
%Now, we can fill out:
%\begin{center}
%	\begin{tabular}{ccc|c|c||c}
%		Valuation&$I$&$H$&$\enot I$&$(\enot I\eand H)$&$(\enot I\eand H)\eif H$\\\hline
%		$v_1$&T&T&F&F\\
%		$v_2$&T&F&F&F\\
%		$v_3$&F&T&T&T\\
%		$v_4$&F&F&T&F
%	\end{tabular}
%\end{center}
%We worked these using the following instructions:
%\begin{itemize}
%	\item  For every valuation go to column $H$ and do the following:
%	\begin{itemize}
%		\item[$\dagger$]  If the value of $H$ is F, put F into column $(\enot I\eand H)$ at the line of the valuation.
%		\item If the value of $H$ is T, go to column of $\enot I$:
%		\begin{itemize}
%			\item if the value of $\enot I$ is T, put T into column $(\enot I\eand H)$ and at the line of the valuation.
%			\item if the value of $\enot I$ is F, put F into column $(\enot I\eand H)$ and at the line of the valuation.
%		\end{itemize}
%	\end{itemize}
%\end{itemize}
%The instruction $\dagger$ is justified by the truth rules of the conjunction: if one of the conjuncts has value F, the conjunction will also have value F.
%
%Now, finally, we need to look at $(\enot I\eand H)\eif H$, and will use the truth rule for $\eif$:
%
%\begin{center}
%	\begin{tabular}{cccc}
%		If \metaX is&and \metaY is  && then $\metaX\eif \metaY$ is\\
%		T&T&$\leadsto$&T\\
%		T&F&$\leadsto$&F\\
%		F&T&$\leadsto$&T\\
%		F&F&$\leadsto$&T
%	\end{tabular}
%\end{center}
%Now, we can fill out:
%\begin{center}
%	\begin{tabular}{ccc|c|c||c}
%		Valuation&$I$&$H$&$\enot I$&$(\enot I\eand H)$&$(\enot I\eand H)\eif H$\\\hline
%		$v_1$&T&T&F&F&T\\
%		$v_2$&T&F&F&F&T\\
%		$v_3$&F&T&T&T&T\\
%		$v_4$&F&F&T&F&T
%	\end{tabular}
%\end{center}
%We worked these out by the following procedure:
%\begin{itemize}
%	\item  For every valuation go to column $(\enot I\eand H)$ and do the following:
%	\begin{itemize}
%		\item[$\star$]  If the value of $(\enot I\eand H)$ is F, put T into column $(\enot I\eand H)\eif H$ at the line of the valuation.
%		\item If the value of $(\enot I\eand H)$ is T go to column $H$ and:
%		\begin{itemize}
%			\item if the value of $H$ is T, put T into column $(\enot I\eand H)\eif H$ at the line of the valuation.
%			\item if the value of $H$ is F, put F into column $(\enot I\eand H)\eif H$ at the line of the valuation.
%		\end{itemize}
%	\end{itemize}
%\end{itemize}
%The instruction $\star$ is justified by the truth rules of the conditional: if the antecedent of the conditional has value F, then the conditional will also value T.
%%\subsection{How to do truth tables}
%
%With this example in mind let us try to give a general instruction for constructing a truth table for a sentence $X$.
%
%\factoidbox{
%	\begin{enumerate}
%		\item Write down the formation tree of $X$.
%		\item Find all atomic sentences on the tree. These will be the leaves of the tree.
%		\item If you have found all atomic sentences, you can start the truth table:
%		\begin{itemize}
%			\item You will need a column for every atomic sentence.
%			\item If there are $n$ atomic sentences, you will need $2^n$ valuations, that is, $2^n$ horizontal lines.
%			\item Make sure you have correctly written down all the different valuations!
%		\end{itemize}
%		\item From the atomic sentences (the leaves of the tree) move upwards to the root (the sentence $X$) and
%		\begin{itemize}
%			\item for each node (constituent sentence) create a column in the truth table;
%			\item make sure you correctly identify the main connective of the sentence heading the column;
%			\item the root, that is, the sentence $X$ should be the last column of the truth table
%		\end{itemize}
%		\item Moving from left to right compute the truth values of each column
%		\begin{itemize}
%			\item Make sure you use the truth rule associated with the main connective of the sentence heading the column.
%			\item Stay in one and the same valuation (horizontal line) when you compute the truth value of the column.
%			\item Compute the truth value for every valuation.
%		\end{itemize}
%		\item You are done when you have computed the truth values of the column of $X$ \emph{and} there are no gaps in the truth table.
%\end{enumerate}}




%\begin{center}
%	\begin{tabular}{r c c c| d e e e e e e e e e e e  e e f}
	%		\multirow{6}{*}{Valuation} & \multirow{6}{*}{$A$} & \multirow{6}{*}{$C$} & \multirow{6}{*}{$D$}
	%		& $\mainconnective{\enot}$ & $( $ & $C$ & $\eor$ & $( $ & $D$ & $\eif$ & $( $ & $A$ & $\eor$ & $\enot$ & $D$ & $)$ & $)$ & $)$ \\[.2em]
	%	& & & & 
	%& $\mainconnectivecolor{(}$ & $C$ & $\mainconnective{\eor}$ & $( $ & $D$ & ${\eif}$ & $( $ & $A$ & $\eor$ & ${\enot}$ & $D$ & $)$ & ${)}$ & $\mainconnectivecolor{)}$ \\[.2em]
	%		
	%		& & & & 
	%		& & $\atomicstage{C}$ & & $\mainconnectivecolor{( }$ & $D$ & $\mainconnective{\eif}$ & $( $ & $A$ & $\eor$ & ${\enot}$ & $D$ & $)$ & $\mainconnectivecolor{)}$ & \\[.2em]
	%		
	%		& & & & & & & & & $\atomicstage{D}$ & & & $A$ & $\mainconnective{\eor}$ & $\enot$& $D$ & & & \\[.2em]
	%		
	%				& & & & & & & & & & & & $\atomicstage{A}$& &$\mainconnective{\enot}$ & ${D}$ & & & \\[.2em]
	%				
	%		& & & & & & & & & & & & & & & $\atomicstage{D}$ & & & \\[.2em]\hline
	%		
	%		
	%		$v_1:$ & T & T & T & \TTbf{F} & & T & T & & T & T & & T & T & F & T & & & \\
	%		$v_2:$ & T & T & F & \TTbf{F} & & T & T & & F & T & & T & T & T & F & & & \\
	%		$v_3:$ & T & F & T & \TTbf{F} & & F & T & & T & T & & T & T & F & T & & & \\
	%		$v_4:$ & T & F & F & \TTbf{F} & & F & T & & F & T & & T & T & T & F & & & \\
	%		$v_5:$ & F & T & T & \TTbf{F} & & T & T & & T & F & & F & F & F & T & & & \\
	%		$v_6:$ & F & T & F & \TTbf{F} & & T & T & & F & T & & F & F & T & F & & & \\
	%		$v_7:$ & F & F & T & \TTbf{T} & & F & F & & T & F & & F & F & F & T & & & \\
	%		$v_8:$ & F & F & F & \TTbf{F} & & F & T & & F & T & & F & F & T & F & & & \\
	%		
	%	\end{tabular}
%\end{center}


Here's another example. 

\begin{center}
	\begin{tabular}{r c c c| d e e e e e e e e e e e  e e f}
		\multirow{6}{*}{Valuation} & \multirow{6}{*}{$A$} & \multirow{6}{*}{$C$} & \multirow{6}{*}{$D$}
		& \tikzmarknode{r1_outerneg}{$\mainconnective{\enot}$} & $( $ & $C$ & $\eor$ & $( $ & $D$ & $\eif$ & $( $ & $A$ & $\eor$ & $\enot$ & $D$ & $)$ & $)$ & $)$ \\[1em]
		
		% Row 2: parents of Row 3; child of Row 1
		& & & &
		& \tikzmarknode{r2_openBigOr}{\mainconnectivecolor{(}}
		& $C$ & \tikzmarknode{r2_bigor}{$\mainconnective{\eor}$}
		& $( $ & $D$ & \tikzmarknode{r2_implies}{${\eif}$}
		& $( $ & $A$ & \tikzmarknode{r2_inneror}{${\eor}$}
		& \tikzmarknode{r2_not}{${\enot}$} & $D$
		& $)$ & \tikzmarknode{r2_closeBigOr}{${)}$} & $\mainconnectivecolor{)}$ \\[1em]
		
		% Row 3: children of Row 2; parents of Row 4
		& & & &
		& & \tikzmarknode{r3_C}{$\atomicstage{C}$}
		& &{\mainconnectivecolor{$($ }}
		& $D$ & \tikzmarknode{r3_implies}{$\mainconnective{\eif}$}
		& $( $ & $A$ & $\eor$ & ${\enot}$ & $D$
		& $)$ & \tikzmarknode{r3_closeCond}{\mainconnectivecolor{)}} & \\[1em]
		
		% Row 4: children of Row 3; parents of Row 5
		& & & & & & & & &
		\tikzmarknode{r4_D}{$\atomicstage{D}$}
		& & & $A$ & \tikzmarknode{r4_inneror}{$\mainconnective{\eor}$}
		& $\enot$ & $D$ & & & \\[1em]
		
		% Row 5: children of Row 4; parent of Row 6
		& & & & & & & & & & & &
		\tikzmarknode{r5_A}{$\atomicstage{A}$}
		& & \tikzmarknode{r5_not}{$\mainconnective{\enot}$}
		& ${D}$ & & & \\[1em]
		
		% Row 6: child of Row 5
		& & & & & & & & & & & & & & &
		\tikzmarknode{r6_D}{$\atomicstage{D}$}
		& & & \\[.2em]\hline
		
		$v_1:$ & T & T & T & \TTbf{F} & & T & T & & T & T & & T & T & F & T & & & \\
		$v_2:$ & T & T & F & \TTbf{F} & & T & T & & F & T & & T & T & T & F & & & \\
		$v_3:$ & T & F & T & \TTbf{F} & & F & T & & T & T & & T & T & F & T & & & \\
		$v_4:$ & T & F & F & \TTbf{F} & & F & T & & F & T & & T & T & T & F & & & \\
		$v_5:$ & F & T & T & \TTbf{F} & & T & T & & T & F & & F & F & F & T & & & \\
		$v_6:$ & F & T & F & \TTbf{F} & & T & T & & F & T & & F & F & T & F & & & \\
		$v_7:$ & F & F & T & \TTbf{T} & & F & F & & T & F & & F & F & F & T & & & \\
		$v_8:$ & F & F & F & \TTbf{F} & & F & T & & F & T & & F & F & T & F & & & \\
	\end{tabular}
	\begin{tikzpicture}[remember picture,overlay]
		\tikzset{arrowformationtree/.style={-{Stealth[length=2.2mm,width=1.6mm]},gray,opacity=0.6}}
		
		
		% Row 2 -> Row 1
		\draw[arrowformationtree] (r2_bigor) to[bend left=18] (r1_outerneg);
		
		% Row 3 -> Row 2
		\draw[arrowformationtree] (r3_C)        to[bend left=18] (r2_bigor);     
		\draw[arrowformationtree] (r3_implies) to[bend right=18] (r2_bigor);  
		
		% Row 4 -> Row 3
		\draw[arrowformationtree] (r4_D)       to[bend left=18] (r3_implies);    
		\draw[arrowformationtree] (r4_inneror) to[bend right=18] (r3_implies);   
		
		% Row 5 -> Row 4
		\draw[arrowformationtree] (r5_A)   to[bend left=18] (r4_inneror);         
		\draw[arrowformationtree] (r5_not) to[bend right=18] (r4_inneror);        
		
		% Row 6 -> Row 5
		\draw[arrowformationtree] (r6_D) to[bend left=18] (r5_not);               
	\end{tikzpicture}
\end{center}

\subsection{How to do truth tables}

With this example in mind let us try to give a general instruction for constructing a truth table for a sentence $X$.


\factoidbox{
	\begin{enumerate}
		\item Write down the formation tree of $X$. Set up the header of the truth table with lines matching the formation tree. 
		\item Consider any atomic sentences which appear in \metaX. If you are doing a truth table for multiple sentences, you'll need to also consider those appearing in the other sentences. Write all these atomic sentences in the header row, at the beginning of the table. 
		\item We now need to determine all the valuations, the different ways of assigning truth values to all these atomic sentences. If there are $n$ atomic sentences, there will be $2^n$ such valuations. Write them as the rows of the truth table. 
		\item Work up through the formation tree, which has now been written as lines in the header of the truth table to determine the truth value of the consituant sentences of \metaX. Record its truth value underneath the main connective of that sentence.
		\begin{itemize}
			\item If it is an atomic sentence, just copy the truth values from the valuation.
			\item If it is a complex sentence, apply the relevant truth rule to its consituatnt parts, remembering that their truth values are written underneath their main connectives.
		\end{itemize}
		\item Highlight or boldface the column under the main connective of the overall sentence, which provides our final answer.
\end{enumerate}}

If you follow these outlines, you should be able to construct truth tables for arbitrary TFL sentences. Of course, the more complicated the sentences are, and the more atomic sentence letters they contain the longer and tedious the truth table---but the more important it becomes to painstakingly stick to the guidelines we have given.





You can provide your answer without explicitly providing the header row matching the formation tree, unless explicitly requested. 
\begin{center}
	\begin{tabular}{r c c c| d e e e e e e e e e e e  e e f}
	{Valuation} &{$A$} &{$C$} &{$D$}
		& ${\enot}$ & $( $ & $C$ & $\eor$ & $( $ & $D$ & $\eif$ & $( $ & $A$ & $\eor$ & $\enot$ & $D$ & $)$ & $)$ & $)$  \\[.2em]\hline
		
		
		$v_1:$ & T & T & T & \TTbf{F} & & T & T & & T & T & & T & T & F & T & & & \\
		$v_2:$ & T & T & F & \TTbf{F} & & T & T & & F & T & & T & T & T & F & & & \\
		$v_3:$ & T & F & T & \TTbf{F} & & F & T & & T & T & & T & T & F & T & & & \\
		$v_4:$ & T & F & F & \TTbf{F} & & F & T & & F & T & & T & T & T & F & & & \\
		$v_5:$ & F & T & T & \TTbf{F} & & T & T & & T & F & & F & F & F & T & & & \\
		$v_6:$ & F & T & F & \TTbf{F} & & T & T & & F & T & & F & F & T & F & & & \\
		$v_7:$ & F & F & T & \TTbf{T} & & F & F & & T & F & & F & F & F & T & & & \\
		$v_8:$ & F & F & F & \TTbf{F} & & F & T & & F & T & & F & F & T & F & & & \\
		
	\end{tabular}
\end{center}




%
%
%When we do these in practice, our sentences can have many subsentences, consider, e.g.~$$\enot (B\eand(\enot B\eif \enot A)))$$ A tool for fitting our truth tables on a page is not to put all the subsentences out as separate entire columns to calculate, but to simply list the values underneath the main connective of the sentence. This is simply to keep the truth table more concise.
%
%We have a formation tree:
%\begin{center}
%		\begin{forest}
%		[$\mainconnective{\enot}(B \eand (\enot B \eif \enot A))$
%		[$\centrearoundmainconnective{(B }{\mainconnective{\eand}}{(\enot B \eif \enot A))}$
%		[$B$]
%		[$(\enot B \mainconnective{\eif} \enot A)$
%		[$\mainconnective{\enot} B$
%		[$B$]
%		]
%		[$\mainconnective{\enot} A$
%		[$A$]
%		]
%		]
%		]
%		]
%	\end{forest}
%\end{center}
%Instead of having a separate column for each subsentence:
%\begin{center}
%\begin{tabular}{cc|c|c|c|c|c}
%	$A$ & $B$ & $\enot A$ & $\enot B$ & $(\enot B \mainconnective{\eif} \enot A)$ & $(B \mainconnective{\eand} (\enot B \eif \enot A))$ & $\mainconnective{\enot}(B \eand (\enot B \eif \enot A))$ \\\hline
%	T & T & F & F & T & T & \textbf{F} \\
%	T & F & F & T & F & F & \textbf{T} \\
%	F & T & T & F & T & T & \textbf{F} \\
%	F & F & T & T & T & F & \textbf{T} \\
%\end{tabular}
%\end{center}
%
%I could write it more concisely:
%
%
%
%
%\begin{center}
%	\begin{tabular}{c c | deeeeeeeeeef}
%		\multirow{5}{*}{$A$} & \multirow{5}{*}{$B$} & $\mainconnective{\enot}$ & $($ & $B$ & $\eand$ & $($ & $\enot$ & $B$ & $\eif$ & $\enot$ & $A$ & $)$ & $)$ \\[.2em]
%	&     &         & $\mainconnectivecolor{(}$ & $B$ & $\mainconnective{\eand}$ & $($ & $\enot$ & $B$ & $\eif$ & $\enot$ & $A$ & $)$ & $\mainconnectivecolor{)}$ \\[.2em]
%	&     &         &     & $\atomicstage{B}$ &          & $\mainconnectivecolor{(}$ & $\enot$ & $B$ & $\mainconnective{\eif}$ & $\enot$ & $A$ & $\mainconnectivecolor{)}$ &  \\[.2em]
%	&     &         &     &     &          &     & $\mainconnective{\enot}$ & $B$ &        & $\mainconnective{\enot}$ & $A$ &     &     \\[.2em]
%	&     &         &     &     &          &     &         & $\atomicstage{B}$ &        &         & $\atomicstage{A}$ &     &     \\[.2em]\hline
%	T & T & \TTbf{F }&   & \textcolor{gray}{T} & T &   & F & \textcolor{gray}{T} & T & F & \textcolor{gray}{T} &   &   \\
%	T & F & \TTbf{T} &   & \textcolor{gray}{F} & F &   & T & \textcolor{gray}{F} & F & F & \textcolor{gray}{T} &   &   \\
%	F & T & \TTbf{F} &   & \textcolor{gray}{T} & F &   & F & \textcolor{gray}{T} & T & T & \textcolor{gray}{F} &   &   \\
%	F & F & \TTbf{T} &   & \textcolor{gray}{F} & F &   & T & \textcolor{gray}{F} & T & T & \textcolor{gray}{F} &   &   \\
%\end{tabular}
%\end{center}
%
%
%One can also just do the header as a single line, but then make sure you also draw the formation tree, to keep track (and for the marker to see that you know what you're doing).
%
%\begin{center}
%	\begin{tabular}{c c | deeeeeeeeeef}
%	{$A$} & {$B$} & $\mainconnective{\enot}$ & $($ & $B$ & $\eand$ & $($ & $\enot$ & $B$ & $\eif$ & $\enot$ & $A$ & $)$ & $)$   \\\hline
%	T & T & \TTbf{F }&   & \textcolor{gray}{T} & T &   & F & \textcolor{gray}{T} & T & F & \textcolor{gray}{T} &   &   \\
%	T & F & \TTbf{T} &   & \textcolor{gray}{F} & F &   & T & \textcolor{gray}{F} & F & F & \textcolor{gray}{T} &   &   \\
%	F & T & \TTbf{F} &   & \textcolor{gray}{T} & F &   & F & \textcolor{gray}{T} & T & T & \textcolor{gray}{F} &   &   \\
%	F & F & \TTbf{T} &   & \textcolor{gray}{F} & F &   & T & \textcolor{gray}{F} & T & T & \textcolor{gray}{F} &   &   \\
%\end{tabular}
%\end{center}
%
%
%
%
%One must always highlight the column under the main connective of the overall sentence, which provides our final answer. You should feel free to use whichever method you find easier.

There are tricks that mean you can miss out some of the gaps by using strategies such as: knowing that $\metaX$ is false is already enough to see that $\metaX\eand\metaY$ is false, so we don't need to continue to work out the truth value of $\metaY$. These ``shortcuts'' are discussed in \S\ref{s:PartialTruthTable}, but we won't read it in this course.

\begin{practiceproblems}\label{pr.TT.TTorC}
	\problempart
	\label{pr.TT.TTorC}
	\label{pr.TT.TTorC}
	Complete truth tables for each of the following:
	\begin{earg}
		\item $A \eif A$ %taut
		\myanswer{\begin{center}
				\begin{tabular}{c | def}
					$A$ & $A$&$\eif$&$A$\\
					\hline
					T & T&\TTbf{T}&T\\
					F & F&\TTbf{T}&F
				\end{tabular}
			\end{center}
		}
		
		\item $C \eif\enot C$ %contingent
		\myanswer{\begin{center}
				\begin{tabular}{c | d e e f}
					$C$ & $C$&$\eif$&$\enot$&$C$\\
					\hline
					T & T & \TTbf{F}& F& T\\
					F & F & \TTbf{T}& T& F\\
				\end{tabular}
		\end{center}}
		\item $(A \eif B) \eif (\enot A\eor B)$ %tautology
		
		%\myanswer{\begin{center}
				%\begin{tabular}{c c | d e e e e e e e f}
				%$A$ & $B$&$(A$&$\eif$&$B)$&$\eif$&$(\enot A$&$\eor$&$B)$ \\
				%\hline
				%T & T & T & T & T & \TTbf{T} & F & T & T \\
				%T & F & T & F & F & \TTbf{T} & F & T & T\\
				%F & T & F & F & T & \TTbf{T} & F & F & T\\
				%F & F & F & T & F & \TTbf{T} & T & F & F & T & F
				%\end{tabular}
				%\end{center}}
				\item $(A \eif B) \eor (B \eif A)$ % taut
				
				\myanswer{
					\begin{center}
						\begin{tabular}{c c | d e e e e e f}
							$A$ & $B$&$(A$&$\eif$&$B)$&$\eor$&$(B$&$\eif$&$A)$ \\
							\hline
							T & T & T & T & T & \TTbf{T} & T & T & T\\
							T & F & T & F & F & \TTbf{T} & F & T & T\\
							F & T & F & T & T & \TTbf{T} & T & F & F \\
							F & F & F & T & F & \TTbf{T} & F & T & F
						\end{tabular}
				\end{center}}
				\item $(A \eand B) \eif (B \eor A)$  %taut
				
				\myanswer{
					\begin{center}
						\begin{tabular}{c c | d e e e e e f}
							$A$ & $B$&$(A$&$\eand$&$B)$&$\eif$&$(B$&$\eor$&$A)$ \\
							\hline
							T & T & T & T & T & \TTbf{T} & T & T & T\\
							T & F & T & F & F & \TTbf{T} & F & T & T\\
							F & T & F & F & T & \TTbf{T} & T & T & F \\
							F & F & F & F & F & \TTbf{T} & F & F & F
						\end{tabular}
				\end{center}}
				\item $\enot(A \eor B) \eif (\enot A \eand \enot B)$ %taut
				
				\myanswer{\begin{center}
						\begin{tabular}{c c | d e e e e e e e e f}
							$A$ & $B$&$\enot$&$(A$&$\eor$&$B)$&$\eif$&$(\enot$&$A$&$\eand$&$\enot$&$B)$\\
							\hline
							T & T & F & T & T & T & \TTbf{T} & F & T & F & F & T\\
							T & F & F & T& T & F & \TTbf{T} & F & T & F & T & F\\
							F & T & F & F & T & T & \TTbf{T} & T & F & F & F & T\\
							F & F & T & F & F & F & \TTbf{T} & T & F & T & T & F
						\end{tabular}
				\end{center}}
				\item $(\enot A \eand \enot B) \eif\enot(A \eor B) $ %taut
				
				%\myanswer{\begin{center}
						%\begin{tabular}{c c | d e e e e e e e e f}
						%$A$ & $B$&$\enot$&$(A$&$\eor$&$B)$&$\leftarrow$&$(\enot$&$A$&$\eand$&$\enot$&$B)$\\
						%\hline
						%T & T & F & T & T & T & \TTbf{T} & F & T & F & F & T\\
						%T & F & F & T& T & F & \TTbf{T} & F & T & F & T & F\\
						%F & T & F & F & T & T & \TTbf{T} & T & F & F & F & T\\
						%F & F & T & F & F & F & \TTbf{T} & T & F & T & T & F
						% \end{tabular}
					%\end{center}}
					
					
					\item $\bigl[(A\eand B) \eand\enot(A\eand B)\bigr] \eand C$ %contradiction
					
					\myanswer{\begin{center}
							\begin{tabular}{c c c | d e e e e e e e e f}
								$A$ & $B$&$C$&$\bigl[(A$&$\eand$&$B)$&$ \eand$&$\enot$&$(A$&$\eand$&$B)\bigr]$&$\eand$&$C$\\
								\hline
								T & T & T & T & T & T & F & F & T & T & T & \TTbf{F} & T\\
								T & T & F & T& T & T & F & F & T & T & T & \TTbf{F}& F\\
								T & F & T & T & F & F & F & T & T & F & F & \TTbf{F} & T\\
								T & F & F & T & F & F & F & T & T & F & F & \TTbf{F} & F\\
								F & T & T & F & F & T & F & T & F & F & T & \TTbf{F} & T\\
								F & T & F & F & F & T & F & T & F & F & T & \TTbf{F} & F\\
								F & F & T & F & F & F & F & T & F & F & F & \TTbf{F} & T\\
								F & F & F & F & F & F & F & T & F & F & F & \TTbf{F} & F
							\end{tabular}
					\end{center}}
					\item $[(A \eand B) \eand C] \eif B$ %taut
					
					\myanswer{\begin{center}
							\begin{tabular}{c c c | d e e e e e f}
								$A$ & $B$&$C$&$[(A$&$\eand$&$B)$&$\eand$&$C]$&$\eif$&$B$\\
								\hline
								T & T & T & T & T & T & T & T & \TTbf{T} & T\\
								T & T & F & T & T & T & F & F & \TTbf{T} & T\\
								T & F & T & T & F & F & F & T & \TTbf{T} & F\\
								T & F & F & T & F & F & F & F & \TTbf{T} & F\\
								F & T & T & F & F & T & F & T & \TTbf{T} & T\\
								F & T & F & F & F & T & F & F & \TTbf{T} & T\\
								F & F & T & F & F & F & F & T & \TTbf{T} & F\\
								F & F & F & F & F & F & F & F & \TTbf{T} & F\\
							\end{tabular}
					\end{center}}
					\item $\enot\bigl[(C\eor A) \eor B\bigr]$ %contingent
					
					\myanswer{\begin{center}
							\begin{tabular}{c c c | d e e e e f}
								$A$ & $B$&$C$&$\enot\bigl[($&$C$&$\eor$&$A)$&$\eor$&$B\bigr]$\\
								\hline
								T & T & T & \TTbf{F} & T & T & T & T & T\\
								T & T & F & \TTbf{F} & F & T & T & T & T\\
								T & F & T & \TTbf{F} & T & T & T & T & F\\
								T & F & F & \TTbf{F} & F & T & T & T & F\\
								F & T & T & \TTbf{F} & T & T & F & T & T\\
								F & T & F & \TTbf{F} & F & F & F & T & T\\
								F & F & T & \TTbf{F} & T & T & F & T & F\\
								F & F & F & \TTbf{T} & F & F & F & F & F
							\end{tabular}
					\end{center}}

	
	
%	\item $\enot [(X \eand Y) \eor (X \eor Y)]$
%	
%	\myanswer{
%		\begin{center}
%			\begin{tabular}{c|c|ccccccc}
%				\cline{2-2}
%				~	&	\enot	&	 [(X 	&	\eand& 	Y) 	&	\eor 	&	(X 	&	\eor 	&	Y)] \\
%				\cline{2-9}
%				&	F	&	T	&	T	&	T	&	T	&	T	&	T	&	T	\\
%				&	F	&	T	&	F	&	F	&	T	&	T	&	T	&	F	\\
%				&	F	&	F	&	F	&	T	&	T	&	F	&	T	&	T	\\
%				&	T	&	F	&	F	&	F	&	F	&	F	&	F	&	F	\\
%				\cline{2-2}
%			\end{tabular}
%		\end{center}
%	}
	
	
	\item $(A \eif B) \eif (\enot B\eif \enot A)$
	
	\myanswer{
		\begin{center}
			\begin{tabular}{cccc|c|ccccc}
				\cline{5-5}
				~	&	(A 	&	\eif	&	B)	&	 \eif 	&	(\enot&	B 	&	\eif 	&	 \enot 	& 	 A) \\
				\cline{2-10}
				&	T	&	T	&	T	&	T		&	F	 &	T	&	T	&	F		&	T	\\
				&	T	&	F	&	F	&	T		&	T	 &	F	&	F	&	F		&	T	\\
				&	F	&	T	&	T	&	T		&	F	 &	T	&	T	&	T		&	F	\\
				&	F	&	T	&	F	&	T		&	T	 &	F	&	T	&	T		&	F	\\
				\cline{5-5}
			\end{tabular}
		\end{center}
	}
	
	\item $[C \eif (D \eor E)] \eand \enot C$
	
	%\myanswer{
		%\begin{center}
		%\begin{tabular}{cccccc|c|cc}
		%\cline{7-7}
		%~	&	[C 	&	\eiff 	&	(D 	&	\eor 	&	E)] 	&	\eand 	&	 \enot 	&	 C \\
		%\cline{2-9}
		%	&	T	&	T	&	T	&	T	&	T	&	F		&	F		&	T	\\
		%	&	T	&	T	&	T	&	T	&	F	&	F		&	F		&	T	\\
		%	&	T	&	T	&	F	&	T	&	T	&	F		&	F		&	T	\\
		%	&	T	&	F	&	F	&	F	&	F	&	F		&	F		&	T	\\
		%	&	F	&	F	&	T	&	T	&	T	&	F		&	T		&	F	\\
		%	&	F	&	F	&	T	&	T	&	F	&	F		&	T		&	F	\\
		%	&	F	&	F	&	F	&	T	&	T	&	F		&	T		&	F	\\
		%	&	F	&	T	&	F	&	F	&	F	&	T		&	T		&	F	\\
		%\cline{7-7}
		%\end{tabular}
		%\end{center}
		%}
	
	\item $\enot(G \eand (B \eand H)) \eif (G \eor (B \eor H))$
	
	%\myanswer{
		%\begin{center}
		%\begin{tabular}{ccccccc|c|ccccc}
		%\cline{8-8}
		%~	&\enot&	(G 	&\eand &	(B 	&	 \eand 	&	 H))	&	\eiff 	&	(G 	& \eor 	& (B 	& \eor	& H))	\\
		%\cline{2-13}
		%	&F	   &	T	&	  T &	T	&	T		&	T	&	F	&	T	&	T	&	T	&	T	&	T	\\
		%	&T	   &	T	&	  F &	T	&	F		&	F	&	T	&	T	&	T	&	T	&	T	&	F	\\
		%	&T	   &	T	&	 F  &	F	&	F		&	T	&	T	&	T	&	T	&	F	&	T	&	T	\\
		%	&T	   &	T	&	 F  &	F	&	F		&	F	&	T	&	T	&	T	&	F	&	F	&	F	\\
		%	&T	   &	F	&	F   &	T	&	T		&	T	&	T	&	F	&	T	&	T	&	T	&	T	\\
		%	&T	   &	F	&	F   &	T	&	F		&	F	&	T	&	F	&	T	&	T	&	T	&	F	\\
		%	&T	   &	F	&	F   &	F	&	F		&	T	&	T	&	F	&	T	&	F	&	T	&	T	\\
		%	&T	   &	F	&	F   &	F	&	F		&	F	&	F	&	F	&	F	&	F	&	F	&	F	\\
		%\cline{8-8}
		%\end{tabular}
		%\end{center}
		%}
	
	\vspace{1em}
	
\end{earg}

\problempart
Write complete truth tables for the following sentences and mark the column that represents the possible truth values for the whole sentence.

\begin{earg}
	
	\item	$(D \eand \enot D) \eif G $
	
	
	\myanswer{\vspace{1em}
		\begin{center}
			\begin{tabular}{ccccc|c|c}
				\cline{6-6}
				&	(D 	&	 \eand 	& 	 \enot	&	 D) 	&	 \eif 	&	 G \\
				\hline
				&	T	&	F		&	F		&	T	&	T	&	T	\\
				&	T	&	F		&	F		&	T	&	T	&	F	\\
				&	F	&	F		&	T		&	F	&	T	&	T	\\
				&	F	&	F		&	T		&	F	&	T	&	F	\\
				\cline{6-6}
			\end{tabular}
		\end{center}
	}
	\vspace{1em}
	
	
	\item	$(\enot P \eor \enot M) \eif M $
	
	%\myanswer{
		%\begin{center}
		%\begin{tabular}{cccccc|c|c}
		%\cline{7-7}
		%	&	(\enot 	&	P 	&	\eor 	&	\enot 	& 	 M) 	& 	\eiff 	&	 M \\
		% \hline
		%	&	F		&	T	&	F	&	F		&	T	&	F	&	T	\\
		%	&	F		&	T	&	T	&	T		&	F	&	F	&	F	\\
		%	&	T		&	F	&	T	&	F		&	T	&	T	&	T	\\
		%	&	T		&	F	&	T	&	T		&	F	&	F	&	F	\\
		%\cline{7-7}
		%\end{tabular}
		%\end{center}
		%}
	\vspace{1em}
	
	
	
	\item	$\enot \enot (\enot A \eand \enot B)  $
	
	\myanswer{
		\begin{center}
			\begin{tabular}{c|c|cccccc}
				\cline{2-2}
				&	\enot		&	 \enot 	&	(\enot 	& 	 A 	& \eand 	& 	\enot 	&	 B)  \\
				\hline
				&	F		&	T		&	F		&	T	&	F	&	F		&	T	\\
				&	F		&	T		&	F		&	T	&	F	&	T		&	F	\\
				&	F		&	T		&	T		&	F	&	F	&	F		&	T	\\
				&	T		&	F		&	T		&	F	&	T	&	T		&	F	\\
				\cline{2-2}
			\end{tabular}
		\end{center}
	}
	\vspace{1em}
	
	
	
	\item 	$[(D \eand R) \eif I] \eif \enot(D \eor R) $
	
	\myanswer{
		\begin{center}
			\begin{tabular}{cccccc|c|cccc}
				\cline{7-7}
				&	[(D 	& 	 \eand 	& 	 R)	& 	\eif 	&	I] 	&	\eif 	&	 \enot 	&	(D 	&	 \eor 	& R) \\
				\hline
				&	T	&	T		&	T	&	T	&	T	&	F	&	F		&	T	&	T		&T	\\
				&	T	&	T		&	T	&	F	&	F	&	T	&	F		&	T	&	T		&T	\\
				&	T	&	F		&	F	&	T	&	T	&	F	&	F		&	T	&	T		&F	\\
				&	T	&	F		&	F	&	T	&	F	&	F	&	F		&	T	&	T		&F	\\
				&	F	&	F		&	T	&	T	&	T	&	F	&	F		&	F	&	T		&T	\\
				&	F	&	F		&	T	&	T	&	F	&	F	&	F		&	F	&	T		&T	\\
				&	F	&	F		&	F	&	T	&	T	&	T	&	T		&	F	&	F		&F	\\
				&	F	&	F		&	F	&	T	&	F	&	T	&	T		&	F	&	F		&F	\\
				\cline{7-7}
			\end{tabular}
		\end{center}
	}
	
	\vspace{1em}
	
	
	%\item	$\enot [(D \eiff O) \eiff A] \eif (\enot D \eand O) $
	%\myanswer{
		%&\begin{center}
		%\begin{tabular}{ccccccc|c|cccc}
		%\cline{8-8}
		%	&	\enot 	&	[(D 	&	\eiff 	&	O) 	&	\eiff 	&	 A]	& 	\eif 	 &	(\enot 	& 	D 	 & 	 \eand &O) \\
		%	\hline
		%	&	F		&	T	&	T	&	T	&	T	&	T	&	T	&	F		&	T	&	F	&T	\\
		%	&	T		&	T	&	T	&	T	&	F	&	F	&	F	&	F		&	T	&	F	&T	\\
		%	&	T		&	T	&	F	&	F	&	F	&	T	&	F	&	F		&	T	&	F	&F	\\
		%	&	F		&	T	&	F	&	F	&	T	&	F	&	T	&	F		&	T	&	F	&F	\\
		%	&	T		&	F	&	F	&	T	&	F	&	T	&	T	&	T		&	F	&	T	&T	\\
		%	&	F		&	F	&	F	&	T	&	T	&	F	&	T	&	T		&	F	&	T	&T	\\
		%	&	F		&	F	&	T	&	F	&	T	&	T	&	T	&	T		&	F	&	F	&F	\\
		%	&	T		&	F	&	T	&	F	&	F	&	F	&	F	&	T		&	F	&	F	&F	\\
		%\cline{8-8}
		%\end{tabular}
		%\end{center}
		%}
	%\vspace{1em}
\end{earg}





\problempart
Can you think of sentences with the following truth table:
\begin{enumerate}
	\item \begin{tabular}{cc|c}
		$A$&$B$&?\\\hline
		T&T&T\\
		T&F&T\\
		F&T&T\\
		F&F&T
	\end{tabular}\myanswer{$A\eor\enot A$}
	\item \begin{tabular}{cc|c}
		$A$&$B$&?\\\hline
		T&T&F\\
		T&F&F\\
		F&T&T\\
		F&F&F
	\end{tabular}\myanswer{$\enot A\eand B$}
	\item \begin{tabular}{cc|c}
		$A$&$B$&?\\\hline
		T&T&T\\
		T&F&F\\
		F&T&T\\
		F&F&T
	\end{tabular}\myanswer{\quad\begin{minipage}{.75\textwidth}
			$\enot( A\eand \enot B)$;\\ or a more systematic answer by simply writing out the different options when the sentence is true: $$((A\eand B)\eor (\enot A\eand B))\eor (\enot A\eand\enot B)$$
	\end{minipage}}
	
\end{enumerate}
\problempart
Suppose $X$ is TFL sentence containing two atomic sentences. Then there are in fact sixteen different possible columns for $X$.
\begin{enumerate}
	\item Can you explain why?\myanswer{\\ The first line might have T or F, the second line might have T or F, etc. There are 4 valuations, so there are $2\times 2\times 2\times 2=2^4=16$ different ways of putting Ts and Fs to these 4 lines. I.e.~16 different columns. }
	\item Can you show for each of these combinations that there is a sentence of TFL with that column describing its truth.
	%\myanswer{\\
		%See section 37.2 of forall$x$:Bristol for the general argument for this.
		%}
	\item {} Can you show there's always a formula just using $\enot$ and $\eand$ with that column describing its truth.
	%\myanswer{\\
		%The previous answer had that every truth table can be given by a sentence with $\eand, \eor$ and $\enot$. We can replace any instances of $\eor$ by $\eand$ using: $$\metaX\eor\metaY\text{ is logically equivalent to }\enot(\enot\metaX\eand\enot\metaY)$$
		%See 37.4 of forall$x$:Bristol for more details.
		%}
\end{enumerate}

If you want additional practice, you can construct truth tables for any of the sentences and arguments in the exercises for the previous chapter.

\end{practiceproblems}

\chapter{Validity and the method of truth tables}
\section{Validity}

\factoidbox{\label{TFLvalid}A TFL-argument $X_1,\ldots,X_n\therefore Y$ is \define{valid} if and only if there is no valuation $v$ such that all the premises, $X_1,\ldots,X_n$ are true relative to $v$ but $Y$ is false relative to $v$.}




This can be determined by the method of truth tables.

Consider the argument:
$$\enot L \eif (J \eor L),\enot J\therefore (\enot L\eif J)$$
We need to check whether there is any valuation which makes both $\enot L \eif (J \eor L)$ and $\enot J$ true whilst making $(\enot L\eif J) $ false. 

We so far have only done complete truth tables for single sentences. Now we need to do a complete truth table which simultaneously works for the three sentences, $\enot L \eif (J \eor L)$, $\enot J$ and $(\enot L\eif J)$.


The atomic sentences that appear in the premises and conclusion are $J$ and $L$. So we will need to fill out a truth table

\begin{center}
	\begin{tabular}{rcc|c|c|c}
		Valuation&$J$&$L$&$\enot L \eif (J \eor L)$&$\enot J$ & $(\enot L\eif J)$\\\hline
		$v_1:$&T&T&&&\\
		$v_2:$&T&F&&&\\
		$v_3:$&F&T&&&\\
		$v_4:$&F&F&&
	\end{tabular}
\end{center}

We will work up through the formation tree of each of our three sentences, starting with the leaves, the atomic sentences, for which we consult the valuation, and applying the truth rules to determine the truth values of higher up sentences on that valuation. 

The result will be:

\begin{center}
	\begin{tabular}{r c c | d e e eeeeee f | d f | d eeee e f}
		\multirow{3}{*}{Valuation} & \multirow{3}{*}{$J$} & \multirow{3}{*}{$L$} 
		& $\mainconnectivecolor{(}$ & $\enot$ & $L$ & $\mainconnective{\eif}$ & $($ & $J$ & $\eor$ & $L$ & $)$ & $\mainconnectivecolor{)}$
		& $\mainconnective{\enot}$ & $J$
		& $\mainconnectivecolor{(}$ & $\enot$ & $L$ & $\mainconnective{\eif}$ & $J$ & $\mainconnectivecolor{)}$ \\[.3em]
		&&& & $\mainconnective{\enot}$ & $L$ & & $\mainconnectivecolor{(}$ & $J$ & $\mainconnective{\eor}$ & $L$ & $\mainconnectivecolor{)}$ & 
		& & $\atomicstage{J}$
		&& $\mainconnective{\enot}$ & $L$ & & $\atomicstage{J}$ & \\[.3em]
		&&& & & $\atomicstage{L}$ & & & $\atomicstage{J}$ & & $\atomicstage{L}$ & & 
		&& 
		&& &$\atomicstage{L}$ & &  & \\[.3em]
		\hline
		$v_1:$ & T & T &
		& F & \textcolor{gray!20}{T} & \TTbf{T} & & \textcolor{gray!20}{T} & T & \textcolor{gray!20}{T} & & &
		\TTbf{F} & \textcolor{gray!20}{T} &
		& F & \textcolor{gray!20}{T} & \TTbf{T} & \textcolor{gray!20}{T} & \\
		$v_2:$ & T & F &
		& T & \textcolor{gray!20}{F} & \TTbf{T} & & \textcolor{gray!20}{T} & T & \textcolor{gray!20}{F} & & &
		\TTbf{F} & \textcolor{gray!20}{T} &
		& T & \textcolor{gray!20}{F} & \TTbf{T} & \textcolor{gray!20}{T} & \\
		$v_3:$ & F & T &
		& F & \textcolor{gray!20}{T} & \TTbf{T} & & \textcolor{gray!20}{F} & T & \textcolor{gray!20}{T} & & &
		\TTbf{T} & \textcolor{gray!20}{F} &
		& F & \textcolor{gray!20}{T} & \TTbf{T} & \textcolor{gray!20}{F} & \\
		$v_4:$ & F & F &
		& T & \textcolor{gray!20}{F} & \TTbf{F} & & \textcolor{gray!20}{F} & F & \textcolor{gray!20}{F} & & &
		\TTbf{T} & \textcolor{gray!20}{F} &
		& T & \textcolor{gray!20}{F} & \TTbf{F} & \textcolor{gray!20}{F} & \\
	\end{tabular}
\end{center}


The only valuation on which both `$\enot L \eif (J \eor L)$' and `$\enot J$' are true is $v_3$, and on $v_3$, `$\enot L\eif J$' is true. So there is no valuation where the premises are true and the conclusion is false. So the argument is valid. 





\begin{highlighted}
	To check {validity} with truth tables, draw a complete truth table for all the sentences simultaneously and look for a row like:
	
	%	To check \alert{validity} with truth tables, look for a row like:
	%\smallskip 
	\begin{center}
		\begin{tabular}{r || c |c |c || c c}
			Valuation & Premise $1$ & Premise $2$ & $\cdots$ & Conclusion\\
			\hline
			$v_1$&&&&\\
			$\bigstar$ $v_2$  & T & T & T & {F} &$\leftsquigarrow$ counterexample\\
			$v_3$&&&&\\
			$v_4$&&&&
		\end{tabular}\\ \textbf{Invalid}
	\end{center}
	%	
	
	%\smallskip
	
	If there is no such row, then it is \emph{valid}. 
\end{highlighted}
(In fact, for full marks to it is invalid, one only needs to give such a counterexample valuation, not complete truth tables.)

We will also say:
\begin{highlighted}
	\metaY is a \define{logical consequence} of $\metaX_1,\ldots,\metaX_n$ if and only if $X_1,\ldots,X_n\therefore Y$ is {valid}
\end{highlighted}

One needs to check every valuation to see if there is one which is a counterexample valuation. 

Consider all assignments of truth values to all the atomic sentences that appear in the premises or the conclusion.


\subsection{Example}

%\begin{earg}
%	\prem If Jones was drunk when he signed the contract, then the contract is void.
%	\prem Jones was not drunk when he signed the contract. 
%	\conc The contract is not void.
%\end{earg}

$D\eif V,\;\enot D\;\therefore\;\enot V$

\begin{center}
	\begin{tabular}{r c c | def | df | df}
		&&&\multicolumn{3}{c}{Premise 1}&\multicolumn{2}{c}{Premise 2}&\multicolumn{2}{c}{Conclusion}\\
		Valuation & $D$ & $V$ & $D$ & $\mainconnective{\eif}$ & $V$ & $\mainconnective{\enot}$ & $D$ & $\mainconnective{\enot}$ & $V$\\[.3em]
		%		Valuation & $D$ & $V$ & $\atomicstage{D}$ & &$\atomicstage{V}$& & $\atomicstage{D}$ &  & $\atomicstage{V}$\\
		\hline
		$v_1:$ & T & T & T & \TTbf{T} & T & \TTbf{F} & T & \TTbf{F} & T\\
		$v_2:$ & T & F & T & \TTbf{F} & F & \TTbf{F} & T & \TTbf{T} & F\\
		$\bigstar$ $v_3:$ & F & T & F & \TTbf{\bfseries T} & T & \TTbf{\bfseries T} & F & \TTbf{\bfseries F} & T\\
		$v_4:$ & F & F & F & \TTbf{T} & F & \TTbf{T} & F & \TTbf{T} & F\\
	\end{tabular}
\end{center}


On $v_3$, the premises are true and the conclusion is false. It is invalid.

\subsection{Example}
Compare: 
$D\eif V,\;D\;\therefore\;V$

\begin{center}
	\begin{tabular}{r c c | def | c | c}
		&&&\multicolumn{3}{c}{Premise 1}&\multicolumn{1}{c}{Premise 2}&\multicolumn{1}{c}{Conclusion}\\
		Valuation & $D$ & $V$ & $D$ & $\mainconnective{\eif}$ & $V$ & $\atomicstage{D}$  & $\atomicstage{V}$ \\[.3em]
		%		Valuation & $D$ & $V$ & $\atomicstage{D}$ & & $\atomicstage{V}$ &  & \\
		\hline
		$v_1:$ & T & T & T & \TTbf{T} & T & \TTbf{T} &\TTbf{T}\\
		$v_2:$ & T & F & T & \TTbf{F} & F & \TTbf{T} & \TTbf{F}\\
		$v_3:$ & F & T & F & \TTbf{T} & T & \TTbf{F} & \TTbf{T}\\
		$v_4:$ & F & F & F & \TTbf{T} & F & \TTbf{F} & \TTbf{F}\\
	\end{tabular}
\end{center}
$v_1$ is the only valuation where premises are both true, and on it, conclusion is also true. So is valid. 

\subsection{Example}

%\begin{earg}
%	\prem If I have a pet, then I have a cat or a dog.
%	\prem I do not have a cat.
%	\conc If I have a pet, then I have a dog. 
%\end{earg}
$P\eif(C\eor D)\,,\,\enot C\;\therefore P\eif D$

\begin{center}
	\begin{tabular}{r c c c | d e e e e e f | d f || d e e e f}
		&&&&\multicolumn{7}{c}{Premise 1} & \multicolumn{2}{c}{Premise 2} & \multicolumn{5}{c}{Conclusion}\\
		% Row 1: main connectives and structure
		Valuation & $P$ & $C$ & $D$
		& $P$ & $\mainconnective{\eif}$ & $\mainconnectivecolor{(}$ & $C$ & $\eor$ & $D$ & $\mainconnectivecolor{)}$
		& $\mainconnective{\enot}$ & $C$
		& $\mainconnectivecolor{(}$ & $P$ & $\mainconnective{\eif}$ & $D$ & $\mainconnectivecolor{)}$ \\[.3em]
		%		% Row 2: children of Row 1
		%		&&&
		%		& $\atomicstage{P}$ &  & $\mainconnectivecolor{(}$ & $C$ & $\mainconnective{\eor}$ & $D$ & $\mainconnectivecolor{)}$
		%		&  & $\atomicstage{C}$
		%		&  & $\atomicstage{P}$ &  & $\atomicstage{D}$ &  \\[.3em]
		\hline
		% v1
		$v_1:$ & T & T & T
		& \textcolor{gray!20}{T} & \TTbf{T} &  & \textcolor{gray!20}{T} & T & \textcolor{gray!20}{T} & 
		& \TTbf{F} & \textcolor{gray!20}{T}
		&  & \textcolor{gray!20}{T} & \TTbf{T} & \textcolor{gray!20}{T} &  \\
		% v2
		$v_2:$ & T & T & F
		& \textcolor{gray!20}{T} & \TTbf{T} &  & \textcolor{gray!20}{T} & T & \textcolor{gray!20}{F} & 
		& \TTbf{F} & \textcolor{gray!20}{T}
		&  & \textcolor{gray!20}{T} & \TTbf{F} & \textcolor{gray!20}{F} &  \\
		% v3
		$v_3:$ & T & F & T
		& \textcolor{gray!20}{T} & \TTbf{T} &  & \textcolor{gray!20}{F} & T & \textcolor{gray!20}{T} & 
		& \TTbf{T} & \textcolor{gray!20}{F}
		&  & \textcolor{gray!20}{T} & \TTbf{T} & \textcolor{gray!20}{T} &  \\
		% v4
		$v_4:$ & T & F & F
		& \textcolor{gray!20}{T} & \TTbf{F} &  & \textcolor{gray!20}{F} & F & \textcolor{gray!20}{F} & 
		& \TTbf{T} & \textcolor{gray!20}{F}
		&  & \textcolor{gray!20}{T} & \TTbf{F} & \textcolor{gray!20}{F} &  \\
		% v5
		$v_5:$ & F & T & T
		& \textcolor{gray!20}{F} & \TTbf{T} &  & \textcolor{gray!20}{T} & T & \textcolor{gray!20}{T} & 
		& \TTbf{F} & \textcolor{gray!20}{T}
		&  & \textcolor{gray!20}{F} & \TTbf{T} & \textcolor{gray!20}{T} &  \\
		% v6
		$v_6:$ & F & T & F
		& \textcolor{gray!20}{F} & \TTbf{T} &  & \textcolor{gray!20}{T} & T & \textcolor{gray!20}{F} & 
		& \TTbf{F} & \textcolor{gray!20}{T}
		&  & \textcolor{gray!20}{F} & \TTbf{T} & \textcolor{gray!20}{F} &  \\
		% v7
		$v_7:$ & F & F & T
		& \textcolor{gray!20}{F} & \TTbf{T} &  & \textcolor{gray!20}{F} & T & \textcolor{gray!20}{T} & 
		& \TTbf{T} & \textcolor{gray!20}{F}
		&  & \textcolor{gray!20}{F} & \TTbf{T} & \textcolor{gray!20}{T} &  \\
		% v8
		$v_8:$ & F & F & F
		& \textcolor{gray!20}{F} & \TTbf{T} &  & \textcolor{gray!20}{F} & F & \textcolor{gray!20}{F} & 
		& \TTbf{T} & \textcolor{gray!20}{F}
		&  & \textcolor{gray!20}{F} & \TTbf{T} & \textcolor{gray!20}{F} &  \\
	\end{tabular}
\end{center}

This is valid.


\subsection{Example}

$A\eor B,\,\enot (B\eand \enot C)\,\therefore \, \enot C\eor A$\begin{center}
	\begin{tabular}{r c c c | def | deeeeef | deeeef}
		&&&& \multicolumn{3}{c}{Premise 1} & \multicolumn{7}{c}{Premise 2} & \multicolumn{6}{c}{Conclusion}\\
		Val & $A$ & $B$ & $C$
		& $A$ & \mainconnective{$\eor$} & $B$
		& \mainconnective{$\enot$} & {(} & $B$ & {$\eand$} & {$\enot$} & $C$ & {)}
		& {(} & {$\enot$} & $C$ & \mainconnective{$\eor$} & $A$ & {)} \\[.3em]
		\hline
		$v_1:$ & T & T & T
		& T & \TTbf{T} & T
		& \TTbf{T} &  & T & F & F & T &
		&  & F & T & \TTbf{T} & T &  \\
		$v_2:$ & T & T & F
		& T & \TTbf{T} & T
		& \TTbf{F} &  & T & T & T & F &
		&  & T & F & \TTbf{T} & T &  \\
		$v_3:$ & T & F & T
		& T & \TTbf{T} & F
		& \TTbf{T} &  & F & F & F & T &
		&  & F & T & \TTbf{T} & T &  \\
		$v_4:$ & T & F & F
		& T & \TTbf{T} & F
		& \TTbf{T} &  & F & F & T & F &
		&  & T & F & \TTbf{T} & T &  \\
		$\bigstar$ $v_5:$ & F & T & T
		& F & \TTbf{T} & T
		& \TTbf{T} &  & T & F & F & T &
		&  & F & T & \TTbf{F} & F &  \\
		$v_6:$ & F & T & F
		& F & \TTbf{T} & T
		& \TTbf{F} &  & T & T & T & F &
		&  & T & F & \TTbf{T} & F &  \\
		$v_7:$ & F & F & T
		& F & \TTbf{F} & F
		& \TTbf{T} &  & F & F & F & T &
		&  & F & T & \TTbf{F} & F &  \\
		$v_8:$ & F & F & F
		& F & \TTbf{F} & F
		& \TTbf{T} &  & F & F & T & F &
		&  & T & F & \TTbf{T} & F &  \\
	\end{tabular}
\end{center}
On $v_5$, premises are both true but conclusion is false. The argument is invalid. 



%\subsection{Example}
%Sometimes we need to consider arguments with more complex logical form. E.g., 
%
%\begin{earg}
%	\prem If it's not raining, then Jones is coming or it is raining.
%	\prem Jones isn't coming
%	\conc If it isn't raining, then Jones is coming.
%\end{earg}
%
%$$\enot R \eif (J \eor R),\enot J\therefore (\enot R\eif J)$$
%
%\begin{center}
%	\begin{tabular}{r c c | d e e eeeeee f | d f || d eeee e f}
	%		&&&\multicolumn{10}{c}{Premise 1}&\multicolumn{2}{c}{Premise 2}&\multicolumn{5}{c}{Conclusion}\\
	%		% Row 1: main connectives and structure
	%		&&
	%		& $\mainconnectivecolor{(}$ & $\enot$ & $R$ & $\mainconnective{\eif}$ & $($ & $J$ & $\eor$ & $R$ & $)$ & $\mainconnectivecolor{)}$
	%		& $\mainconnective{\enot}$ & $J$
	%		& $\mainconnectivecolor{(}$ & $\enot$ & $R$ & $\mainconnective{\eif}$ & $J$ & $\mainconnectivecolor{)}$ \\[.3em]
	%		% Row 2: children of Row 1; parent of Row 3
	%		&&& & $\mainconnective{\enot}$ & $R$ & & $\mainconnectivecolor{(}$ & $J$ & $\mainconnective{\eor}$ & $R$ & $\mainconnectivecolor{)}$ & 
	%		& & $\atomicstage{J}$
	%		&& $\mainconnective{\enot}$ & $R$ & & $\atomicstage{J}$ & \\[.3em]
	%		% Row 3: children of Row 2; parent of Row 4. Also includes valuation as its the last row.
	%		Valuation&$J$&$R$& & & $\atomicstage{R}$ & & & $\atomicstage{J}$ & & $\atomicstage{R}$ & & 
	%		&& 
	%		&& &$\atomicstage{R}$ & &  & \\[.3em]
	%		\hline
	%		% Row 4: valuation v1
	%		$v_1:$ & T & T &
	%		& F & \textcolor{gray!20}{T} & \TTbf{T} & & \textcolor{gray!20}{T} & T & \textcolor{gray!20}{T} & & &
	%		\TTbf{F} & \textcolor{gray!20}{T} &
	%		& F & \textcolor{gray!20}{T} & \TTbf{T} & \textcolor{gray!20}{T} & \\
	%		% Row 5: valuation v2
	%		$v_2:$ & T & F &
	%		& T & \textcolor{gray!20}{F} & \TTbf{T} & & \textcolor{gray!20}{T} & T & \textcolor{gray!20}{F} & & &
	%		\TTbf{F} & \textcolor{gray!20}{T} &
	%		& T & \textcolor{gray!20}{F} & \TTbf{T} & \textcolor{gray!20}{T} & \\
	%		% Row 6: valuation v3
	%		$v_3:$ & F & T &
	%		& F & \textcolor{gray!20}{T} & \TTbf{T} & & \textcolor{gray!20}{F} & T & \textcolor{gray!20}{T} & & &
	%		\TTbf{T} & \textcolor{gray!20}{F} &
	%		& F & \textcolor{gray!20}{T} & \TTbf{T} & \textcolor{gray!20}{F} & \\
	%		% Row 7: valuation v4
	%		$v_4:$ & F & F &
	%		& T & \textcolor{gray!20}{F} & \TTbf{F} & & \textcolor{gray!20}{F} & F & \textcolor{gray!20}{F} & & &
	%		\TTbf{T} & \textcolor{gray!20}{F} &
	%		& T & \textcolor{gray!20}{F} & \TTbf{F} & \textcolor{gray!20}{F} & \\
	%	\end{tabular}
%\end{center}
%
%
%The argument is valid. 

\subsection{Inconsistent premises}
\begin{earg}
	\prem Jenny is happy.
	\prem Jenny is not happy.
	\conc The moon is made of cheese.
\end{earg}

$$H,\,\enot H\,\therefore\, C$$

\begin{center}
	\begin{tabular}{r c c | c | c | c}
		&&& Premise 1 & Premise 2 & Conclusion\\
		Valuation& $H$ & $C$ & $H$ & $\enot H$ & $C$\\[.3em]
		\hline
		$v_1:$ & T & T & T & F & T\\
		$v_2:$ & T & F & T & F & F\\
		$v_3:$ & F & T & F & T & T\\
		$v_4:$ & F & F & F & T & F\\
	\end{tabular}
\end{center}
Valid. There is no counterexample valuation!

We can make a more general observation: any argument with premises that are \emph{inconsistent} is valid. We will now carefully introduce this notion of consistency and other logical notions. 

\chapter{Other logical notions}

\section{Consistency}

We have just seen an example where the premises are inconsistent: Here's the careful definition of this:


\begin{highlighted}
	A collection of TFL-sentences $X_1,\ldots,X_n$ (with possibly $n=1$) is \define{consistent} iff there is a valuation relative to which $X_1,\ldots,X_n$ are true. Otherwise the collection is \define{inconsistent}.
\end{highlighted}

\newglossaryentry{consistency in TFL}
{
	name=consistency (in TFL),
	text=consistent,
	description={A property held by sentences if and only if there is some \gls{valuation} that makes all the sentences true}
}




\begin{highlighted}
		To check if a collection is \define{consistent} with truth tables, look for a row like:
	
	
	\smallskip 
	\begin{center}
%		\begin{minipage}{.5\linewidth}
%			\begin{center}
				\begin{tabular}{r || c |c |c | c}
					Valns & $\metaX_1$ & $\metaX_2$ & $\cdots$ & $\metaX_n$\\
					\hline
					$v_1$&&&&\\
					$\bigstar$ $v_2$  & T & T & T & {T} \\
					$v_3$&&&&\\
					$v_4$&&&&
				\end{tabular}\\ \textbf{consistent}
%			\end{center}
%		\end{minipage}%
\end{center}

If there is no such row then it is inconsistent. 
	%	
	
	%\smallskip
	
	%	If no such row exists, then the set is \textbf{inconsistent}.
\end{highlighted}

`$H$' and `$\enot H$' are inconsistent. 

Inconsistent collections have truth tables more like: 
%\begin{minipage}{.5\linewidth}
	\begin{center}
		\begin{tabular}{r || c |c |c | c}
			Valns & $\metaX_1$ & $\metaX_2$ & $\cdots$ & $\metaX_n$\\
			\hline
			$v_1$&F&&&\\
			$v_2$  &&F&& \\
			$v_3$&&&&F\\
			$v_4$&&&F&
		\end{tabular}\\ \textbf{inconsistent}
	\end{center}
%\end{minipage}
%\end{center}




\section{Tautologies and contradictions}\label{s:TautologiesAndContradictions}

\factoidbox{
	A sentence $\metaX$ of TFL is a \define{logical truth} or \define{tautology} if and only if it is true on every valuation.
}

\newglossaryentry{tautology}
{
	name=tautology,
	description={A sentence that is true on every \gls{valuation}}
}

``Tautology'' is a name which is specific to TFL. The notion of being a ``logical truth'' is broader and can also apply in the case of First Order Logic. 

We can determine whether a sentence is a tautology just by using truth tables. A complete truth table had a line for each valuation, so we just need to check if it is true on every line of the truth table. 


Our main example of a tautology is $A\eor\enot A$. Whatever the weather is like you know it's either raining or not. Similarly here, whether or not $A$ is true, we already know $A\eor \enot A$. We can check this with a truth table.
\begin{center}
	\begin{tabular}{c|deef}
		$A$&$A$&$\eor$&$\enot$&$A$\\\hline
		T&T&\TTbf{T}&F&T\\
		F&F&\TTbf{T}&T&F
	\end{tabular}
\end{center}
Since there is a $T$ under the main connective on every line of the truth table, $A\eor \enot A$ is true, whatever the valuation is, i.e.~whether $A$ is true or false.


\factoidbox{
	$\metaX$ is a \define{contradiction} iff it is false on every valuation.
}
\newglossaryentry{contradiction of TFL}
{
	name=contradiction (of TFL),
	text = contradiction,
	description={A sentence that is false on every \gls{valuation}}
}

We can determine whether a sentence is a contradiction just by using truth tables. If the sentence is false on every line of a complete truth table, then it is false on every valuation, so it is a contradiction.
%In the example of \S\ref{s:CompleteTruthTables}, `$[(C\eiff C) \eif C] \eand \enot(C \eif C)$' is a contradiction.

Our core example of a contradiction is $A\eand\enot A$. Whether $A$ is true or false, $A\eand\enot A$ is false. This can again be checked using truth tables, observing that there is an F under the main connective in each line of the truth table.

\begin{center}
	\begin{tabular}{c|deef}
		$A$&$A$&$\eand$&$\enot$&$A$\\\hline
		T&T&\TTbf{F}&F&T\\
		F&F&\TTbf{F}&T&F
	\end{tabular}
\end{center}


\factoidbox{
	$\metaX$ is \define{contingent} iff it is neither a tautology nor a contradiction.
}





\begin{highlighted}
		To check these notions on a truth table, see if the value recoreded underneath the main connective is always $T$, always $F$ or is mixed:
		
	%\smallskip 
	\begin{center}
		\begin{minipage}{.3\linewidth}
			\begin{center}
				\begin{tabular}{r | c}
					Valns & $\metaX$\\
					\hline
					$v_1$&T\\
					$v_2$  & T \\
					%			$v_3$&T\\
					\vdots & \vdots \\
					$v_n$ & T \\
				\end{tabular}\\ \textbf{logical truth /} \\
				\textbf{tautology}
			\end{center}
		\end{minipage}
		\begin{minipage}{.35\linewidth}
			\begin{center}
				\begin{tabular}{r | c}
					Valns & $\metaX$\\
					\hline
					$v_1$&F\\
					$v_2$  & F \\
					%			$v_3$&T\\
					\vdots & \vdots \\
					$v_n$ & F\\
				\end{tabular}\\ \textbf{contradiction}
			\end{center}
		\end{minipage}
		\begin{minipage}{.3\linewidth}
			\begin{center}
				\begin{tabular}{r | c}
					Valns & $\metaX$\\
					\hline
					$v_1$&\\
					$v_2$  & T \\
					%			$v_3$&T\\
					\vdots &  \\
					$v_n$ &F \\
				\end{tabular}\\ \textbf{contingent}
			\end{center}
		\end{minipage}
	\end{center}
	%	
	
	%\smallskip
	
	%	Otherwise \textbf{contingent}.
\end{highlighted}

\section{Logical equivalence}
Here is another useful notion:
\factoidbox{
	$\metaX$ and $\metaY$ are \define{logically equivalent} iff, on each valuation they are either both true or both false. 
}
\newglossaryentry{logically equivalent}
{
	name=logical equivalence (in TFL),
	text = logically equivalent,
	description={A property held by pairs of sentences if and only if the sentences have the same truth value on every valuation}
}
 Again, it is easy to test for logical equivalence using truth tables. Consider the sentences `$\enot(P \eor Q)$' and `$\enot P \eand \enot Q$'. Are they logically equivalent? To find out, we construct a truth table.
\begin{center}
	\begin{tabular}{c c|d e e f |d e e e f}
		$P$&$Q$&\enot&$(P$&\eor&$Q)$&\enot&$P$&\eand&\enot&$Q$\\
		\hline
		T & T & \TTbf{F} & T & T & T & F & T & \TTbf{F} & F & T\\
		T & F & \TTbf{F} & T & T & F & F & T & \TTbf{F} & T & F\\
		F & T & \TTbf{F} & F & T & T & T & F & \TTbf{F} & F & T\\
		F & F & \TTbf{T} & F & F & F & T & F & \TTbf{T} & T & F
	\end{tabular}
\end{center}
Look at the columns for the main logical operators; negation for the first sentence, conjunction for the second. On the first three rows, both are false. On the final row, both are true. Since they match on every row, the two sentences are logically equivalent.


One can show that `$(A \eand B) \eand C$' and  `$A \eand (B \eand C)$' are logically equivalent.


\begin{highlighted}
		To check for {logical equivalence} with truth tables, look for a row where one of them is true and the other false, for example:
	
	%\smallskip 
	\begin{center}
		%		\begin{minipage}{.5\linewidth}
			%			\begin{center}
				\begin{tabular}{r || c |c }
					Valns & $\metaX$ & $\metaY$ \\
					\hline
					$v_1$&&\\
					$\bigstar$ $v_2$  & T&F\\
					$v_3$&&\\
					$v_4$&&
				\end{tabular}\\ \textbf{Not logically equivalent}
				%			\end{center}
			%		\end{minipage}%
		%		\begin{minipage}{.5\linewidth}
			%			\begin{center}
				%			\begin{tabular}{r || c |c }
					%				Valns & $\metaX_1$ & $\metaX_2$ \\
					%				\hline
					%				$v_1$&&\\
					%				$\bigstar$ $v_2$  & T&F\\
					%				$v_3$&&\\
					%				$v_4$&&
					%			\end{tabular}\\ \textbf{}
				%		\end{center}
			%		\end{minipage}
	\end{center}
	%	
	
	%\smallskip
	
	%	If no such row exists, then the set is \textbf{inconsistent}.
\end{highlighted}


\begin{practiceproblems}
	\problempart
	Revisit your answers to \ref{pr.TT.TTorC}. Determine which sentences were tautologies, which were contradictions, and which were neither tautologies nor contradictions.
	\solutions
	
	\
	
	\problempart
	\label{pr.TT.consistent}
	Use truth tables to determine whether these sentences are jointly consistent, or jointly inconsistent:
	\begin{earg}
		\item $A\eif A$, $\enot A \eif \enot A$, $A\eand A$, $A\eor A$ %consistent
		\item $A\eor B$, $A\eif C$, $B\eif C$ %consistent
		\item $B\eand(C\eor A)$, $A\eif B$, $\enot(B\eor C)$  %inconsistent
		\item $A\eiff(B\eor C)$, $C\eif \enot A$, $A\eif \enot B$ %consistent
	\end{earg}
	
	
	
	\problempart Determine whether each sentence is a tautology, a contradiction, or a contingent sentence, using a complete truth table.
	\begin{earg}
		\item $\enot B \eand B$ \vspace{.5ex}%contra
		
		
		\item $\enot D \eor D$ \vspace{.5ex}%taut
		
		
		\item $(A\eand B) \eor (B\eand A)$\vspace{.5ex} %contingent
		
		
		\item $\enot[A \eif (B \eif A)]$\vspace{.5ex} %contra
		
		
		\item $A \eiff [A \eif (B \eand \enot B)]$ \vspace{.5ex}%contra
		
		
		\item $[(A \eand B) \eiff B] \eif (A \eif B)$ \vspace{.5ex}% contingent.
		
	\end{earg}
	
	
	
	\noindent\problempart
	\label{pr.TT.equiv}
	Determine whether each the following sentences are logically equivalent using complete truth tables. If the two sentences really are logically equivalent, write ``equivalent.'' Otherwise write, ``Not equivalent.''
	\begin{earg}
		\item $A$ and $\enot A$
		\item $A \eand \enot A$ and $\enot B \eiff B$
		\item $[(A \eor B) \eor C]$ and $[A \eor (B \eor C)]$
		\item $A \eor (B \eand C)$ and $(A \eor B) \eand (A \eor C)$
		\item $[A \eand (A \eor B)] \eif B$ and $A \eif B$\end{earg}
	
	
	\problempart
	\label{pr.TT.equiv2}
	Determine whether each the following sentences are logically equivalent using complete truth tables. If the two sentences really are equivalent, write ``equivalent.'' Otherwise write, ``not equivalent.''
	\begin{earg}
		\item $A\eif A$ and $A \eiff A$
		\item $\enot(A \eif B)$ and $\enot A \eif \enot B$
		\item $A \eor B$ and $\enot A \eif B$
		\item$(A \eif B) \eif C$ and $A \eif (B \eif C)$
		\item $A \eiff (B \eiff C)$ and $A \eand (B \eand C)$
	\end{earg}
	
	
	\problempart
	\label{pr.TT.consistent2}
	Determine whether each collection of sentences is jointly consistent or jointly inconsistent using a complete truth table.
	\begin{earg}
		\item $A \eand \enot B$, $\enot(A \eif B)$, $B \eif A$\vspace{.5ex} %Consistent
		
		%\begin{tabular}{ccccccccccccccc}
		%1. 	&	A 					 & \eand 		&  \enot & B & & \enot  		& 	 (A	  & 	 \eif	 	 & 	 B)		 & 	 & 	 B	 	 & 	\eif 	 	 & 	A 	 	 & 	 Consistent \\
		%\cline{2-5} \cline{7-10}\cline{12-14}
		%	& 	T 					 & 	 F	 		&  F	 & T & & F	 		& 	 T	  & 	 T	 	 & 	T 	 	 & 	 & 	 T	 	 & 	 T	 	 & T	 	 	&	  \\
		%\cline{2-14}
		%	& \multicolumn{1}{|r}{T}& 	\textbf{T}	 & T	 & F & & \textbf{T}	 & 	 T	 & 	 F	 	 & 	 F	 	 & 	 & 	 F	 	 & 	 \textbf{T}	 	 & 	 \multicolumn{1}{r|}{T}	 	 & 	  \\
		%\cline{2-14}
		%	& 	 F	 				 & 	 F	 & 	 F	 & T & 	& 	 F	 & 	 F	 & 	 T	 	 & 	 T	 	 & 	  & 	 T	 	 & 	 F	 	 & 	 F	 	 & 	  \\
		%	& 	 F	  				& 	 F	 & 	 T	 & 	F&  & 	 F	 & 	 F	 & 	 T	 	 & 	 F	 	 & 	  & 	 F	 	 & 	 T	 	 & 	 F	 	 & 	  \\
		%\end{tabular}
		
		\item $A \eor B$, $A \eif \enot A$, $B \eif \enot B$ \vspace{.5ex}%inconsistent.
		
		%\begin{tabular}{ccccccccccccccc}
		%2. &A	 & \eor 	 & B 	 & 	 	 & A 	 & \eif 	 & 	\enot & A 	 & 	 	 & B 	 & \eif 	 & \enot	 & 	B 	 & 	Inconsistent \\
		%\cline{2-4}\cline{6- 9} \cline{11-14}
		%   &	T	 & 	 T	 &T  	 & 	 	 & T	 & 	 F	 & 	F 	 & T 	 & 	 	 & 	T 	 & 	F 	 & 	 F	 & 	T 	 & 	 \\
		%   &	 T	& 	 T	 & F 	 & 	 	 & 	T 	 & 	 F	 & 	 F	 & 	 T	 & 	 	 & 	F 	 & 	 T	 & 	 T	 & 	 F	 & 	 \\
		%   &	 F	& 	 T	 & 	 T	 & 	 	 & 	F 	 & 	 T	 & 	 T	 & 	F 	 & 	 	 & 	 T	 & 	 F	 & 	 F	 & 	 T	 & 	 \\
		%   &	 F	& 	 F	 & 	 F	 & 	 	 & 	 F	 & 	 T	 & 	 T	 & 	 F	 & 	 	 & 	 F	 & 	 T	 & 	 T	 & 	 F	 & 	 \\
		%\end{tabular}
		
		\item $\enot(\enot A \eor B) $, $A \eif \enot C$, $A \eif (B \eif C)$\vspace{.5ex} %Inconsistent
		
		%3. &\enot & (\enot & A & \eor &B) &  &A  & \eif 	 &\enot 	 &C & 	 & A &\eif 	& (B 	 &\eif 	& C)	 &Consistent \\
		%\cline{2-6}\cline{8-11} \cline{13-17}
		%   &	F 	& 	F	 & 	T & T	 & T & 	  & T & F	 & 	 F&T 	 & 	 &T & T	 & T	 &T 	 &T 	 & \\
		%   &	 F	& 	F	 & 	T & T	 & T & 	  & T & T	 & 	 T& F	 & 	 &T & F	 & T	 & F	 &F 	 & \\
		%
		%  &	 T & 	F 	& 	T & F	 & F & 	  & T & F	 & 	 F& T	 & 	 &T & T	 & F	 & T	 &T 	 & \\
		%\cline{2-17}
		%   &	 \multicolumn{1}{|r}{{\color{red}T}}		&  F	 & 	T & F	 & 	F &  & 	T & {\color{red}T}	 & 	 T&F 	& 	 &T & {\color{red}T}	 & F	 & T	 &\multicolumn{1}{r|}{F} 	 & \\
		%\cline{2-17}
		%   &	 F	& 	T	 & 	F & T	 & 	T &  & 	F & T	 & 	 F& T	 & 	 &F	 & F	 & T	 & T	 &T 	 & \\
		%   &	 F	& 	 T	& 	F & T	 & 	T &  & 	F & T	 & 	T & F 	& 	 &F	 & T	 & T	 &F 	 &F 	 & \\
		%   &	 F	& 	 T	& 	F & T	 & 	F &  & 	F & T	 & 	F & T	 & 	 &F	 & T	 & F	 & T	 &T 	 & \\
		%   &	 F	& 	 T	& 	F & T	 & 	F &  & 	F & T	 & 	T & F	 & 	 &F	 & T	 & F	 & T	 &F 	 & \\
		%\end{tabular}
		%
		
		
		\item $A \eif B$, $A \eand \enot B$\vspace{.5ex} %Inconsistent
		
		\item $A \eif (B \eif C)$, $(A \eif B) \eif C$, $A \eif C$\vspace{.5ex} % consistent.
		
	\end{earg}
	
	\noindent\problempart
	\label{pr.TT.consistent3}
	Determine whether each collection of sentences is jointly consistent or jointly inconsistent, using a complete truth table.
	\begin{earg}
		\item $\enot B$, $A \eif B$, $A$ \vspace{.5ex}%inconsistent.
		\item $\enot(A \eor B)$, $A \eiff B$, $B \eif A$\vspace{.5ex} %Consistent
		\item $A \eor B$, $\enot B$, $\enot B \eif \enot A$\vspace{.5ex} %Inconsistent
		\item $A \eiff B$, $\enot B \eor \enot A$, $A \eif B$\vspace{.5ex} %consistent.
		\item $(A \eor B) \eor C$, $\enot A \eor \enot B$, $\enot C \eor \enot B$\vspace{.5ex} %consistent
	\end{earg}
	
	
	
	
	\solutions
	\problempart
	\label{pr.TT.concepts}
	Answer each of the questions below and justify your answer.
	\begin{earg}
		\item Suppose $\metaX$ is true and $\metaY$ is true\\Is $\metaX\eor\metaY$ true.\myanswer{\\true}
		\item Suppose $\metaX$ is false \\Is $\metaX\eor\metaY$ true?\myanswer{\\false. If $\metaY$ is also false, then $\metaX\eor\metaY$ is false. }
		\item Suppose $\metaX$ is false \\Is $\metaX\eif\metaY$ true. \myanswer{\\true}
		%\item Suppose $\entails \metaX$.\\ What can you say about $\metaX$?\myanswer{\\There are no valuations where the premises (in this case nothing) are true and the conclusion, \metaX, is false. So there are no valuations where the conclusion is false. I.e.~$\metaX$ is true on all valuations. It is a tautology.}
		%\item Suppose that \metaX and \metaY are logically equivalent.\\ What can you say about $\metaX\eiff\metaY$?
		%\myanswer{\item[] \metaX and \metaY have the same truth value on every line of a complete truth table, so $\metaX\eiff\metaY$ is true on every line. It is a tautology.}
		%\item Suppose $\metaX\entails\metaY$. \\Is $\metaX\therefore\metaY$ valid?\myanswer{\\Yes}
		\item Suppose that $\metaX$, $\metaY$ and $\metaZ$  are jointly inconsistent. \\What can you say about $\metaX\eand\metaY\eand\metaY$?
		\myanswer{\item[] Since the sentences are jointly inconsistent, there is no valuation on which they are all true. So their conjunction is false on every valuation. It is a contradiction}
		\item Suppose $\metaX$ is a tautology and  $\metaX\therefore\metaY$ is a valid argument. \\What can you say about $\metaY$?
		\item Suppose that $(\metaX\eand\metaY)\eif\metaZ$ is not a tautology.\\ What can you say about whether $\metaX, \metaY \therefore\metaZ$ is valid?
		\myanswer{\item[] Since the sentence $(\metaX\eand\metaY)\eif\metaZ$ is not a tautology, there is some line on which it is false. Since it is a conditional, on that line, \metaX and \metaY are true and \metaZ is false. So the argument is invalid.}
		\item Suppose that \metaX is a contradiction.\\ What can you say about whether $\metaX, \metaY \entails \metaZ$?
		\myanswer{\item[] Since \metaX is false on every line of a complete truth table, there is no line on which \metaX and \metaY are true and \metaZ is false. So the entailment holds.}
		%%\item Suppose that \metaZ is a tautology.\\ What can you say about whether $\metaX, \metaY\entails \metaZ$?
		%%\myanswer{\item[] Since \metaZ is true on every line of a complete truth table, there is no line on which \metaX and \metaY are true and \metaZ is false. So the entailment holds.}
		\item Suppose that \metaX and \metaY are logically equivalent.\\ What can you say about $(\metaX\eor\metaY)$?
		\myanswer{\item[] Not much. Since $\metaX$ and $\metaY$ are true on exactly the same lines of the truth table, their disjunction is true on exactly the same lines. So, their disjunction is logically equivalent to them.}
		\item Suppose that \metaX and \metaY are \emph{not} logically equivalent.\\ What can you say about $\metaX\eor\metaY$?
		\myanswer{\item[] \metaX and \metaY have different truth values on at least one line of a complete truth table, and $(\metaX\eor\metaY)$ will be true on that line. On other lines, it might be true or false. So $\metaX\eor\metaY$ is either a tautology or it is contingent; it is \emph{not} a contradiction.}
	\end{earg}
	
	\noindent\problempart
	\label{pr.TT.meta}
	Are the following statements true? Why?
	\begin{itemize}
	\item if $Y$ is a logical consequence of $X$, then $X\eif Y$ is a logical truth.\myanswer{Yes}
	\item if $X\eif Y$ is a logical truth, then $Y$ is a logical consequence of $X$.\myanswer{Yes}
	\item if $X\eif Y\eand\enot Y$ is a logical truth, then $X$ is a logical contradiction.\myanswer{Yes}
	\item if $X\eif Y\eand\enot Y$ is true relative to a valuation, then $X$ is a logical contradiction.\myanswer{No}
	\item if $X\eor\enot X\eif Y$ is a logical truth, then $Y$ is a tautology.\myanswer{Yes}
	\item if $X\eor\enot X\eif Y$ is true relative to a valuation, then $Y$ is a tautology.\myanswer{No}
	\end{itemize}
	
	\problempart
	Consider the following principle:
	\begin{ebullet}
		\item Suppose $\metaX$ and $\metaY$ are logically equivalent. Suppose an argument contains $\metaX$ (either as a premise, or as the conclusion). The validity of the argument would be unaffected, if we replaced $\metaX$ with $\metaY$.
	\end{ebullet}
	Is this principle correct? Explain your answer.
	
\end{practiceproblems}

\chapter{Truth table shortcuts}\teachingnote{Not taught in the course. You may use these techniques if you wish.\\}

Sometimes shortcuts can be taken when doing truth tables.
%With practice, you will quickly become adept at filling out truth tables.
In this section, we want to give you some permissible shortcuts to help you along the way.
These will never be {required} but they can speed things up.




\section{Working through truth tables}
You will quickly find that you do not need to copy the truth value of each atomic sentence, but can simply refer back to them. So you can speed things up by writing:
\begin{center}
	\begin{tabular}{c c|d e e e e f}
		$P$&$Q$&$(P$&\eor&$Q)$&\eiff&\enot&$P$\\
		\hline
		T & T &  & T &  & \TTbf{F} & F\\
		T & F &  & T &  & \TTbf{F} & F\\
		F & T &  & T & & \TTbf{T} & T\\
		F & F &  & F &  & \TTbf{F} & T
	\end{tabular}
\end{center}
You also know for sure that a disjunction is true whenever one of the disjuncts is true. So if you find a true disjunct, there is no need to work out the truth values of the other disjuncts. Thus you might offer:
\begin{center}
	\begin{tabular}{c c|d e e e e e e f}
		$P$&$Q$& $(\enot$ & $P$&\eor&\enot&$Q)$&\eor&\enot&$P$\\
		\hline
		T & T & F & & F & F& & \TTbf{F} & F\\
		T & F &  F & & T& T& &  \TTbf{T} & F\\
		F & T & & &  & & & \TTbf{T} & T\\
		F & F & & & & & &\TTbf{T} & T
	\end{tabular}
\end{center}
Equally, you know for sure that a conjunction is false whenever one of the conjuncts is false. So if you find a false conjunct, there is no need to work out the truth value of the other conjunct. Thus you might offer:
\begin{center}
	\begin{tabular}{c c|d e e e e e e f}
		$P$&$Q$&\enot &$(P$&\eand&\enot&$Q)$&\eand&\enot&$P$\\
		\hline
		T & T &  &  & &  & & \TTbf{F} & F\\
		T & F &   &  &&  & & \TTbf{F} & F\\
		F & T & T &  & F &  & & \TTbf{T} & T\\
		F & F & T &  & F & & & \TTbf{T} & T
	\end{tabular}
\end{center}
A similar short cut is available for conditionals. You immediately know that a conditional is true if either its consequent is true, or its antecedent is false. Thus you might present:
\begin{center}
	\begin{tabular}{c c|d e e e e e f}
		$P$&$Q$& $((P$&\eif&$Q$)&\eif&$P)$&\eif&$P$\\
		\hline
		T & T & &  & & & & \TTbf{T} & \\
		T & F &  &  & && & \TTbf{T} & \\
		F & T & & T & & F & & \TTbf{T} & \\
		F & F & & T & & F & &\TTbf{T} &
	\end{tabular}
\end{center}
So `$((P \eif Q) \eif P) \eif P$' is a tautology. In fact, it is an instance of \emph{Peirce's Law}, named after Charles Sanders Peirce.

\section{Testing for validity and logical consequence}
When we use truth tables to test for validity or logical consequence, we are checking for \emph{bad} lines: lines where the premises are all true and the conclusion is false. Note:
\begin{earg}
	\item[\textbullet] Any line where the conclusion is true is not a bad line.
	\item[\textbullet] Any line where some premise is false is not a bad line.
\end{earg}
Since \emph{all} we are doing is looking for bad lines, we should bear this in mind. So: if we find a line where the conclusion is true, we do not need to evaluate anything else on that line: that line definitely isn't bad. Likewise, if we find a line where some premise is false, we do not need to evaluate anything else on that line.

With this in mind, consider how we might test the following for validity:
$$\enot L \eif (J \eor L), \enot L \therefore J$$
The \emph{first} thing we should do is evaluate the conclusion. If we find that the conclusion is \emph{true} on some line, then that is not a bad line. So we can simply ignore the rest of the line. So at our first stage, we are left with something like:
\begin{center}
	\begin{tabular}{c c|d e e e e f |d f|c}
		$J$&$L$&\enot&$L$&\eif&$(J$&\eor&$L)$&\enot&$L$&$J$\\
		\hline
		%J   L   -   L      ->     (J   v   L)
		T & T & &&&&&&&& {T}\\
		T & F & &&&&&&&& {T}\\
		F & T & &&?&&&&?&& {F}\\
		F & F & &&?&&&&?&& {F}
	\end{tabular}
\end{center}
where the blanks indicate that we are not going to bother doing any more investigation (since the line is not bad) and the question-marks indicate that we need to keep investigating.

The easiest premise to evaluate is the second, so we next do that:
\begin{center}
	\begin{tabular}{c c|d e e e e f |d f|c}
		$J$&$L$&\enot&$L$&\eif&$(J$&\eor&$L)$&\enot&$L$&$J$\\
		\hline
		%J   L   -   L      ->     (J   v   L)
		T & T & &&&&&&&& {T}\\
		T & F & &&&&&&&& {T}\\
		F & T & &&&&&&{F}&& {F}\\
		F & F & &&?&&&&{T}&& {F}
	\end{tabular}
\end{center}
Note that we no longer need to consider the third line on the table: it will not be a bad line, because (at least) one of premises is false on that line. Finally, we complete the truth table:
\begin{center}
	\begin{tabular}{c c|d e e e e f |d f|c}
		$J$&$L$&\enot&$L$&\eif&$(J$&\eor&$L)$&\enot&$L$&$J$\\
		\hline
		%J   L   -   L      ->     (J   v   L)
		T & T & &&&&&&&& {T}\\
		T & F & &&&&&&&& {T}\\
		F & T & &&&&&&{F}& & {F}\\
		F & F & T &  & \TTbf{F} &  & F & & {T} & & {F}
	\end{tabular}
\end{center}
The truth table has no bad lines, so the argument is valid. (Any valuation on which all the premises are true is a valuation on which the conclusion is true.)

It might be worth illustrating the tactic again. Let us check whether the following argument is valid
$$A\eor B, \enot (A\eand C), \enot (B \eand \enot D) \therefore (\enot C\eor D)$$
At the first stage, we determine the truth value of the conclusion. Since this is a disjunction, it is true whenever either disjunct is true, so we can speed things along a bit. We can then ignore every line apart from the few lines where the conclusion is false.
\begin{center}
	\begin{tabular}[t]{c c c c | c|c|c|d e e f }
		$A$ & $B$ & $C$ & $D$ & $A\eor B$ & $\enot (A\eand C)$ & $\enot (B\eand \enot D)$ & $(\enot$ &$C$& $\eor$ & $D)$\\
		\hline
		T & T & T & T & & & & &  &  \TTbf{T} & \\
		T & T & T & F & ? & ? & ? & F & &  \TTbf{F} & \\
		T & T & F & T &  & &   & & &  \TTbf{T} & \\
		T & T & F & F &  &  &   & T & &  \TTbf{T} &\\
		T & F & T & T &  &  &  & & &  \TTbf{T} & \\
		T & F & T & F & ? & ? & ?  & F &  &  \TTbf{F} &\\
		T & F & F & T & & & & & & \TTbf{T} &\\
		T & F & F & F & & & & T &  & \TTbf{T} & \\
		F & T & T & T & & & & & & \TTbf{T} & \\
		F & T & T & F & ? & ? & ? & F &  & \TTbf{F} &\\
		F & T & F & T & & &  & & & \TTbf{T} & \\
		F & T & F & F & & & &T & & \TTbf{T} & \\
		F & F & T & T & & & & & & \TTbf{T} & \\
		F & F & T & F & ? & ? & ? & F & & \TTbf{F} & \\
		F & F & F & T & & & & & & \TTbf{T} & \\
		F & F & F & F & & & & T& & \TTbf{T} & \\
	\end{tabular}
\end{center}
We must now evaluate the premises. We use shortcuts where we can:
\begin{center}
	\begin{tabular}[t]{c c c c | d e f |d e e f |d e e e f |d e e f }
		$A$ & $B$ & $C$ & $D$ & $A$ & $\eor$ & $B$ & $\enot$ & $(A$ &$\eand$ &$ C)$ & $\enot$ & $(B$ & $\eand$ & $\enot$ & $D)$ & $(\enot$ &$C$& $\eor$ & $D)$\\
		\hline
		T & T & T & T & & && & && & && & & & &  &  \TTbf{T} & \\
		T & T & T & F & &\TTbf{T}& & \TTbf{F}& &T& & & & & & & F & &  \TTbf{F} & \\
		T & T & F & T & & && & && & &&  & &   & & &  \TTbf{T} & \\
		T & T & F & F & & && & && & &&  &  &   & T & &  \TTbf{T} & \\
		T & F & T & T & & && & && & &&  &  &  & & &  \TTbf{T} & \\
		T & F & T & F & &\TTbf{T}& &\TTbf{F}& &T& &  && & & & F & & \TTbf{F} & \\
		T & F & F & T & & && & && & && & & & & & \TTbf{T} & \\
		T & F & F & F & & && & && & && & & & T &  & \TTbf{T} & \\
		F & T & T & T& & && & && & & & & & & & & \TTbf{T} & \\
		F & T & T & F & &\TTbf{T}& & \TTbf{T}& & F& & \TTbf{F}& & T& T&  & F &  & \TTbf{F} & \\
		F & T & F & T & & && & && & && & &  & & & \TTbf{T} & \\
		F & T & F & F& & && & && & && & & &T & & \TTbf{T} & \\
		F & F & T & T & & && & && & && & & & & & \TTbf{T} & \\
		F & F & T & F & & \TTbf{F} & & & & & & &&  &  &  & F & & \TTbf{F} & \\
		F & F & F & T & & && & && & && & & & & & \TTbf{T} & \\
		F & F & F & F & & && & && & && & & & T& & \TTbf{T} & \\
	\end{tabular}
\end{center}
If we had used no shortcuts, we would have had to write 256 `T's or `F's on this table. Using shortcuts, we only had to write 37. We have saved ourselves a \emph{lot} of work.

We have been discussing shortcuts in testing for logically validity, but exactly the same shortcuts can be used in testing for logical consequence. By employing a similar notion of \emph{bad} lines, you can save yourself a huge amount of work.

\begin{practiceproblems}
	\problempart
	\label{pr.TT.TTorC2}
	Using shortcuts, determine whether each sentence is a tautology, a contradiction, or neither.
	\begin{earg}
		
		\item $\enot B \eand B$ %contra
		\myanswer{ \hfill Contradiction\begin{center}
				\begin{tabular}{c | d e e f }
					$B$ & $\enot$&$B$&$\eand$&$B$\\
					\hline
					T & F & & \TTbf{F}\\
					F & & & \TTbf{F} & 
				\end{tabular}
		\end{center}}
		\item $\enot D \eor D$ %taut
		\myanswer{\hfill Tautology
			\begin{center}
				\begin{tabular}{c | d e e f }
					$D$ & $\enot$&$D$&$\eor$&$D$\\
					\hline
					T &  & & \TTbf{T}\\
					F & T & & \TTbf{T}
				\end{tabular}
		\end{center}}
		\item $(A\eand B) \eor (B\eand A)$ %contingent
		\myanswer{\hfill Neither
			\begin{center}
				\begin{tabular}{c c | d e e e e e f }
					$A$ & $B$ & $(A$&$\eand$&$B)$&$\eor$&$(B$&$\eand$&$A)$\\
					\hline
					T & T & & T & & \TTbf{T}\\
					T & F & & F & & \TTbf{F} & & F \\
					F & T & & F & & \TTbf{F} & & F \\
					F & F & & F & & \TTbf{F} & & F 
				\end{tabular}
		\end{center}}
		\item $\enot[A \eif (B \eif A)]$ %contra
		\myanswer{\hfill Contradiction
			\begin{center}
				\begin{tabular}{c c | d e e e e f }
					$A$ & $B$ & $\enot[$&$A$&$\eif$&$(B$&$\eif$&$A)]$\\
					\hline
					T & T & \TTbf{F} &  & T & & T\\
					T & F & \TTbf{F} &  & T & & T \\
					F & T & \TTbf{F} &  & T & &  \\
					F & F & \TTbf{F} &  & T & &  
				\end{tabular}
		\end{center}}
		\item $A \eiff [A \eif (B \eand \enot B)]$ %contra
		\myanswer{\hfill Contradiction
			\begin{center}
				\begin{tabular}{c c | d e e e e e e f }
					$A$ & $B$ & $A$&$\eiff$&$[A$&$\eif$&$(B$&$\eand$&$\enot$&$B)]$\\
					\hline
					T & T & & \TTbf{F} &  & F & & F& F\\
					T & F & & \TTbf{F} &  & F & & F \\
					F & T & & \TTbf{F} &  & T & &  \\
					F & F & & \TTbf{F} &  & T & &  
				\end{tabular}
		\end{center}}
		\item $\enot(A\eand B) \eiff A$ %contingent
		\myanswer{\hfill Neither
			\begin{center}
				\begin{tabular}{c c | d e e e e f }
					$A$ & $B$ & $\enot$&$(A$&$\eand$&$B)$&$\eiff$&$A$\\
					\hline
					T & T & F & & T & & \TTbf{F} \\
					T & F & T & & F & & \TTbf{T} \\
					F & T & T & & F & & \TTbf{F} \\
					F & F & T & & F & & \TTbf{F} \\
				\end{tabular}
		\end{center}}
		\item $A\eif(B\eor C)$ %contingent
		\myanswer{\hfill Neither
			\begin{center}
				\begin{tabular}{c c c | d e e e f }
					$A$ & $B$ & $C$ & $A$&$\eif$&$(B$&$\eor$&$C)$\\
					\hline
					T & T & T & & \TTbf{T} & & T \\
					T & T & F & & \TTbf{T} & & T \\
					T & F & T & & \TTbf{T} & & T \\
					T & F & F & & \TTbf{F} & & F \\
					F & T & T & & \TTbf{T} & &  \\
					F & T & F & & \TTbf{T} & &  \\
					F & F & T & & \TTbf{T} & &  \\
					F & F & F & & \TTbf{T} & &  
				\end{tabular}
		\end{center}}
		\item $(A \eand\enot A) \eif (B \eor C)$ %tautology
		\myanswer{\hfill Tautology
			\begin{center}
				\begin{tabular}{c c c | d e e e e e e f }
					$A$ & $B$ & $C$ & $(A$&$\eand$&$\enot$&$A)$&$\eif$&$(B$&$\eor$&$C)$\\
					\hline
					T & T & T & & \TTbf{F} & F& & T \\
					T & T & F & & \TTbf{F} &F & & T \\
					T & F & T & & \TTbf{F} &F & & T \\
					T & F & F & & \TTbf{F} & F& & T \\
					F & T & T & & \TTbf{F} & & & T \\
					F & T & F & & \TTbf{F} & & & T \\
					F & F & T & & \TTbf{F} & & & T \\
					F & F & F & & \TTbf{F} & &  & T \\ 
				\end{tabular}
		\end{center}}
		\item $(B\eand D) \eiff [A \eiff(A \eor C)]$%contingent
		\myanswer{\hfill Neither
			\begin{center}
				\begin{tabular}{c c c c | d e e e e e e e f }
					$A$ & $B$ & $C$ & $D$ & $(B$&$\eand$&$D)$&$\eiff $&$[A$&$\eiff$&$(A$&$\eor$&$C)]$\\
					\hline
					T & T & T & T & & T & & \TTbf{T} &  & T & &  T\\
					T & T & T & F & & F & & \TTbf{F} &  & T & &  T\\
					T & T & F & T & & T & & \TTbf{T} &  & T & &  T\\
					T & T & F & F & & F & & \TTbf{F} &  & T & &  T\\
					T & F & T & T & & F & & \TTbf{F} &  & T & &  T\\
					T & F & T & F & & F & & \TTbf{F} &  & T & &  T\\
					T & F & F & T & & F & & \TTbf{F} &  & T & &  T\\
					T & F & F & F & & F & & \TTbf{F} &  & T & &  T\\
					F & T & T & T & & T & & \TTbf{F} &  & F & &  T\\
					F & T & T & F & & F & & \TTbf{T} &  & F & &  T\\
					F & T & F & T & & T & & \TTbf{T} &  & T & &  F\\
					F & T & F & F & & F & & \TTbf{F} &  & T & &  F\\
					F & F & T & T & & F & & \TTbf{T} &  & F & &  T\\
					F & F & T & F & & F & & \TTbf{T} &  & F & &  T\\
					F & F & F & T & & F & & \TTbf{F} &  & T & &  F\\
					F & F & F & F & & F & & \TTbf{F} &  & T & &  F
				\end{tabular}
		\end{center}}
	\end{earg}
\end{practiceproblems}

\chapter{Partial truth tables}\label{s:PartialTruthTable}\teachingnote{Not taught in course. This knowledge will not be needed for the exam. (Although you can use these techniques in the exam if you wish.)}

Sometimes, we do not need to know what happens on every line of a truth table. Sometimes, just a line or two will do.

\paragraph{Tautology.}
In order to show that a sentence is a tautology, we need to show that it is true on every valuation. That is to say, we need to know that it comes out true on every line of the truth table. So we need a complete truth table.

To show that a sentence is \emph{not} a tautology, however, we only need one line: a line on which the sentence is false. Therefore, in order to show that some sentence is not a tautology, it is enough to provide a single valuation---a single line of the truth table---which makes the sentence false.

Suppose that we want to show that the sentence `$(U \eand T) \eif (S \eand W)$' is \emph{not} a tautology. We set up a \define{partial truth table}:
\begin{center}
	\begin{tabular}{c c c c |d e e e e e f}
		$S$&$T$&$U$&$W$&$(U$&\eand&$T)$&\eif    &$(S$&\eand&$W)$\\
		\hline
		&   &   &   &    &   &    &\TTbf{F}&    &   &
	\end{tabular}
\end{center}
We have only left space for one line, rather than 16, since we are only looking for one line on which the sentence is false. For just that reason, we have filled in `F' for the entire sentence.

The main logical operator of the sentence is a conditional. In order for the conditional to be false, the antecedent must be true and the consequent must be false. So we fill these in on the table:
\begin{center}
	\begin{tabular}{c c c c |d e e e e e f}
		$S$&$T$&$U$&$W$&$(U$&\eand&$T)$&\eif    &$(S$&\eand&$W)$\\
		\hline
		&   &   &   &    &  T  &    &\TTbf{F}&    &   F &
	\end{tabular}
\end{center}
In order for the `$(U\eand T)$' to be true, both `$U$' and `$T$' must be true.
\begin{center}
	\begin{tabular}{c c c c|d e e e e e f}
		$S$&$T$&$U$&$W$&$(U$&\eand&$T)$&\eif    &$(S$&\eand&$W)$\\
		\hline
		& T & T &   &  T &  T  & T  &\TTbf{F}&    &   F &
	\end{tabular}
\end{center}
Now we just need to make `$(S\eand W)$' false. To do this, we need to make at least one of `$S$' and `$W$' false. We can make both `$S$' and `$W$' false if we want. All that matters is that the whole sentence turns out false on this line. Making an arbitrary decision, we finish the table in this way:
\begin{center}
	\begin{tabular}{c c c c|d e e e e e f}
		$S$&$T$&$U$&$W$&$(U$&\eand&$T)$&\eif    &$(S$&\eand&$W)$\\
		\hline
		F & T & T & F &  T &  T  & T  &\TTbf{F}&  F &   F & F
	\end{tabular}
\end{center}
We now have a partial truth table, which shows that `$(U \eand T) \eif (S \eand W)$' is not a tautology. Put otherwise, we have shown that there is a valuation which makes `$(U \eand T) \eif (S \eand W)$' false, namely, the valuation which makes `$S$' false, `$T$' true, `$U$' true and `$W$' false.

\paragraph{Contradiction.}
Showing that something is a contradiction requires a complete truth table: we need to show that there is no valuation which makes the sentence true; that is, we need to show that the sentence is false on every line of the truth table.

However, to show that something is \emph{not} a contradiction, all we need to do is find a valuation which makes the sentence true, and a single line of a truth table will suffice. We can illustrate this with the same example.
\begin{center}
	\begin{tabular}{c c c c|d e e e e e f}
		$S$&$T$&$U$&$W$&$(U$&\eand&$T)$&\eif    &$(S$&\eand&$W)$\\
		\hline
		&  &  &  &   &   &   &\TTbf{T}&  &  &
	\end{tabular}
\end{center}
To make the sentence true, it will suffice to ensure that the antecedent is false. Since the antecedent is a conjunction, we can just make one of them false. For no particular reason, we choose to make `$U$' false; and then we can assign whatever truth value we like to the other atomic sentences.
\begin{center}
	\begin{tabular}{c c c c|d e e e e e f}
		$S$&$T$&$U$&$W$&$(U$&\eand&$T)$&\eif    &$(S$&\eand&$W)$\\
		\hline
		F & T & F & F &  F &  F  & T  &\TTbf{T}&  F &   F & F
	\end{tabular}
\end{center}

\paragraph{Truth functional equivalence.}
To show that two sentences are logically equivalent, we must show that the sentences have the same truth value on every valuation. So this requires a  complete truth table.

To show that two sentences are \emph{not} logically equivalent, we only need to show that there is a valuation on which they have different truth values. So this requires only a one-line partial truth table: make the table so that one sentence is true and the other false.

\paragraph{Consistency.}
To show that some sentences are jointly consistent, we must show that there is a valuation which makes all of the sentences true,so this requires only a partial truth table with a single line.

To show that some sentences are jointly inconsistent, we must show that there is no valuation which makes all of the sentence true. So this requires a complete truth table: You must show that on every row of the table at least one of the sentences is false.

\paragraph{Validity.}
To show that an argument is valid, we must show that there is no valuation which makes all of the premises true and the conclusion false. So this  requires a complete truth table.  (Likewise for logical consequence.)

To show that argument is \emph{invalid}, we must show that there is a valuation which makes all of the premises true and the conclusion false. So this requires only a one-line partial truth table on which all of the premises are true and the conclusion is false. (Likewise for a failure of logical consequence.)


\
\\This table summarises what is required:

\begin{center}
	\begin{tabular}{l l l}
		%\cline{2-3}
		& \textbf{Yes} & \textbf{No}\\
		\hline
		%\cline{2-3}
		tautology? & complete truth table & one-line partial truth table\\
		contradiction? &  complete truth table  & one-line partial truth table\\
		%contingent? & two-line partial truth table & complete truth table\\
		equivalent? & complete truth table & one-line partial truth table\\
		consistent? & one-line partial truth table & complete truth table\\
		valid? & complete truth table & one-line partial truth table\\
		logical consequence? & complete truth table & one-line partial truth table\\
	\end{tabular}
\end{center}
\label{table.CompleteVsPartial}

%
%\begin{practiceproblems}
%	\problempart
%	\label{pr.TT.equiv3}
%	Use complete or partial truth tables (as appropriate) to determine whether these pairs of sentences are logically equivalent:
%	\begin{earg}
%		\item $A$, $\enot A$
%		\myanswer{\hfill Not logically equivalent
%			\begin{center}
%				\begin{tabular}{c | c | c}
%					$A$ &$A$&$\enot A$\\
%					\hline
%					T & T & F
%				\end{tabular}
%		\end{center}}
%		\item $A$, $A \eor A$
%		\myanswer{\hfill Logically equivalent
%			\begin{center}
%				\begin{tabular}{c | c | c}
%					$A$ &$A$&$A \eor A$\\
%					\hline
%					T & T & T\\
%					T & T & T
%				\end{tabular}
%		\end{center}}
%		\item $A\eif A$, $A \eiff A$ %Yes
%		\myanswer{\hfill Logically equivalent
%			\begin{center}
%				\begin{tabular}{c | c | c}
%					$A$ &$A \eif A$&$A \eiff A$\\
%					\hline
%					T & T & T\\
%					F & T & T
%				\end{tabular}
%		\end{center}}
%		\item $A \eor \enot B$, $A\eif B$ %No
%		\myanswer{\hfill Not logically equivalent
%			\begin{center}
%				\begin{tabular}{c c | c | c}
%					$A$ & $B$ &$A\eor \enot B$&$A \eif B$\\
%					\hline
%					T & F & {T} & F
%				\end{tabular}
%		\end{center}}
%		\item $A \eand \enot A$, $\enot B \eiff B$ %Yes
%		\myanswer{\hfill Logically equivalent
%			\begin{center}
%				\begin{tabular}{c c | d e e f | d e e f}
%					$A$ & $B$ &$A$&$\eand$&$\enot$&$A$&$\enot$&$B$&$\eiff$&$B$\\
%					\hline
%					T & T & & \TTbf{F} & F & & F & & \TTbf{F} & \\
%					T & F & & \TTbf{F} & F & & T & & \TTbf{F} & \\
%					F & T & & \TTbf{F} &  & & F & & \TTbf{F} & \\
%					F & F & & \TTbf{F} &  & &  T & & \TTbf{F} & \\
%				\end{tabular}
%		\end{center}}
%		
%		\item $\enot(A \eand B)$, $\enot A \eor \enot B$ %Yes
%		\myanswer{\hfill Logically equivalent
%			\begin{center}
%				\begin{tabular}{c c | d e e f | d e e e f}
%					$A$ & $B$ &$\enot$&$(A$&$\eand$&$B)$&$\enot$&$A$&$\eor$&$\enot$&$B$\\
%					\hline
%					T & T &\TTbf{F}& & T & & F & & \TTbf{F} & F\\
%					T & F & \TTbf{T} & & F & & F & & \TTbf{T} & T\\
%					F & T & \TTbf{T} & & F & & T & & \TTbf{T} & F\\
%					F & F & \TTbf{T} & & F & &  T& & \TTbf{T} & T\\
%				\end{tabular}
%		\end{center}}
%		\item $\enot(A \eif B)$, $\enot A \eif \enot B$ %No
%		\myanswer{\hfill Not logically equivalent
%			\begin{center}
%				\begin{tabular}{c c | d e e f | d e e e f}
%					$A$ & $B$ &$\enot$&$(A$&$\eif$&$B)$&$\enot$&$A$&$\eif$&$\enot$&$B$\\
%					\hline
%					T & T &\TTbf{F}& & T & & F & & \TTbf{T} & F\\
%				\end{tabular}
%		\end{center}}
%		\item $(A \eif B)$, $(\enot B \eif \enot A)$ %Yes
%		\myanswer{\hfill Logically equivalent
%			\begin{center}
%				\begin{tabular}{c c | c | d e e e f}
%					$A$ & $B$ &$(A \eif B)$&$(\enot$&$B$&$\eif$&$\enot$&$A)$\\
%					\hline
%					T & T &T& F & & \TTbf{T} &  \\
%					T & F &F & T & & \TTbf{F} & F  \\
%					F & T &T& F & & \TTbf{T} &  \\
%					F & F &T& T & & \TTbf{T} & T 
%				\end{tabular}
%		\end{center}}
%	\end{earg}
%	
%	\solutions
%	\problempart
%	\label{pr.TT.satisfiable4}
%	Use complete or partial truth tables (as appropriate) to determine whether these sentences are jointly satisfiable, or jointly unsatisfiable:
%	\begin{earg}
%		\item $A \eand B$, $C\eif \enot B$, $C$ %unsatisfiable
%		\myanswer{\hfill Jointly unsatisfiable
%			\begin{center}
%				\begin{tabular}{c c c | c | d e f | c c c c  }
%					$A$ & $B$ & $C$ & $A \eand B$ & $C$&$\eif$&$\enot B$&$C$\\
%					\hline
%					T & T & T & T & & \TTbf{F} & F& T \\
%					T & T & F & T & & \TTbf{T} & & F \\
%					T & F & T & F & & \TTbf{T} &T & T \\
%					T & F & F & F & & \TTbf{T} & & F \\
%					F & T & T & F & & \TTbf{F} &F & T \\
%					F & T & F & F & & \TTbf{T} & & F \\
%					F & F & T & F & & \TTbf{T} & T& T \\
%					F & F & F & F & & \TTbf{T} & &  F \\ 
%				\end{tabular}
%		\end{center}}
%		\item $A\eif B$, $B\eif C$, $A$, $\enot C$ %unsatisfiable
%		\myanswer{\hfill Jointly unsatisfiable
%			\begin{center}
%				\begin{tabular}{c c c | c | c | c | c }
%					$A$ & $B$ & $C$ & $A \eif B$ & $B \eif C$ & $A$ & $\enot C$\\
%					\hline
%					T & T & T & T & T & T & F\\
%					T & T & F & T & F & T & T\\
%					T & F & T & F & T & T & F\\
%					T & F & F & F & T & T & T\\
%					F & T & T & T & T & F & F\\
%					F & T & F & T & F & F & T\\
%					F & F & T & T & T & F & F\\
%					F & F & F & T & T & F & T
%				\end{tabular}
%		\end{center}}
%		\item $A \eor B$, $B\eor C$, $C\eif \enot A$ %satisfiable
%		\myanswer{\hfill Jointly satisfiable
%			\begin{center}
%				\begin{tabular}{c c c | c | c | d e f }
%					$A$ & $B$ & $C$ & $A \eor B$ & $B \eor C$ & $C$&$\eif$& $\enot A$\\
%					\hline
%					F & T & T & T & T & & \TTbf{T} & T\\
%				\end{tabular}
%		\end{center}}
%		\item $A$, $B$, $C$, $\enot D$, $\enot E$, $F$ %satisfiable
%		\myanswer{\hfill Jointly satisfiable
%			\begin{center}
%				\begin{tabular}{c c c c c c| c | c | c | c |c | c }
%					$A$ & $B$ & $C$ & $D$ & $E$ & $F$ & $A$ & $B$ & $C$ & $\enot D$ & $\enot E$ & $F$\\
%					\hline
%					T & T & T & F & F & T & T & T& T& T& T& T
%				\end{tabular}
%		\end{center}}
%		% solutions to last two problems omitted
%	\end{earg}
%	
%	\solutions
%	\problempart
%	\label{pr.TT.valid4}
%	Use complete or partial truth tables (as appropriate) to determine whether each argument is valid or invalid:
%	\begin{earg}
%		\item $A\eor\bigl[A\eif(A\eiff A)\bigr] \therefore A$ %invalid
%		\myanswer{\hfill Invalid
%			\begin{center}
%				\begin{tabular}{c |  d e e e e e f | c}
%					$A$ & $A$&$\eor$&$\bigl[A$&$\eif$&$(A$&$\eiff$&$A)\bigr]$ & $A$\\
%					\hline
%					F & & \TTbf{T} & & T & & & & F
%				\end{tabular}
%		\end{center}}
%		\item $A\eiff\enot(B\eiff A) \therefore A$ %invalid
%		\myanswer{\hfill Invalid
%			\begin{center}
%				\begin{tabular}{c c | d e e f | c}
%					$A$&$B$ & $A$&$\eiff$&$\enot$&$(B\eiff A)$ & $A$\\
%					\hline
%					F & F & & \TTbf{T} & F & T & F
%				\end{tabular}
%		\end{center}}
%		
%		\item $A\eif B, B \therefore A$ %invalid
%		\myanswer{\hfill Invalid
%			\begin{center}
%				\begin{tabular}{c c | c | c | c}
%					$A$&$B$ & $A \eif B$ & $B$ & $A$\\
%					\hline
%					F & T & T & T & F
%				\end{tabular}
%		\end{center}}
%		
%		\item $A\eor B, B\eor C, \enot B \therefore A \eand C$ %valid
%		\myanswer{\hfill Valid
%			\begin{center}
%				\begin{tabular}{c c c | c | c | c | c }
%					$A$ & $B$ & $C$ & $A \eor B$ & $B \eor C$ & $\enot B$ & $A \eand C$\\
%					\hline
%					T & T & T &  &  &  & T\\
%					T & T & F &  &  & F & F\\
%					T & F & T &  &  &  & T\\
%					T & F & F & T & F & T & F\\
%					F & T & T &  &  & F & F\\
%					F & T & F &  &  & F & F\\
%					F & F & T & F &  & T & F\\
%					F & F & F & F &  & T & F
%				\end{tabular}
%		\end{center}}
%		
%		\item $A\eiff B, B\eiff C \therefore A\eiff C$ %valid
%		\myanswer{\hfill Valid
%			\begin{center}
%				\begin{tabular}{c c c | c | c | c }
%					$A$ & $B$ & $C$ & $A \eiff B$ & $B \eiff C$ & $A \eiff C$\\
%					\hline
%					T & T & T &  &  & T\\
%					T & T & F & T & F & F\\
%					T & F & T &  &  & T\\
%					T & F & F & F &  & F\\
%					F & T & T & F &  & F\\
%					F & T & F &  &  & T\\
%					F & F & T & T & F & F\\
%					F & F & F &  &  & T
%				\end{tabular}
%		\end{center}}
%	\end{earg}
%	
%	\problempart
%	\label{pr.TT.TTorC3}
%	Determine whether each sentence is a tautology, a contradiction, or a contingent sentence. Justify your answer with a complete or partial truth table where appropriate.
%	
%	% truth tables in LaTeX generated by http://www.curtisbright.com/logic/. Be sure to give him a shout out.
%	
%	\begin{earg}
%		\item  $A \eif \enot A$ 
%		\myanswer{\hfill Contingent
%			\[
%			\begin{array}{c|cccc}
%				A&A&\eif&\enot&A\\\hline
%				T&T&\mathbf{F}&F&T\\
%				F&F&\mathbf{T}&T&F
%			\end{array}
%			\]
%		}
%		%	T letter, 2 connectives
%		\item $A \eif (A \eand (A \eor B))$ 
%		\myanswer{ \hfill Tautology
%			\[
%			\begin{array}{cc|ccc@{}ccc@{}ccc@{}c@{}c}
%				A&B&A&\eif&(&A&\eand&(&A&\eor&B&)&)\\\hline
%				T&T&T&\mathbf{T}&&T&T&&T&T&T&&\\
%				T&F&T&\mathbf{T}&&T&T&&T&T&F&&\\
%				F&T&F&\mathbf{T}&&F&F&&F&T&T&&\\
%				F&F&F&\mathbf{T}&&F&F&&F&F&F&&
%			\end{array}
%			\]
%			\vspace{6pt}
%		}
%		%			2 letters, 3 connectives
%		
%		\item $(A \eif B) \eiff (B \eif A)$
%		\myanswer{ \hfill Contingent
%			\[
%			\begin{array}{cc|c@{}ccc@{}ccc@{}ccc@{}c}
%				A&B&(&A&\rightarrow&B&)&\leftrightarrow&(&B&\rightarrow&A&)\\\hline
%				T&T&&T&T&T&&\mathbf{T}&&T&T&T&\\
%				T&F&&T&F&F&&\mathbf{F}&&F&T&T&\\
%				F&T&&F&T&T&&\mathbf{F}&&T&F&F&\\
%				F&F&&F&T&F&&\mathbf{T}&&F&T&F&
%			\end{array}
%			\]}
%		%		2 letters, 3 connectives
%		
%		\item $A \eif \enot(A \eand (A \eor B)) $
%		\myanswer{ \hfill Contingent
%			\[
%			\begin{array}{cc|cccc@{}ccc@{}ccc@{}c@{}c}
%				A&B&A&\rightarrow&\enot&(&A&\eand&(&A&\eor&B&)&)\\\hline
%				T&T&T&\mathbf{F}&F&&T&T&&T&T&T&&\\
%				T&F&T&\mathbf{F}&F&&T&T&&T&T&F&&\\
%				F&T&F&\mathbf{T}&T&&F&F&&F&T&T&&\\
%				F&F&F&\mathbf{T}&T&&F&F&&F&F&F&&
%			\end{array}
%			\]
%		}
%		
%		% 2 letters, 4 connectives
%		
%		\item $\enot B \eif [(\enot A \eand A) \eor B]$
%		\myanswer{ \hfill Contingent
%			\[
%			\begin{array}{cc|cccc@{}c@{}cccc@{}ccc@{}c}
%				A&B&\enot&B&\rightarrow&(&(&\enot&A&\eand&A&)&\eor&B&)\\\hline
%				T&T&F&T&\mathbf{T}&&&F&T&F&T&&T&T&\\
%				T&F&T&F&\mathbf{F}&&&F&T&F&T&&F&F&\\
%				F&T&F&T&\mathbf{T}&&&T&F&F&F&&T&T&\\
%				F&F&T&F&\mathbf{F}&&&T&F&F&F&&F&F&
%			\end{array}
%			\]}
%		%	2 letters, 5 connectives
%		
%		\item $\enot(A \eor B) \eiff (\enot A \eand \enot B)$
%		\myanswer{ \hfill Tautology
%			\[
%			\begin{array}{cc|cc@{}ccc@{}ccc@{}ccccc@{}c}
%				A&B&\enot&(&A&\eor&B&)&\leftrightarrow&(&\enot&A&\eand&\enot&B&)\\\hline
%				T&T&F&&T&T&T&&\mathbf{T}&&F&T&F&F&T&\\
%				T&F&F&&T&T&F&&\mathbf{T}&&F&T&F&T&F&\\
%				F&T&F&&F&T&T&&\mathbf{T}&&T&F&F&F&T&\\
%				F&F&T&&F&F&F&&\mathbf{T}&&T&F&T&T&F&
%			\end{array}
%			\]}
%		%2 letters, 6 connectives
%		
%		\item $[(A \eand B) \eand C] \eif B$
%		\myanswer{ \hfill Tautology
%			\[
%			\begin{array}{ccc|c@{}c@{}ccc@{}ccc@{}ccc}
%				A&B&C&(&(&A&\eand&B&)&\eand&C&)&\rightarrow&B\\\hline
%				T&T&T&&&T&T&T&&T&T&&\mathbf{T}&T\\
%				T&T&F&&&T&T&T&&F&F&&\mathbf{T}&T\\
%				T&F&T&&&T&F&F&&F&T&&\mathbf{T}&F\\
%				T&F&F&&&T&F&F&&F&F&&\mathbf{T}&F\\
%				F&T&T&&&F&F&T&&F&T&&\mathbf{T}&T\\
%				F&T&F&&&F&F&T&&F&F&&\mathbf{T}&T\\
%				F&F&T&&&F&F&F&&F&T&&\mathbf{T}&F\\
%				F&F&F&&&F&F&F&&F&F&&\mathbf{T}&F
%			\end{array}
%			\]}
%		
%		%3 letters, 3 connectives
%		
%		\item $\enot\bigl[(C\eor A) \eor B\bigr]$
%		\myanswer{ \hfill Contingent
%			\[
%			\begin{array}{ccc|cc@{}c@{}ccc@{}ccc@{}c}
%				A&B&C&\enot&(&(&C&\eor&A&)&\eor&B&)\\\hline
%				T&T&T&\mathbf{F}&&&T&T&T&&T&T&\\
%				T&T&F&\mathbf{F}&&&F&T&T&&T&T&\\
%				T&F&T&\mathbf{F}&&&T&T&T&&T&F&\\
%				T&F&F&\mathbf{F}&&&F&T&T&&T&F&\\
%				F&T&T&\mathbf{F}&&&T&T&F&&T&T&\\
%				F&T&F&\mathbf{F}&&&F&F&F&&T&T&\\
%				F&F&T&\mathbf{F}&&&T&T&F&&T&F&\\
%				F&F&F&\mathbf{T}&&&F&F&F&&F&F&
%			\end{array}
%			\]}
%		%	 	3 letters, 3 connectives
%		
%		\item $\bigl[(A\eand B) \eand\enot(A\eand B)\bigr] \eand C$
%		\myanswer{ \hfill Contradiction
%			\[
%			\begin{array}{ccc|c@{}c@{}ccc@{}cccc@{}ccc@{}c@{}ccc}
%				A&B&C&(&(&A&\eand&B&)&\eand&\enot&(&A&\eand&B&)&)&\eand&C\\\hline
%				T&T&T&&&T&T&T&&F&F&&T&T&T&&&\mathbf{F}&T\\
%				T&T&F&&&T&T&T&&F&F&&T&T&T&&&\mathbf{F}&F\\
%				T&F&T&&&T&F&F&&F&T&&T&F&F&&&\mathbf{F}&T\\
%				T&F&F&&&T&F&F&&F&T&&T&F&F&&&\mathbf{F}&F\\
%				F&T&T&&&F&F&T&&F&T&&F&F&T&&&\mathbf{F}&T\\
%				F&T&F&&&F&F&T&&F&T&&F&F&T&&&\mathbf{F}&F\\
%				F&F&T&&&F&F&F&&F&T&&F&F&F&&&\mathbf{F}&T\\
%				F&F&F&&&F&F&F&&F&T&&F&F&F&&&\mathbf{F}&F
%			\end{array}
%			\]}
%		
%		% 	3 letters, 5 connectives
%		
%		\item $(A \eand B) ]\eif[(A \eand C) \eor (B \eand D)]$
%		\myanswer{ \hfill Contingent
%			\[
%			\begin{array}{cccc|c@{}c@{}ccc@{}c@{}ccc@{}c@{}ccc@{}ccc@{}ccc@{}c@{}c}
%				A&B&C&D&(&(&A&\eand&B&)&)&\eif&(&(&A&\eand&C&)&\eor&(&B&\eand&D&)&)\\\hline
%				T&T&T&T&&&T&T&T&&&\mathbf{T}&&&T&T&T&&T&&T&T&T&&\\
%				T&T&F&F&&&T&T&T&&&\mathbf{F}&&&T&F&F&&F&&T&F&F&&\\
%			\end{array}
%			\]}
%		
%		%	4 letters, 5 connectives
%	\end{earg}
%	
%	\noindent\problempart
%	\label{pr.TT.TTorC4}
%	Determine whether each sentence is a tautology, a contradiction, or a contingent sentence. Justify your answer with a complete or partial truth table where appropriate.
%	\begin{earg}
%		\item  $\enot (A \eor A)$\vspace{.5ex}		\hfill			\myanswer{		Contradiction}	%	1 letter, 2 connectives
%		\item $(A \eif B) \eor (B \eif A)$\vspace{.5ex}	\hfill		\myanswer{			Tautology	}	%	2 letters, 2 connectives
%		\item $[(A \eif B) \eif A] \eif A$\vspace{.5ex}	\hfill			\myanswer{	  	Tautology		 }   %2 letters, 3 connectives
%		\item $\enot[( A \eif B) \eor (B \eif A)]$\vspace{.5ex}	\hfill		\myanswer{		Contradiction}	%	2 letters, 4 connectives
%		\item $(A \eand B) \eor (A \eor B)$\vspace{.5ex} 	\hfill			\myanswer{		Contingent	}	%2 letters, 5 connectives
%		\item $\enot(A\eand B) \eiff A$\vspace{.5ex} 		\hfill			\myanswer{	Contingent		}	%2 letters, 3 connectives
%		\item $A\eif(B\eor C)$\vspace{.5ex} 				\hfill			\myanswer{	Contingent		}	%3 letters, 2 connectives
%		\item $(A \eand\enot A) \eif (B \eor C)$\vspace{.5ex} 	\hfill		\myanswer{	Tautology		}	%3 letters, 4 connectives 
%		\item $(B\eand D) \eiff [A \eiff(A \eor C)]$\vspace{.5ex}\hfill		\myanswer{		Contingent	}	%	4 letters, 4 connectives
%		\item $\enot[(A \eif B) \eor (C \eif D)]$\vspace{.5ex} 	\hfill		\myanswer{	 Contingent 	}	%4 letters, 4 connectives
%	\end{earg}
%	
%	
%	\problempart
%	Some brackets are redundant. Which ones? To find out check the claims below and eventually propose further conventions for omitting some brackets.\begin{earg}
%		
%		\item `$((A \eand B) \eand C)$' and `$(A \eand (B \eand C))$' have the same truth table
%		\myanswer{\begin{center}
%				\begin{tabular}{c c c | d e e e f | d e e e f }
%					$A$ & $B$ & $C$ & $(A$&$\eand$& $B)$ &$ \eand$ & $C$ & $A$ & $\eand$ & $(B$&$\eand$&$C)$\\
%					\hline
%					T & T & T & T & T & T &  \TTbf{T} & T &T & \TTbf{T} & T & T& T \\
%					T & T & F & T& T & T &  \TTbf{F} & F & T& \TTbf{F} & T & F& F\\
%					T & F & T & T & F & F &  \TTbf{F} & T & T & \TTbf{F} & F & F & T \\
%					T & F & F &  T & F & F &  \TTbf{F} & F & T& \TTbf{F} & F & F & F\\
%					F & T & T & F & F & T & \TTbf{F} & T & F& \TTbf{F} & T & T & T\\
%					F & T & F & F & F & T & \TTbf{F} & F & F& \TTbf{F} &  T & F & F\\
%					F & F & T & F & F & F & \TTbf{F} & T & F& \TTbf{F} & F& F & T\\
%					F & F & F & F & F & F & \TTbf{F} & F & F& \TTbf{F} &  F& F & F\\
%				\end{tabular}
%		\end{center}}
%		
%		\item `$((A \eor B) \eor C)$' and `$(A \eor (B \eor C))$' have the same truth table
%		\myanswer{\begin{center}
%				\begin{tabular}{c c c | d e e e f | d e e e f }
%					$A$ & $B$ & $C$ & $(A$&$\eor$& $B)$ &$ \eor$ & $C$ & $A$ & $\eor$ & $(B$&$\eor$&$C)$\\
%					\hline
%					T & T & T & T & T & T &  \TTbf{T} & T &T & \TTbf{T} & T & T& T \\
%					T & T & F & T& T & T &  \TTbf{T} & F & T& \TTbf{T} & T & T& F\\
%					T & F & T & T & T & F &  \TTbf{T} & T & T & \TTbf{T} & F & T & T \\
%					T & F & F &  T & T& F &  \TTbf{T} & F & T& \TTbf{T} & F & F & F\\
%					F & T & T & F & T & T & \TTbf{T} & T & F& \TTbf{T} & T & T & T\\
%					F & T & F & F & T & T & \TTbf{T} & F & F& \TTbf{T} &  T & T & F\\
%					F & F & T & F & F & F & \TTbf{T} & T & F& \TTbf{T} & F& T & T\\
%					F & F & F & F & F & F & \TTbf{F} & F & F& \TTbf{F} &  F& F & F\\
%				\end{tabular}
%		\end{center}}
%		
%		\item `$((A \eor B) \eand C)$' and `$(A \eor (B \eand C))$' do not have the same truth table
%		\myanswer{\begin{center}
%				\begin{tabular}{c c c | d e e e f | d e e e f }
%					$A$ & $B$ & $C$ & $(A$&$\eor$& $B)$ &$ \eand$ & $C$ & $A$ & $\eor$ & $(B$&$\eand$&$C)$\\
%					\hline
%					T & T & T & T & T & T &  \TTbf{T} & T &T & \TTbf{T} & T & T& T \\
%					T & T & F & T& T & T &  \TTbf{F} & F & T& \TTbf{T} & T & F& F\\
%					T & F & T & T & T & F &  \TTbf{T} & T & T & \TTbf{T} & F & F & T \\
%					T & F & F &  T & T& F &  \TTbf{F} & F & T& \TTbf{T} & F & F & F\\
%					F & T & T & F & T & T & \TTbf{T} & T & F& \TTbf{T} & T & T & T\\
%					F & T & F & F & T & T & \TTbf{F} & F & F& \TTbf{F} &  T & F & F\\
%					F & F & T & F & F & F & \TTbf{F} & T & F& \TTbf{F} & F& F & T\\
%					F & F & F & F & F & F & \TTbf{F} & F & F& \TTbf{F} &  F& F & F\\
%				\end{tabular}
%		\end{center}}
%		
%		\item `$((A \eif B) \eif C)$' and `$(A \eif (B \eif C))$' do not have the same truth table
%		\myanswer{\begin{center}
%				\begin{tabular}{c c c | d e e e f | d e e e f }
%					$A$ & $B$ & $C$ & $(A$&$\eif$& $B)$ &$ \eif$ & $C$ & $A$ & $\eif$ & $(B$&$\eif$&$C)$\\
%					\hline
%					T & T & T & T & T & T &  \TTbf{T} & T &T & \TTbf{T} & T & T& T \\
%					T & T & F & T& T & T &  \TTbf{F} & F & T& \TTbf{F} & T & F& F\\
%					T & F & T & T & F & F &  \TTbf{T} & T & T & \TTbf{T} & F & T & T \\
%					T & F & F &  T & F& F &  \TTbf{T} & F & T& \TTbf{T} & F & T & F\\
%					F & T & T & F & T & T & \TTbf{T} & T & F& \TTbf{T} & T & T & T\\
%					F & T & F & F & T & T & \TTbf{F} & F & F& \TTbf{T} &  T & F & F\\
%					F & F & T & F & T & F & \TTbf{T} & T & F& \TTbf{T} & F& T & T\\
%					F & F & F & F & T & F & \TTbf{F} & F & F& \TTbf{T} &  F& T & F\\
%				\end{tabular}
%		\end{center}}
%	\end{earg}
%	
%	\noindent\problempart
%	Determine whether each the following pairs of sentences are logically equivalent using complete truth tables. If the two sentences really are logically equivalent, write ``equivalent.'' Otherwise write, ``not equivalent.''
%	\begin{earg}
%		\item $A$ and $A \eor A$
%		\item $A$ and $A \eand A$
%		\item $A \eor \enot B$ and $A\eif B$
%		\item $(A \eif B)$ and $(\enot B \eif \enot A)$
%		\item $\enot(A \eand B)$ and $\enot A \eor \enot B$
%		\item $ ((U \eif (X \eor X)) \eor U)$ and $\enot (X \eand (X \eand U))$
%		\item $ ((C \eand (N \eiff C)) \eiff C)$ and $(\enot \enot \enot N \eif C)$
%		\item $[(A \eor B) \eand C]$ and $[A \eor (B \eand C)]$
%		\item $((L \eand C) \eand I)$ and $L \eor C$
%	\end{earg}
%	
%	
%	\noindent\problempart
%	\label{pr.TT.satisfiable5}
%	Determine whether each collection of sentences is jointly consistent or jointly inconstent. Justify your answer with a complete or partial truth table where appropriate.
%	\begin{earg}
%		\item $A\eif A$, $\enot A \eif \enot A$, $A\eand A$, $A\eor A$ \vspace{.5ex} \hfill \myanswer{Consistent}
%		\item $A \eif \enot A$, $\enot A \eif A$\vspace{.5ex} \hfill \myanswer{Insatisfiable }
%		\item $A\eor B$, $A\eif C$, $B\eif C$\vspace{.5ex} \hfill \myanswer{Consistent}
%		\item $A \eor B$, $A \eif C$, $B \eif C$, $\enot C$\vspace{.5ex} \hfill \myanswer{	Insatisfiable}
%		\item $B\eand(C\eor A)$, $A\eif B$, $\enot(B\eor C)$\vspace{.5ex}  \hfill \myanswer{Insatisfiable}
%		\item $(A \eiff B) \eif B$,  $B \eif \enot (A \eiff B)$, $A \eor B$ \vspace{.5ex} \hfill \myanswer{Consistent}
%		\item $A\eiff(B\eor C)$, $C\eif \enot A$, $A\eif \enot B$\vspace{.5ex} \hfill \myanswer{Consistent}
%		\item  $A \eiff B$,  $\enot B \eor \enot A$,  $A \eif  B$ \vspace{.5ex} \hfill \myanswer{ Consistent}
%		\item $A \eiff B$, $A \eif C$, $B \eif D$, $\enot(C \eor D)$\vspace{.5ex} \hfill \myanswer{Consistent}
%		\item $\enot (A \eand \enot B)$,  $B \eif \enot A$, $\enot B$  \vspace{.5ex} \hfill \myanswer{Consistent}
%	\end{earg}
%	
%	\noindent\problempart Determine whether each argument is valid or invalid. Justify your answer with a complete or partial truth table where appropriate.
%	\label{pr.TT.valid5} 
%	\begin{enumerate}
%		\item $A\eif(A\eand\enot A)\therefore \enot A$ \hfill \myanswer{ Valid}
%		\item $A \eor B$, $A \eif B$, $B \eif A \therefore  A \eiff B$  \hfill \myanswer{ Valid}
%		\item $A\eor(B\eif A)\therefore \enot A \eif \enot B$ \hfill \myanswer{Valid}
%		\item $A \eor B$, $A \eif B$, $ B \eif A \therefore  A \eand B$ \hfill \myanswer{Valid}
%		\item $(B\eand A)\eif C$, $(C\eand A)\eif B\therefore (C\eand B)\eif A$ \hfill \myanswer{Invalid}
%		\item $\enot (\enot A \eor \enot B)$, $A \eif \enot C \therefore  A \eif (B \eif C)$ \hfill \myanswer{ Invalid}
%		\item $A \eand (B \eif C)$, $\enot C \eand (\enot B \eif \enot A)\therefore C \eand \enot C$ \hfill \myanswer{ Valid}
%		\item $A \eand B$, $\enot A \eif \enot C$, $B \eif \enot D \therefore  A \eor B$ \hfill \myanswer{ Invalid}
%		\item $A \eif B\therefore (A \eand B) \eor (\enot A \eand \enot B)$ \hfill \myanswer{ Invalid}
%		\item $\enot A \eif B$,$ \enot B \eif C $,$ \enot C \eif A \therefore  \enot A \eif (\enot B \eor \enot C)$ \hfill \myanswer{Invalid}
%		
%	\end{enumerate}
%	
%	\noindent\problempart Determine whether each argument is valid or invalid. Justify your answer with a complete or partial truth table where appropriate.
%	\label{pr.TT.valid6} 
%	\begin{enumerate}
%%		\item $A\eiff\enot(B\eiff A)\therefore A$ \hfill \myanswer{ Invalid}
%		\item $A\eor B$, $B\eor C$, $\enot A\therefore B \eand C$ \hfill \myanswer{ Invalid}
%		\item $A \eif C$, $E \eif (D \eor B)$, $B \eif \enot D\therefore (A \eor C) \eor (B \eif (E \eand D))$ \hfill \myanswer{ Invalid}
%		\item $A \eor B$, $C \eif A$, $C \eif B\therefore A \eif (B \eif C)$ \hfill \myanswer{ Invalid}
%		\item $A \eif B$, $\enot B \eor A\therefore A \eiff B$ \hfill \myanswer{ Valid}
%	\end{enumerate}
%	
%	
%	\noindent\problempart
%	\label{pr.TT.meta}
%	Are the following statements true? Why?
%	\begin{itemize}
%		\item if $Y$ is a logical consequence of $X$, then $X\eif Y$ is a logical truth.\myanswer{Yes}
%		\item if $X\eif Y$ is a logical truth, then $Y$ is a logical consequence of $X$.\myanswer{Yes}
%		\item if $X\eif Y\eand\enot Y$ is a logical truth, then $X$ is a logical contradiction.\myanswer{Yes}
%		\item if $X\eif Y\eand\enot Y$ is true relative to a valuation, then $X$ is a logical contradiction.\myanswer{No}
%		\item if $X\eor\enot X\eif Y$ is a logical truth, then $Y$ is a tautology.\myanswer{Yes}
%		\item if $X\eor\enot X\eif Y$ is true relative to a valuation, then $Y$ is a tautology.\myanswer{No}
%	\end{itemize}
%\end{practiceproblems}

%
%\chapter{TFL vs English connectives }\label{s:ParadoxesOfMaterialConditional}
%\teachingnote{Not covered in this course. This is just a taster for students who are interested.\\}
%Consider the sentence:
%\begin{earg}
%	\item[\ex{n:JanBald}] Jan is neither bald nor not-bald.
%\end{earg}
%To symbolize this sentence in TFL, we would offer something like `$\enot J \eand \enot \enot J$'. This a contradiction (check this with a truth-table), but sentence \ref{n:JanBald} does not itself seem like a contradiction; for we might have happily go on to add `Jan is on the borderline of baldness'!
%
%Third, consider the following sentence:
%\begin{earg}
%	\item[\ex{n:GodParadox}]	It's not the case that, if God exists, She answers malevolent prayers.
%	%	Aaliyah wants to kill Zebedee. She knows that, if she puts chemical A into Zebedee's water bottle, Zebedee will drink the contaminated water and die. Equally, Bathsehba wants to kill Zebedee. She knows that, if she puts chemical B into Zebedee's water bottle, then Zebedee will drink the contaminated water and die. But chemicals A and B neutralize each other; so that if both are added to the water bottle, then Zebedee will not die.
%\end{earg}
%Symbolizing this in TFL, we would offer something like `$\enot (G \eif M)$'. Now, `$\enot (G \eif M)$' entails `$G$' (again, check this with a truth table). So if we symbolize sentence \ref{n:GodParadox} in TFL, it seems to entail that God exists. But that's strange: surely even an atheist can accept sentence \ref{n:GodParadox}, without contradicting herself!
%
%One lesson of this is that the symbolization of \ref{n:GodParadox} as `$\enot(G \eif M)$' shows that \ref{n:GodParadox} does not express what we intend. Perhaps we should rephrase it as
%\begin{earg}
%	\item[\ex{n:GodParadox2}] If God exists, She does not answer malevolent prayers.
%\end{earg}
%and symbolize \ref{n:GodParadox2} as `$G \eif \enot M$'.  Now, if atheists are right, and there is no God, then `$G$' is false and so `$G \eif \enot M$' is true, and the puzzle disappears. However, if `$G$' is false, `$G \eif M$', i.e.\ `If God exists, She answers malevolent prayers', is \emph{also} true!
%
%In different ways, these  examples highlight some of the limits of working with a language (like TFL) that can \emph{only} handle truth-functional connectives. Moreover, these limits give rise to some interesting questions in philosophical logic. The case of Jan's baldness (or otherwise) raises the general question of what logic we should use when dealing with \emph{vague} discourse. The case of the atheist raises the question of how to deal with the (so-called) \emph{paradoxes of the material conditional}. Part of the purpose of this course is to equip you with the tools to explore these questions of \emph{philosophical logic}. But we have to walk before we can run; we have to become proficient in using TFL, before we can adequately discuss its limits, and consider alternatives.


%\chapter{FROM JS. DELETING!}
%\section{Other logical notions}
%
%It is worth highlighting a peculiar feature of the definition of validity: there are valid arguments without a premise. Consider the argument
%$$\therefore A\eor\enot A.$$
%The argument is valid if there is no valuation such that all premises are true but the conclusion false, that is, if there is no valuation such that $A\eor\enot A$ is false. This can be quickly verified as there are only two valuations to consider:
%
%\begin{center}
%	\begin{tabular}{rc|deeef}
%		Valuation&$A$&$\enot$&$ A$&$\eor$&$ A$\\\hline
%		$v_1$&T & F &&\textbf{T}\\
%		$v_2$&F & T& & \textbf{T}
%	\end{tabular}
%\end{center}
%
%$A\eor\enot A$ is true on both $v_1$ and $v_2$. The argument is valid. The conclusions of arguments with no premises are called \define{logical truths} or \define{tautologies}.
%
%\factoidbox{\label{logical truth}A TFL-sentence $X$ is called a \define{logical truth} or \define{tautology} if and only if the TFL-argument $\therefore X$ is valid.
%
%$X$ is a logical truth iff it is true on all valuations.}\todo{add in the true on all valuations!}
%
%A logical truth is a TFL-sentence that is true relative to every valuation. Alternatively, one can say that a TFL sentence is a logical truth if and only if it is a \define{logical consequence} of (\define{follows from}) any TFL sentence (Exercise: explain why).
%
%
%\factoidbox{\label{TFLcons}$Y$ is a \define{logical consequence} of $X_1,\ldots,X_n$ if and only if the argument $X_1,\ldots,X_n\therefore Y$ is valid.}
%
%Notice one important  difference between validity and consequence: the former is a property of TFL-arguments while the latter is a property of TFL-sentences!
%
%A logical truth follows from every sentence. In contrast, every sentence  follows from a \define{logical contradiction}.
%\factoidbox{\label{TFLcontra}$X$ is a \define{logical contradiction} if and only if there is no valuation $v$ such that X is true relative to $v$.}
%Every sentence $Y$ follows from a contradiction $X$, since $X\therefore Y$ is valid: there is no valuation on which $X$ is true (and $Y$ false).
%
%We end this section by introducing three further important logical notions: \define{consistency}, \define{contingency} and \define{logical equivalence}.
%\factoidbox{\label{TFLconsist}A collection of TFL-sentences $X_1,\ldots,X_n$ (with possibly $n=1$) is \define{consistent} iff there is a valuation relative to which $X_1,\ldots,X_n$ are true. Otherwise the collection is \define{inconsistent}.}
%
%\factoidbox{\label{TFLconting}A sentence of TFL  is called \define{contingent}. If it is part of a consistent collection and not a logical truth.}
%Any collection of sentences that contains a logical contradiction is inconsistent, and so a logical contradiction is not a contingent sentence. A contingent sentence is a sentence for which we can find a valuation relative to which it is true, but we can also find a valuation relative to which it is false. (Exercise: every sentence of TFL is either a logical truth, a logical contradiction, or a contingent sentence. Why?) 
%
%The final notion we introduce is that of logical equivalence.
%\factoidbox{\label{TFLequiv}$X$ and $Y$ are \define{logically equivalent} if and only if $X$ is a logical consequence of $Y$ and $Y$ a logical consequence of $X$.}
%
%In terms of truth tables, $X$ and $Y$ are logically equivalent if and only if, for every valuation, $v$: $X$ is true relative to $v$ if and only if $Y$ is true relative to $v$.
%
%If two sentences are logically equivalent they mean the same thing from the perspective of TFL: at least semantically TFL cannot tell the two sentences apart. Examples of logically equivalent sentences include $A$ and $A\eand A$, or $\enot A\eor B$ and $A\eif B$. This can be checked by means of truth tables (Exercise!).
%
%Again, it is easy to test for logical equivalence using truth tables. Consider the sentences `$\enot(P \eor Q)$' and `$\enot P \eand \enot Q$'. Are they logically equivalent? To find out, we construct a truth table.
%\begin{center}
%	\begin{tabular}{rc c|d e e f |d e e e f}
%		Valuation&$P$&$Q$&\enot&$(P$&\eor&$Q)$&\enot&$P$&\eand&\enot&$Q$\\
%		\hline
%		$v_1:$&T & T & \TTbf{F} & T & T & T & F & T & \TTbf{F} & F & T\\
%		$v_2:$&	T & F & \TTbf{F} & T & T & F & F & T & \TTbf{F} & T & F\\
%		$v_3:$&	F & T & \TTbf{F} & F & T & T & T & F & \TTbf{F} & F & T\\
%		$v_4:$&	F & F & \TTbf{T} & F & F & F & T & F & \TTbf{T} & T & F
%	\end{tabular}
%\end{center}
%
%On every valuation where `$\enot(P \eor Q)$' is true, `$\enot P \eand \enot Q$' is also true; and on every valuation where 
%`$\enot P \eand \enot Q$' is true,  `$\enot(P \eor Q)$' is true. So they are logically equivalent. 
%
%
%
%%\color{red}
%%
%%
%%To better grasp what a valuation is, it makes sense to look at the \define{truth table} of the sentence `$A\eand B$', which looks as follows 
%%(from now on, in contrast to \S\ref{sec:tt}, we no longer display the explanatory text, and replace $\leadsto$ by a vertical line):\todo{J added this sent. Remove it!!! The squiggly is for the metavar version. }
%%
%%\begin{center}
%%\begin{tabular}{ccc|c}
%%$A$&$B$&&$A\eand B$\\\hline
%%T & T && T\\
%%T & F && F\\
%%F & T && F\\
%%F & F && F
%%\end{tabular}
%%\end{center}
%%
%%Now the first two rows of every horizontal line in the above table specify a valuation for the atomic sentence $A$ and $B$: according to the valuation given by the first line both $A$ and $B$ are true, according to the valuation given by the second line $A$ is true but $B$ is false, and so on. Let use $v_1,v_2,\ldots$ as names for different valuations, then we can make valuations explicit in the truth table above:
%%\begin{center}
%%\begin{tabular}{c|ccc|c}
%%Valuation&$A$&$B$&&$A\eand B$\\\hline
%%$v_1$&T & T && T\\
%%$v_2$&T & F && F\\
%%$v_3$&F & T && F\\
%%$v_4$&F & F && F
%%\end{tabular}
%%\end{center}
%
%
%
%%
%% The truth table then tells us that `$A\eand B$' is true relative to the valuation $v_1$, but false relative to the $v_2,v_3$, and $v_4$. We can generalize idea and give an inductive definition of when \define{a sentence of TFL is true relative to a given valuation}.  The definition specifies rules to compute the truth of sentence for every formation rule of the definition of a TFL sentence.
%%
%% \factoidbox{\label{Truthint}Let $v$ be a valuation. Then
%%	\begin{enumerate}
%%		\item An atomic sentence $X$ is true relative to $v$, if and only if $v$ assigns the value T to $X$.
%%		\item a sentence $\enot\metaX$ is true relative to $v$, if and only if $\metaX$ is not true relative to $v$.
%%		\item a sentence $(\metaX\eand\metaY)$ is true relative to $v$, if and only if $\metaX$ and $\metaY$ are both true relative to $v$.
%%		\item a sentence $(\metaX\eor\metaY)$ is true relative to $v$, if and only if $\metaX$ or $\metaY$is true relative to $v$.
%%		\item a sentence $(\metaX\eif\metaY)$  is true relative to $v$, if and only if $\metaX$ is not true or $\metaY$ is true relative to $v$
%%		%\item If \metaX and \metaY are sentences, then $(\metaX\eiff\metaY)$ is a sentence.
%%		%\item Nothing else is a sentence.
%%	\end{enumerate}
%%	}
%%
%%	\newglossaryentry{Truthval}
%%{
%%name=Truth relative to a valuation,
%%description={a list of recursive rules that allow us to determine whether a sentence of TFL is true given a particular valuation, p.~\pageref{Truthval}}
%%}
%%
%%Let's go through the five rules of the definition step by step. The first rule should be relatively immediate: an atomic sentence is true relative to a valuation, if and only if the valuation says it has value ``True''. For rule 2, we look at the truth table for negation: the truth table tells us that a sentence $\enot X$ is true whenever $X$ is false (not true). That is, $\enot X$ is true relative to a valuation, if $X$ is not true (false) relative to that valuation. For the case of conjunction, we can revisit the discussion we used to motivate talking about truth relative to a valuation. We say that the conjunction `$A\eand B$' was only true relative to the valuation $v_1$, that is, the valuation relative to which both `$A$' and `$B$' are true. Our reasoning was not specific to the specific conjunction `$A\eand B$' but applicable to all sentences of the form $X\eand Y$ with arbitrary conjuncts $X$ and $Y$.
%%
%%For the fourth rule we need to look at the truth table for disjunction:
%%\begin{center}
%%\begin{tabular}{ccc|c}
%%$X$&$Y$&&$X\eor Y$\\\hline
%%T & T && T\\
%%T & F && T\\
%%F & T && T\\
%%F & F && F
%%\end{tabular}
%%\end{center}
%%According to the truth table $X\eor Y$ is true on lines 1-3, that is, inspecting these three lines one can see that it suffices for $X$ or $Y$ to be true for $X\eor Y$ to be true. This is precisely what the third rule says.
%%
%%For the conditional (rule 5) we also reexamine the truth table $X\eif Y$:
%%\begin{center}
%%\begin{tabular}{ccc|c}
%%$X$&$Y$&&$X\eif Y$\\\hline
%%T & T && T\\
%%T & F && F\\
%%F & T && T\\
%%F & F && T
%%\end{tabular}
%%\end{center}
%%According to the truth table a sentence $X\eif Y$ is true on line 1, 3, and 4. Let's check whether the rule is correct: $X\eif Y$ has to be true, if and only if $X$ is false or $Y$ is true. If $X$ is false, we are either in line 3 or in line 4, but in both lines $X\eif Y$ turns out true. If $Y$ is true, we are either in line 1 or line 3 of the truth table and $X\eif Y$ is true in both lines. So rule 5 is correct.
%%
%%With the definition in place we can now determine whether an arbitrary sentence of TFL is true relative to a given valuation $v$. This requires giving the \define{complete truth table} for this sentence. For example, consider the sentence $A\eand(B\eor C)$. This sentence has 3 atomic sentences which means that there are $2^3$ different ways to assign truth values to the three atomic sentences, which in turn means that there are $2^3$ valuation to consider in the complete truth table: 
%%
%%\begin{center}
%%\begin{tabular}{ccc|c}
%%$A$&$B$&$C$&\\\hline
%%T& T &T&\\
%%T & T & F&\\
%%T & F & T& \\
%%T & F &F&\\
%%F & T &T&\\
%%F& T & F & \\
%%F & F & T &\\
%%F & F & F& 
%%\end{tabular}
%%\end{center}
%%
%%
%%To give the complete truth table of this sentence we first need to give the truth table for $B\eor C$ using the truth rule for disjunction:
%%
%%\begin{center}
%%\begin{tabular}{ccc|c}
%%$A$&$B$&$C$&$B\eor C$\\\hline
%%T& T &T&T\\
%%T & T & F&T\\
%%T & F & T&T \\
%%T & F &F&F\\
%%F & T &T&T\\
%%F& T & F & T\\
%%F & F & T &T\\
%%F & F & F& F
%%\end{tabular}
%%\end{center}
%%
%%The final step in the construction of the truth table is to combine the columns of $A$ and $B\eor C$ and to compute the truth values of $A\wedge (B\eor C)$. We get the following truthtable:
%%
%%\begin{center}
%%\begin{tabular}{ccc|c||c}
%%$A$&$B$&$C$&$B\eor C$&$A\eand(B\eor C)$\\\hline
%%\textbf{T} & \textbf{T} &\textbf{T}& T & \textbf{T}\\
%%\textbf{T} & \textbf{T} & F& T & \textbf{T}\\
%%\textbf{T} & F & \textbf{T}& T& \textbf{T}\\
%%T & F &F&F& F\\
%%F & T &T&T&F\\
%%F& T & F & T &F\\
%%F & F & T &T &F\\
%%F & F & F& F& F
%%\end{tabular}
%%\end{center}
%%
%%By inspecting the truth table we see that `$A\eand(B\eor C)$' is true relative relative to an interpretation $v$ if and only if `$A$' is true relative relative to v and at least one of `$B$' and `$C$' is also true relative to $v$. To this effect we only need to check the lines 1,2, and 3 of the truth table, as these are the only lines in which `$A\eand(B\eor C)$' is true. Using the above strategy we can construct truth tables for all sentences of TFL. Given a truth table for some sentence of TFL it is easy to check under which conditions (valuations) the sentence is true (false). In \S\ref{sec:ctt} we will provide and discuss a \emph{How to guide} for constructing complete truth tables.
%%
%% \section{Validity and other logical notions}\label{sec:valtt}
%% In \S\ref{s:Valid} we said that an argument was valid, if and only if there is no interpretation such that all premises are true but the conclusion false. At that point the definition was suggestive, but we lacked a clear understanding of `interpretation' and when a sentence is true relative to an interpretation. But for the language of TFL we can now turn the informal definition in a precise and rigorous definition.
%%
%% Recall that an argument consisted of a number of premises together with a conclusion. A TFL-argument then can be written as $X_1,\ldots,X_n\therefore Y$ where $X_1,\ldots,X_n, Y$ are sentences of TFL, and $X_1,\ldots,X_n$ are the premises and $Y$ the conclusion of the argument.
%%
%% \factoidbox{\label{TFLvalid}A TFL-argument $X_1,\ldots,X_n\therefore Y$ is \define{valid} if and only if there is no valuation $v$ such that $X_1$ is true relative to $v$ and \ldots ~and $X_n$ is true relative to $v$, but $Y$ is false relative to $v$.}
%%
%% We can now investigate whether a TFL-argument is valid or not. The difficult bit is to show that there is no valuation on which all premises are true but the conclusion false. After all there are many different valuations. Fortunately, only the atomic sentences that occur in the premises and the conclusion of the argument will be relevant for deciding whether an argument is valid or not. Consider the argument:
%% $$\enot B,A\eif B\therefore\enot A$$
%% In this case there are only four different ways to assign truth values to $A$ and $B$, that is, we only need to consider four valuation. This means that we can check whether the argument is valid via the following truth table:
%%
%% \begin{center}
%%\begin{tabular}{c|cc|c|c||c}
%%Valuation&$A$&$B$&$\enot B$&$A\eif B$&$\enot A$\\\hline
%%$v_1$&T & T & F & T & F\\
%%$v_2$&T & F & T & F & F\\
%%$v_3$&F & T & F & T & T\\
%%$v_4$&F & F & \textbf{T}& \textbf{T} & \textbf{T}
%%\end{tabular}
%%\end{center}
%%
%%In the truth table we have used the truth table for negation to compute the truth value of $\enot B$ ($\enot A$) from $B$ ($A$) and the truth value of the conditional to obtain the value of $A\eif B$ from the values of $A$ and $B$. By inspecting the truth table we see that only relative to valuation $v_4$ all the premises of the argument ($\enot B$ and $A\eif B$) are true. But relative to $v_4$ the conclusion $\enot A$ is true likewise. There is no valuation relative to which all premises are true, but the conclusion false. The argument is valid.
%%
%%Unfortunately, when we consider arguments that involve more atomic sentences we need to consider more valuations: for $n$ atomic sentences there are $2^n$ different options for assigning truth values to these atomic sentences, that is we need to consider $2^n$ different valuations. Hence, the more atomic sentences one needs to consider the longer the truth tables will be. Fortunately, there is a straightforward strategy that makes it easy to find all the different valuations.
%%
%%
%
%\subsection{Summary of how to use truth tables to check these notions}
%We have already seen that truth tables can be used to find out relative to which valuations a sentence is true or whether an argument is valid. They can also be used to determine whether sentences follow from each other or whether they are consistent or inconsistent. Let sum how to check for the various logical notions using truth tables:
%
%\begin{description}
%\item[Validity:]Construct a truth table with columns for all atomic sentences occurring in the premises and the conclusion; columns for all subsentences of the premises and conclusion, columns for all the premises and the conclusion. If in all valuation (horizontal line) in which all premises are true the conclusion is true too, then the argument is valid. Otherwise it is invalid.
%\item[Consequence]To check that $Y$ is a consequence of $X_1,\ldots,X_n$, we need to check whether $X_1,\ldots,X_n\therefore Y$ is valid (see above).
%\item[Logical Equivalence] To check whether $X$ and $Y$ are logically equivalent, we need to check whether $X\therefore Y$ and $Y\therefore X$ are valid (see above). This will be the case if in a truth table for both $X$ and $Y$ whenever $X$ is true relative to valuation so is $Y$ and vice versa.
%\item[Logical truth/Tautology] $X$ is a tautology if and only if $\therefore X$ is valid, that is, if in the truth table for $X$, $X$ receives value T relative to all valuations.
%\item[Logical contradiction]  $X$ is a logical contradiction if and only if in the truth table for $X$, $X$ receives value F relative to all valuations.
%\item[Consistency]To check whether $X_1,\ldots,X_n$ are consistent construct a truth table with columns for all atomic sentences occurring in $X_1,\ldots,X_n$, columns for all subsentences of $X_1,\ldots,X_n$, and columns for $X_1,\ldots,X_n$. If there is a valuation (horizontal line) relative to which all of $X_1,\ldots,X_n$ receive value T, they are consistent. Otherwise they are inconsistent.
%\end{description}
%
%\begin{practiceproblems}\label{pr.TT.TTorC}
%\problempart
%\label{pr.TT.TTorC}
%\label{pr.TT.TTorC}
%Complete truth tables for each of the following:
%\begin{earg}
%\item $A \eif A$ %taut
%\myanswer{\begin{center}
%\begin{tabular}{c | def}
%$A$ & $A$&$\eif$&$A$\\
%\hline
% T & T&\TTbf{T}&T\\
%F & F&\TTbf{T}&F
%\end{tabular}
%\end{center}
%}
%
%\item $C \eif\enot C$ %contingent
%\myanswer{\begin{center}
%\begin{tabular}{c | d e e f}
%$C$ & $C$&$\eif$&$\enot$&$C$\\
%\hline
% T & T & \TTbf{F}& F& T\\
%F & F & \TTbf{T}& T& F\\
%\end{tabular}
%\end{center}}
%\item $(A \eif B) \eif (\enot A\eor B)$ %tautology
%
%%\myanswer{\begin{center}
%%\begin{tabular}{c c | d e e e e e e e f}
%%$A$ & $B$&$(A$&$\eif$&$B)$&$\eif$&$(\enot A$&$\eor$&$B)$ \\
%%\hline
%%T & T & T & T & T & \TTbf{T} & F & T & T \\
%%T & F & T & F & F & \TTbf{T} & F & T & T\\
%%F & T & F & F & T & \TTbf{T} & F & F & T\\
%%F & F & F & T & F & \TTbf{T} & T & F & F & T & F
% %\end{tabular}
%%\end{center}}
%\item $(A \eif B) \eor (B \eif A)$ % taut
%
%\myanswer{
%\begin{center}
%\begin{tabular}{c c | d e e e e e f}
%$A$ & $B$&$(A$&$\eif$&$B)$&$\eor$&$(B$&$\eif$&$A)$ \\
%\hline
%T & T & T & T & T & \TTbf{T} & T & T & T\\
%T & F & T & F & F & \TTbf{T} & F & T & T\\
%F & T & F & T & T & \TTbf{T} & T & F & F \\
%F & F & F & T & F & \TTbf{T} & F & T & F
% \end{tabular}
%\end{center}}
%\item $(A \eand B) \eif (B \eor A)$  %taut
%
%\myanswer{
%\begin{center}
%\begin{tabular}{c c | d e e e e e f}
%$A$ & $B$&$(A$&$\eand$&$B)$&$\eif$&$(B$&$\eor$&$A)$ \\
%\hline
%T & T & T & T & T & \TTbf{T} & T & T & T\\
%T & F & T & F & F & \TTbf{T} & F & T & T\\
%F & T & F & F & T & \TTbf{T} & T & T & F \\
%F & F & F & F & F & \TTbf{T} & F & F & F
% \end{tabular}
%\end{center}}
%\item $\enot(A \eor B) \eif (\enot A \eand \enot B)$ %taut
%
%\myanswer{\begin{center}
%\begin{tabular}{c c | d e e e e e e e e f}
%$A$ & $B$&$\enot$&$(A$&$\eor$&$B)$&$\eif$&$(\enot$&$A$&$\eand$&$\enot$&$B)$\\
%\hline
%T & T & F & T & T & T & \TTbf{T} & F & T & F & F & T\\
%T & F & F & T& T & F & \TTbf{T} & F & T & F & T & F\\
%F & T & F & F & T & T & \TTbf{T} & T & F & F & F & T\\
%F & F & T & F & F & F & \TTbf{T} & T & F & T & T & F
% \end{tabular}
%\end{center}}
%\item $(\enot A \eand \enot B) \eif\enot(A \eor B) $ %taut
%
%%\myanswer{\begin{center}
%%\begin{tabular}{c c | d e e e e e e e e f}
%%$A$ & $B$&$\enot$&$(A$&$\eor$&$B)$&$\leftarrow$&$(\enot$&$A$&$\eand$&$\enot$&$B)$\\
%%\hline
%%T & T & F & T & T & T & \TTbf{T} & F & T & F & F & T\\
%%T & F & F & T& T & F & \TTbf{T} & F & T & F & T & F\\
%%F & T & F & F & T & T & \TTbf{T} & T & F & F & F & T\\
%%F & F & T & F & F & F & \TTbf{T} & T & F & T & T & F
%% \end{tabular}
%%\end{center}}
%
%
%\item $\bigl[(A\eand B) \eand\enot(A\eand B)\bigr] \eand C$ %contradiction
%
%\myanswer{\begin{center}
%\begin{tabular}{c c c | d e e e e e e e e f}
%$A$ & $B$&$C$&$\bigl[(A$&$\eand$&$B)$&$ \eand$&$\enot$&$(A$&$\eand$&$B)\bigr]$&$\eand$&$C$\\
%\hline
%T & T & T & T & T & T & F & F & T & T & T & \TTbf{F} & T\\
%T & T & F & T& T & T & F & F & T & T & T & \TTbf{F}& F\\
%T & F & T & T & F & F & F & T & T & F & F & \TTbf{F} & T\\
%T & F & F & T & F & F & F & T & T & F & F & \TTbf{F} & F\\
%F & T & T & F & F & T & F & T & F & F & T & \TTbf{F} & T\\
%F & T & F & F & F & T & F & T & F & F & T & \TTbf{F} & F\\
%F & F & T & F & F & F & F & T & F & F & F & \TTbf{F} & T\\
%F & F & F & F & F & F & F & T & F & F & F & \TTbf{F} & F
%\end{tabular}
%\end{center}}
%\item $[(A \eand B) \eand C] \eif B$ %taut
%
%\myanswer{\begin{center}
%\begin{tabular}{c c c | d e e e e e f}
%$A$ & $B$&$C$&$[(A$&$\eand$&$B)$&$\eand$&$C]$&$\eif$&$B$\\
%\hline
%T & T & T & T & T & T & T & T & \TTbf{T} & T\\
%T & T & F & T & T & T & F & F & \TTbf{T} & T\\
%T & F & T & T & F & F & F & T & \TTbf{T} & F\\
%T & F & F & T & F & F & F & F & \TTbf{T} & F\\
%F & T & T & F & F & T & F & T & \TTbf{T} & T\\
%F & T & F & F & F & T & F & F & \TTbf{T} & T\\
%F & F & T & F & F & F & F & T & \TTbf{T} & F\\
%F & F & F & F & F & F & F & F & \TTbf{T} & F\\
%\end{tabular}
%\end{center}}
%\item $\enot\bigl[(C\eor A) \eor B\bigr]$ %contingent
%
%\myanswer{\begin{center}
%\begin{tabular}{c c c | d e e e e f}
%$A$ & $B$&$C$&$\enot\bigl[($&$C$&$\eor$&$A)$&$\eor$&$B\bigr]$\\
%\hline
%T & T & T & \TTbf{F} & T & T & T & T & T\\
%T & T & F & \TTbf{F} & F & T & T & T & T\\
%T & F & T & \TTbf{F} & T & T & T & T & F\\
%T & F & F & \TTbf{F} & F & T & T & T & F\\
%F & T & T & \TTbf{F} & T & T & F & T & T\\
%F & T & F & \TTbf{F} & F & F & F & T & T\\
%F & F & T & \TTbf{F} & T & T & F & T & F\\
%F & F & F & \TTbf{T} & F & F & F & F & F
%\end{tabular}
%\end{center}}
%\end{earg}
%
%
%\problempart
%Some brackets are redundant. Which ones? To find out check the claims below and eventually propose further conventions for omitting some brackets.\begin{earg}
%
%	\item `$((A \eand B) \eand C)$' and `$(A \eand (B \eand C))$' have the same truth table
%\myanswer{\begin{center}
%\begin{tabular}{c c c | d e e e f | d e e e f }
%$A$ & $B$ & $C$ & $(A$&$\eand$& $B)$ &$ \eand$ & $C$ & $A$ & $\eand$ & $(B$&$\eand$&$C)$\\
%\hline
%T & T & T & T & T & T &  \TTbf{T} & T &T & \TTbf{T} & T & T& T \\
%T & T & F & T& T & T &  \TTbf{F} & F & T& \TTbf{F} & T & F& F\\
%T & F & T & T & F & F &  \TTbf{F} & T & T & \TTbf{F} & F & F & T \\
%T & F & F &  T & F & F &  \TTbf{F} & F & T& \TTbf{F} & F & F & F\\
%F & T & T & F & F & T & \TTbf{F} & T & F& \TTbf{F} & T & T & T\\
%F & T & F & F & F & T & \TTbf{F} & F & F& \TTbf{F} &  T & F & F\\
%F & F & T & F & F & F & \TTbf{F} & T & F& \TTbf{F} & F& F & T\\
%F & F & F & F & F & F & \TTbf{F} & F & F& \TTbf{F} &  F& F & F\\
%\end{tabular}
%\end{center}}
%
%	\item `$((A \eor B) \eor C)$' and `$(A \eor (B \eor C))$' have the same truth table
%\myanswer{\begin{center}
%\begin{tabular}{c c c | d e e e f | d e e e f }
%$A$ & $B$ & $C$ & $(A$&$\eor$& $B)$ &$ \eor$ & $C$ & $A$ & $\eor$ & $(B$&$\eor$&$C)$\\
%\hline
%T & T & T & T & T & T &  \TTbf{T} & T &T & \TTbf{T} & T & T& T \\
%T & T & F & T& T & T &  \TTbf{T} & F & T& \TTbf{T} & T & T& F\\
%T & F & T & T & T & F &  \TTbf{T} & T & T & \TTbf{T} & F & T & T \\
%T & F & F &  T & T& F &  \TTbf{T} & F & T& \TTbf{T} & F & F & F\\
%F & T & T & F & T & T & \TTbf{T} & T & F& \TTbf{T} & T & T & T\\
%F & T & F & F & T & T & \TTbf{T} & F & F& \TTbf{T} &  T & T & F\\
%F & F & T & F & F & F & \TTbf{T} & T & F& \TTbf{T} & F& T & T\\
%F & F & F & F & F & F & \TTbf{F} & F & F& \TTbf{F} &  F& F & F\\
%\end{tabular}
%\end{center}}
%
%	\item `$((A \eor B) \eand C)$' and `$(A \eor (B \eand C))$' do not have the same truth table
%\myanswer{\begin{center}
%\begin{tabular}{c c c | d e e e f | d e e e f }
%$A$ & $B$ & $C$ & $(A$&$\eor$& $B)$ &$ \eand$ & $C$ & $A$ & $\eor$ & $(B$&$\eand$&$C)$\\
%\hline
%T & T & T & T & T & T &  \TTbf{T} & T &T & \TTbf{T} & T & T& T \\
%T & T & F & T& T & T &  \TTbf{F} & F & T& \TTbf{T} & T & F& F\\
%T & F & T & T & T & F &  \TTbf{T} & T & T & \TTbf{T} & F & F & T \\
%T & F & F &  T & T& F &  \TTbf{F} & F & T& \TTbf{T} & F & F & F\\
%F & T & T & F & T & T & \TTbf{T} & T & F& \TTbf{T} & T & T & T\\
%F & T & F & F & T & T & \TTbf{F} & F & F& \TTbf{F} &  T & F & F\\
%F & F & T & F & F & F & \TTbf{F} & T & F& \TTbf{F} & F& F & T\\
%F & F & F & F & F & F & \TTbf{F} & F & F& \TTbf{F} &  F& F & F\\
%\end{tabular}
%\end{center}}
%
%	\item `$((A \eif B) \eif C)$' and `$(A \eif (B \eif C))$' do not have the same truth table
%\myanswer{\begin{center}
%\begin{tabular}{c c c | d e e e f | d e e e f }
%$A$ & $B$ & $C$ & $(A$&$\eif$& $B)$ &$ \eif$ & $C$ & $A$ & $\eif$ & $(B$&$\eif$&$C)$\\
%\hline
%T & T & T & T & T & T &  \TTbf{T} & T &T & \TTbf{T} & T & T& T \\
%T & T & F & T& T & T &  \TTbf{F} & F & T& \TTbf{F} & T & F& F\\
%T & F & T & T & F & F &  \TTbf{T} & T & T & \TTbf{T} & F & T & T \\
%T & F & F &  T & F& F &  \TTbf{T} & F & T& \TTbf{T} & F & T & F\\
%F & T & T & F & T & T & \TTbf{T} & T & F& \TTbf{T} & T & T & T\\
%F & T & F & F & T & T & \TTbf{F} & F & F& \TTbf{T} &  T & F & F\\
%F & F & T & F & T & F & \TTbf{T} & T & F& \TTbf{T} & F& T & T\\
%F & F & F & F & T & F & \TTbf{F} & F & F& \TTbf{T} &  F& T & F\\
%\end{tabular}
%\end{center}}
%\end{earg}
%
%\problempart
%Write complete truth tables for the following sentences and mark the column that represents the truth values for the whole sentence.
%
%\begin{earg}
%
%
% \item $\enot [(X \eand Y) \eor (X \eor Y)]$
%
%\myanswer{
%\begin{center}
%\begin{tabular}{c|c|ccccccc}
%\cline{2-2}
%~	&	\enot	&	 [(X 	&	\eand& 	Y) 	&	\eor 	&	(X 	&	\eor 	&	Y)] \\
%\cline{2-9}
%	&	F	&	T	&	T	&	T	&	T	&	T	&	T	&	T	\\
%	&	F	&	T	&	F	&	F	&	T	&	T	&	T	&	F	\\
%	&	F	&	F	&	F	&	T	&	T	&	F	&	T	&	T	\\
%	&	T	&	F	&	F	&	F	&	F	&	F	&	F	&	F	\\
%\cline{2-2}
%\end{tabular}
%\end{center}
%}
%
%
%\item $(A \eif B) \eif (\enot B\eif \enot A)$
%
%\myanswer{
%\begin{center}
%\begin{tabular}{cccc|c|ccccc}
%\cline{5-5}
%~	&	(A 	&	\eif	&	B)	&	 \eif 	&	(\enot&	B 	&	\eif 	&	 \enot 	& 	 A) \\
%\cline{2-10}
%	&	T	&	T	&	T	&	T		&	F	 &	T	&	T	&	F		&	T	\\
%	&	T	&	F	&	F	&	T		&	T	 &	F	&	F	&	F		&	T	\\
%	&	F	&	T	&	T	&	T		&	F	 &	T	&	T	&	T		&	F	\\
%	&	F	&	T	&	F	&	T		&	T	 &	F	&	T	&	T		&	F	\\
%\cline{5-5}
%\end{tabular}
%\end{center}
%}
%
%\item $[C \eif (D \eor E)] \eand \enot C$
%
%%\myanswer{
%%\begin{center}
%%\begin{tabular}{cccccc|c|cc}
%%\cline{7-7}
%%~	&	[C 	&	\eiff 	&	(D 	&	\eor 	&	E)] 	&	\eand 	&	 \enot 	&	 C \\
%%\cline{2-9}
%%	&	T	&	T	&	T	&	T	&	T	&	F		&	F		&	T	\\
%%	&	T	&	T	&	T	&	T	&	F	&	F		&	F		&	T	\\
%%	&	T	&	T	&	F	&	T	&	T	&	F		&	F		&	T	\\
%%	&	T	&	F	&	F	&	F	&	F	&	F		&	F		&	T	\\
%%	&	F	&	F	&	T	&	T	&	T	&	F		&	T		&	F	\\
%%	&	F	&	F	&	T	&	T	&	F	&	F		&	T		&	F	\\
%%	&	F	&	F	&	F	&	T	&	T	&	F		&	T		&	F	\\
%%	&	F	&	T	&	F	&	F	&	F	&	T		&	T		&	F	\\
%%\cline{7-7}
%%\end{tabular}
%%\end{center}
%%}
%
%\item $\enot(G \eand (B \eand H)) \eif (G \eor (B \eor H))$
%
%%\myanswer{
%%\begin{center}
%%\begin{tabular}{ccccccc|c|ccccc}
%%\cline{8-8}
%%~	&\enot&	(G 	&\eand &	(B 	&	 \eand 	&	 H))	&	\eiff 	&	(G 	& \eor 	& (B 	& \eor	& H))	\\
%%\cline{2-13}
%%	&F	   &	T	&	  T &	T	&	T		&	T	&	F	&	T	&	T	&	T	&	T	&	T	\\
%%	&T	   &	T	&	  F &	T	&	F		&	F	&	T	&	T	&	T	&	T	&	T	&	F	\\
%%	&T	   &	T	&	 F  &	F	&	F		&	T	&	T	&	T	&	T	&	F	&	T	&	T	\\
%%	&T	   &	T	&	 F  &	F	&	F		&	F	&	T	&	T	&	T	&	F	&	F	&	F	\\
%%	&T	   &	F	&	F   &	T	&	T		&	T	&	T	&	F	&	T	&	T	&	T	&	T	\\
%%	&T	   &	F	&	F   &	T	&	F		&	F	&	T	&	F	&	T	&	T	&	T	&	F	\\
%%	&T	   &	F	&	F   &	F	&	F		&	T	&	T	&	F	&	T	&	F	&	T	&	T	\\
%%	&T	   &	F	&	F   &	F	&	F		&	F	&	F	&	F	&	F	&	F	&	F	&	F	\\
%%\cline{8-8}
%%\end{tabular}
%%\end{center}
%%}
%
%\vspace{1em}
%
%\end{earg}
%
%\problempart
%Write complete truth tables for the following sentences and mark the column that represents the possible truth values for the whole sentence.
%
%\begin{earg}
%
%\item	$(D \eand \enot D) \eif G $
%
%
%\myanswer{\vspace{1em}
%\begin{center}
%\begin{tabular}{ccccc|c|c}
%\cline{6-6}
%	&	(D 	&	 \eand 	& 	 \enot	&	 D) 	&	 \eif 	&	 G \\
% \hline
%	&	T	&	F		&	F		&	T	&	T	&	T	\\
%	&	T	&	F		&	F		&	T	&	T	&	F	\\
%	&	F	&	F		&	T		&	F	&	T	&	T	\\
%	&	F	&	F		&	T		&	F	&	T	&	F	\\
%\cline{6-6}
%\end{tabular}
%\end{center}
%}
%\vspace{1em}
%
%
%\item	$(\enot P \eor \enot M) \eif M $
%
%%\myanswer{
%%\begin{center}
%%\begin{tabular}{cccccc|c|c}
%%\cline{7-7}
%%	&	(\enot 	&	P 	&	\eor 	&	\enot 	& 	 M) 	& 	\eiff 	&	 M \\
%% \hline
%%	&	F		&	T	&	F	&	F		&	T	&	F	&	T	\\
%%	&	F		&	T	&	T	&	T		&	F	&	F	&	F	\\
%%	&	T		&	F	&	T	&	F		&	T	&	T	&	T	\\
%%	&	T		&	F	&	T	&	T		&	F	&	F	&	F	\\
%%\cline{7-7}
%%\end{tabular}
%%\end{center}
%%}
%\vspace{1em}
%
%
%
%\item	$\enot \enot (\enot A \eand \enot B)  $
%
%\myanswer{
%\begin{center}
%\begin{tabular}{c|c|cccccc}
%\cline{2-2}
%	&	\enot		&	 \enot 	&	(\enot 	& 	 A 	& \eand 	& 	\enot 	&	 B)  \\
% \hline
%	&	F		&	T		&	F		&	T	&	F	&	F		&	T	\\
%	&	F		&	T		&	F		&	T	&	F	&	T		&	F	\\
%	&	F		&	T		&	T		&	F	&	F	&	F		&	T	\\
%	&	T		&	F		&	T		&	F	&	T	&	T		&	F	\\
%\cline{2-2}
%\end{tabular}
%\end{center}
%}
%\vspace{1em}
%
%
%
%\item 	$[(D \eand R) \eif I] \eif \enot(D \eor R) $
%
%\myanswer{
%\begin{center}
%\begin{tabular}{cccccc|c|cccc}
%\cline{7-7}
%	&	[(D 	& 	 \eand 	& 	 R)	& 	\eif 	&	I] 	&	\eif 	&	 \enot 	&	(D 	&	 \eor 	& R) \\
%	 \hline
%	&	T	&	T		&	T	&	T	&	T	&	F	&	F		&	T	&	T		&T	\\
%	&	T	&	T		&	T	&	F	&	F	&	T	&	F		&	T	&	T		&T	\\
%	&	T	&	F		&	F	&	T	&	T	&	F	&	F		&	T	&	T		&F	\\
%	&	T	&	F		&	F	&	T	&	F	&	F	&	F		&	T	&	T		&F	\\
%	&	F	&	F		&	T	&	T	&	T	&	F	&	F		&	F	&	T		&T	\\
%	&	F	&	F		&	T	&	T	&	F	&	F	&	F		&	F	&	T		&T	\\
%	&	F	&	F		&	F	&	T	&	T	&	T	&	T		&	F	&	F		&F	\\
%	&	F	&	F		&	F	&	T	&	F	&	T	&	T		&	F	&	F		&F	\\
%\cline{7-7}
%\end{tabular}
%\end{center}
%}
%
%\vspace{1em}
%
%
%%\item	$\enot [(D \eiff O) \eiff A] \eif (\enot D \eand O) $
%%\myanswer{
%%&\begin{center}
%%\begin{tabular}{ccccccc|c|cccc}
%%\cline{8-8}
%%	&	\enot 	&	[(D 	&	\eiff 	&	O) 	&	\eiff 	&	 A]	& 	\eif 	 &	(\enot 	& 	D 	 & 	 \eand &O) \\
%%	\hline
%%	&	F		&	T	&	T	&	T	&	T	&	T	&	T	&	F		&	T	&	F	&T	\\
%%	&	T		&	T	&	T	&	T	&	F	&	F	&	F	&	F		&	T	&	F	&T	\\
%%	&	T		&	T	&	F	&	F	&	F	&	T	&	F	&	F		&	T	&	F	&F	\\
%%	&	F		&	T	&	F	&	F	&	T	&	F	&	T	&	F		&	T	&	F	&F	\\
%%	&	T		&	F	&	F	&	T	&	F	&	T	&	T	&	T		&	F	&	T	&T	\\
%%	&	F		&	F	&	F	&	T	&	T	&	F	&	T	&	T		&	F	&	T	&T	\\
%%	&	F		&	F	&	T	&	F	&	T	&	T	&	T	&	T		&	F	&	F	&F	\\
%%	&	T		&	F	&	T	&	F	&	F	&	F	&	F	&	T		&	F	&	F	&F	\\
%%\cline{8-8}
%%\end{tabular}
%%\end{center}
%%}
%%\vspace{1em}
%\end{earg}
%
%
%
%\problempart
% Can you think of sentences with the following truth table:
%\begin{enumerate}
%\item \begin{tabular}{cc|c}
%$A$&$B$&?\\\hline
%T&T&T\\
%T&F&T\\
%F&T&T\\
%F&F&T
%\end{tabular}\myanswer{$A\eor\enot A$}
%\item \begin{tabular}{cc|c}
%$A$&$B$&?\\\hline
%T&T&F\\
%T&F&F\\
%F&T&T\\
%F&F&F
%\end{tabular}\myanswer{$\enot A\eand B$}
%\item \begin{tabular}{cc|c}
%$A$&$B$&?\\\hline
%T&T&T\\
%T&F&F\\
%F&T&T\\
%F&F&T
%\end{tabular}\myanswer{\quad\begin{minipage}{.75\textwidth}
%$\enot( A\eand \enot B)$;\\ or a more systematic answer by simply writing out the different options when the sentence is true: $$((A\eand B)\eor (\enot A\eand B))\eor (\enot A\eand\enot B)$$
%\end{minipage}}
%
%\end{enumerate}
%\problempart
%Suppose $X$ is TFL sentence containing two atomic sentences. Then there are in fact sixteen different possible columns for $X$.
%\begin{enumerate}
%\item Can you explain why?\myanswer{\\ The first line might have T or F, the second line might have T or F, etc. There are 4 valuations, so there are $2\times 2\times 2\times 2=2^4=16$ different ways of putting Ts and Fs to these 4 lines. I.e.~16 different columns. }
%\item Can you show for each of these combinations that there is a sentence of TFL with that column describing its truth.
%%\myanswer{\\
%%See section 37.2 of forall$x$:Bristol for the general argument for this.
%%}
%\item {} Can you show there's always a formula just using $\enot$ and $\eand$ with that column describing its truth.
%%\myanswer{\\
%%The previous answer had that every truth table can be given by a sentence with $\eand, \eor$ and $\enot$. We can replace any instances of $\eor$ by $\eand$ using: $$\metaX\eor\metaY\text{ is logically equivalent to }\enot(\enot\metaX\eand\enot\metaY)$$
%%See 37.4 of forall$x$:Bristol for more details.
%%}
%\end{enumerate}
%
%If you want additional practice, you can construct truth tables for any of the sentences and arguments in the exercises for the previous chapter.
%
%\problempart
%\label{pr.TT.valid}
%\label{pr.TT.valid}
%Use truth tables to determine whether each argument is valid or invalid.
%\begin{earg}
%\item $A\eif A \therefore A$  \hfill \myanswer{Invalid (see line 2)}
%\myanswer{\begin{center}
%\begin{tabular}{c | d e f | c}
%$A$ &$A$&$\eif$&$A$&$A$\\
%\hline
% T & T & \TTbf{T} & T & T\\
% F & F & \TTbf{T} & F & F
% \end{tabular}
%\end{center}}
%\item $A\eif(A\eand\enot A) \therefore \enot A$  \hfill \myanswer{Valid}
%\myanswer{\begin{center}
%\begin{tabular}{c | d e e e e f | df}
%$A$&$A$&$\eif$&$(A$&$\eand$&$\enot$&$A)$&$\enot$&$A$\\
%\hline
% T & T & \TTbf{F} & T & F& F&T&\TTbf{F}&T\\
% F & F & \TTbf{T} & F & F&T&F&\TTbf{T}&F
% \end{tabular}
%\end{center}}
%\item $A\eor(B\eif A) \therefore \enot A \eif \enot B$  \hfill \myanswer{Valid}
%\myanswer{\begin{center}
%\begin{tabular}{c c | d e e e f | d e e e f}
%$A$ & $B$ & $A$&$\eor$&$(B$&$\eif$&$A)$&$\enot$&$A$&$\eif$&$\enot$&$B$\\
%\hline
%T & T & T & \TTbf{T} & T & T & T & F & T & \TTbf{T} & F & T \\
%T & F & T & \TTbf{T} & F & T & T & F & T & \TTbf{T} & T & F \\
%F & T & F & \TTbf{F} & T & F & F & T & F & \TTbf{F} & F & T \\
%F & F & F & \TTbf{T} & F & T & F & T & F & \TTbf{T} & T & F
%\end{tabular}
%\end{center}}
%\item $A\eor B, B\eor C, \enot A \therefore B \eand C$  \hfill \myanswer{Invalid (see line 6)}
%\myanswer{\begin{center}
%\begin{tabular}{c c c | d e f | d e f | d f | d e f}
%$A$ & $B$ & $C$ & $A$&$\eor$&$B$&$B$&$\eor$&$C$&$\enot$&$A$&$B$&$\eand$&$C$\\
%\hline
%T & T & T & T & \TTbf{T} & T & T & \TTbf{T} & T & \TTbf{F} & T & T & \TTbf{T} & T \\
%T & T & F & T & \TTbf{T} & T & T & \TTbf{T} & F & \TTbf{F} & T & T &\TTbf{F} & F \\
%T & F & T & T & \TTbf{T} & F & F & \TTbf{T} & T & \TTbf{F} & T & F & \TTbf{F} & T \\
%T & F & F & T & \TTbf{T} & F & F & \TTbf{F} & F & \TTbf{F} & T & F & \TTbf{F} & F\\
%T & T & T & F & \TTbf{T} & T & T & \TTbf{T} & T & \TTbf{T} & F & T & \TTbf{T} & T \\
%T & T & F & F & \TTbf{T} & T & T & \TTbf{T} & F & \TTbf{T} & F & T &\TTbf{F} & F \\
%T & F & T & F & \TTbf{F} & F & F & \TTbf{T} & T & \TTbf{T} & F & F & \TTbf{F} & T \\
%T & F & F & F & \TTbf{F} & F & F & \TTbf{F} & F & \TTbf{T} & F & F & \TTbf{F} & F
%\end{tabular}
%\end{center}}
%\item $(B\eand A)\eif C, (C\eand A)\eif B \therefore (C\eand B)\eif A$  \hfill \myanswer{Invalid (see line 5)}
%\myanswer{\begin{center}
%\begin{tabular}{c c c | d e e e f | d e e e f | d e e e f}
%$A$ & $B$ & $C$ & $(B$&$\eand$&$A)$&$\eif$&$C$&$(C$&$\eand$&$A)$&$\eif$&$B$&$(C$&$\eand$&$ B)$&$\eif$&$A$\\
%\hline
%T & T & T & T & T & T & \TTbf{T} & T & T & T & T & \TTbf{T} & T & T & T & T & \TTbf{T} & T\\
%T & T & F & T & T & T & \TTbf{F} & F & F & F & T & \TTbf{T} & T & F & F & T & \TTbf{T} & T\\
%T & F & T & F & F & T & \TTbf{T} & T & T & T & T & \TTbf{F} & F & T & F & F & \TTbf{T} & T\\
%T & F & F & F & F & T & \TTbf{T} & F & F & F & T & \TTbf{T} & F & F & F & F & \TTbf{T} & T\\
%F & T & T & T & F & F & \TTbf{T} & T & T & F & F & \TTbf{T} & T & T & T & T & \TTbf{F} & F\\
%F & T & F & T & F & F & \TTbf{T} & F & F & F & F & \TTbf{T} & T & F & F & T & \TTbf{T} & F\\
%F & F & T & F & F & F & \TTbf{T} & T & T & F & F & \TTbf{T} & F & T & F & F & \TTbf{T} & F\\
%F & F & F & F & F & F & \TTbf{T} & F & F & F & F & \TTbf{T} & F & F & F & F & \TTbf{T} & F
%\end{tabular}
%\end{center}}
%\end{earg}
%
%\problempart Determine whether each sentence is a tautology, a contradiction, or a contingent sentence, using a complete truth table.
%\begin{earg}
%\item $\enot B \eand B$ \vspace{.5ex} \hfill \myanswer{Contradiction}
%
%
%\item $\enot D \eor D$ \vspace{.5ex} \hfill \myanswer{Tautology}
%
%
%\item $(A\eand B) \eor (B\eand A)$\vspace{.5ex} \hfill \myanswer{Contingent}
%
%
%\item $\enot[A \eif (B \eif A)]$\vspace{.5ex} \hfill \myanswer{Contradiction}
%
%
%%\item $(A \eif [A \eif (B \eand \enot B)]$ \vspace{.5ex} \hfill \myanswer{Contradiction}
%
%
%\item $[(A \eand B) \eif B] \eif (A \eif B)$ \vspace{.5ex} \hfill \myanswer{Contingent}
%
%
%\end{earg}
%
%\noindent\problempart
%\label{pr.TT.equiv}
%Determine whether each the following sentences are logically equivalent using complete truth tables. If the two sentences really are logically equivalent, write ``equivalent.'' Otherwise write, ``Not equivalent.''
%\begin{earg}
%\item $A$ and $\enot A$
%\item $A \eand \enot A$ and $\enot B \eif B$
%\item $[(A \eor B) \eor C]$ and $[A \eor (B \eor C)]$
%\item $A \eor (B \eand C)$ and $(A \eor B) \eand (A \eor C)$
%\item $[A \eand (A \eor B)] \eif B$ and $A \eif B$\end{earg}
%
%
%\problempart
%\label{pr.TT.equiv2}
%Determine whether each the following sentences are logically equivalent using complete truth tables. If the two sentences really are equivalent, write ``equivalent.'' Otherwise write, ``not equivalent.''
%\begin{earg}
%\item $A\eif A$ and $(A \eif A)\eand (\enot A\eif\enot A)$
%\item $\enot(A \eif B)$ and $\enot A \eif \enot B$
%\item $A \eor B$ and $\enot A \eif B$
%\item$(A \eif B) \eif C$ and $A \eif (B \eif C)$
%\item $A \eif (B \eand C)$ and $A \eand (B \eand C)$
%\end{earg}
%
%\problempart
%\label{pr.TT.satisfiable2}
%
%\noindent\problempart
%\label{pr.TT.satisfiable3}
%Determine whether each collection of sentences is consistent or inconsistent, using a complete truth table.
%\begin{earg}
%\item $\enot B$, $A \eif B$, $A$ \vspace{.5ex} \hfill \myanswer{Inconsistent}
%\item $\enot(A \eor B)$, $A \eif B$, $B \eif A$\vspace{.5ex} \hfill \myanswer{Consistent}
%\item $A \eor B$, $\enot B$, $\enot B \eif \enot A$\vspace{.5ex} \hfill \myanswer{Inconsistent}
%\item $B \eif A$, $\enot B \eor \enot A$, $A \eif B$\vspace{.5ex} \hfill \myanswer{Consistent}
%\item $(A \eor B) \eor C$, $\enot A \eor \enot B$, $\enot C \eor \enot B$\vspace{.5ex} \hfill \myanswer{Consistent}
%\end{earg}
%
%\noindent\problempart
%\label{pr.TT.valid2}
%Determine whether each argument is valid or invalid, using a complete truth table.
%\begin{earg}
%\item $A\eif B$, $B \therefore  A$ \hfill \myanswer{Invalid}
%
%\item $A\eif B$, $B\eif C \therefore A\eif C$ \hfill \myanswer{Valid}
%
%\item $A \eif B$, $A \eif C\therefore B \eif C$ \hfill \myanswer{Invalid}.
%
%%\item $A \eif B$, $B \eif A\therefore A \eiff B$ \hfill \myanswer{Valid}
%\end{earg}
%
%\noindent\problempart
%\label{pr.TT.valid3}
%Determine whether each argument is valid or invalid, using a complete truth table. If invalid, provide a valuation that is a counterexample.
%\begin{earg}
%\item $A\eor\bigl[A\eif(A\eif A)\bigr] \therefore  A $\vspace{.5ex} \hfill \myanswer{Invalid}
%\item $A\eor B$, $B\eor C$, $\enot B \therefore A \eand C$\vspace{.5ex} \hfill \myanswer{Valid}
%\item $A \eif B$, $\enot A\therefore \enot B$ \vspace{.5ex} \hfill \myanswer{Invalid}
%\item $A$, $B\therefore \enot(A\eif \enot B)$ \vspace{.5ex} \hfill \myanswer{Valid}
%\item $\enot(A \eand B)$, $A \eor B$, $A \eif B\therefore C$ \vspace{.5ex} \hfill \myanswer{Valid}
%\end{earg}
%
%\noindent\problempart
%\label{pr.TT.meta}
%Are the following statements true? Why?
%\begin{itemize}
%\item if $Y$ is a logical consequence of $X$, then $X\eif Y$ is a logical truth.\myanswer{Yes}
%\item if $X\eif Y$ is a logical truth, then $Y$ is a logical consequence of $X$.\myanswer{Yes}
%\item if $X\eif Y\eand\enot Y$ is a logical truth, then $X$ is a logical contradiction.\myanswer{Yes}
%\item if $X\eif Y\eand\enot Y$ is true relative to a valuation, then $X$ is a logical contradiction.\myanswer{No}
%\item if $X\eor\enot X\eif Y$ is a logical truth, then $Y$ is a tautology.\myanswer{Yes}
%\item if $X\eor\enot X\eif Y$ is true relative to a valuation, then $Y$ is a tautology.\myanswer{No}
%\end{itemize}
%\end{practiceproblems}
